### 2019-12-27 paper.tex
bootstrapped pub workflow

% -*- coding: utf-8; -*-
% vim: set fileencoding=utf-8 :
\documentclass[english,submission]{programming}
%% First parameter: the language is 'english'.
%% Second parameter: use 'submission' for initial submission, remove it for camera-ready (see 5.1)

\usepackage[backend=biber]{biblatex}
\addbibresource{references.bib}

\begin{document}

\title{Wildcard: Browser Extensions for All}
\subtitle{}% optional

\author{Geoffrey Litt}
\authorinfo{is bla bla bla}
\author{Daniel Jackson}
\authorinfo{is bla bla bla}
\affiliation{Massachusetts Institute of Technology}

\keywords{programming journal, paper formatting, submission preparation} % please provide 1--5 keywords

%%%%%%%%%%%%%%%%%%
%% These data MUST be filled for your submission. (see 5.3)
\paperdetails{
  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
  %% Choose exactly the one that best describes this work. (see 2.1)
  perspective=art,
  %% State one or more areas, separated by a comma. (see 2.2)
  %% Please see list of areas in http://programming-journal.org/cfp/
  %% The list is open-ended, so use other areas if yours is/are not listed.
  area={Social Coding, General-purpose programming},
  %% You may choose the license for your paper (see 3.)
  %% License options include: cc-by (default), cc-by-nc
  % license=cc-by,
}
%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%
%% These data are provided by the editors. May be left out on submission.
%\paperdetails{
%  submitted=2016-08-10,
%  published=2016-10-11,
%  year=2016,
%  volume=1,
%  issue=1,
%  articlenumber=1,
%}
%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please go to https://dl.acm.org/ccs/ccs.cfm and generate your Classification
% System [view CCS TeX Code] stanz and copy _all of it_ to this place.
%% From HERE
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002944.10011122.10003459</concept_id>
<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10010405.10010476.10010477</concept_id>
<concept_desc>Applied computing~Publishing</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}
\ccsdesc[500]{Applied computing~Publishing}

% To HERE
%%%%%%%%%%%%%%%%%%%%%%%

\maketitle

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}

\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

This is the introduction of the paper.

\hypertarget{demos}{%
\section{Demos}\label{demos}}

\acks
\printbibliography

\end{document}

% Local Variables:
% TeX-engine: luatex
% End:

### 2019-12-27 templates/programming-template/example-programming.tex
bootstrapped pub workflow

% -*- coding: utf-8; -*-
% vim: set fileencoding=utf-8 :
\documentclass[english,submission]{programming}
%% First parameter: the language is 'english'.
%% Second parameter: use 'submission' for initial submission, remove it for camera-ready (see 5.1)

\usepackage[backend=biber]{biblatex}
\addbibresource{example.bib}

%
% Packages and Commands specific to article (see 3)
%
% These ones  are used in the guide, replace with your own.
% 
\usepackage{multicol}
\lstdefinelanguage[programming]{TeX}[AlLaTeX]{TeX}{%
  deletetexcs={title,author,bibliography},%
  deletekeywords={tabular},
  morekeywords={abstract},%
  moretexcs={chapter},%
  moretexcs=[2]{title,author,subtitle,keywords,maketitle,titlerunning,authorinfo,affiliation,authorrunning,paperdetails,acks,email},
  moretexcs=[3]{addbibresource,printbibliography,bibliography},%
}%
\lstset{%
  language={[programming]TeX},%
  keywordstyle=\firamedium,
  stringstyle=\color{RosyBrown},%
  texcsstyle=*{\color{Purple}\mdseries},%
  texcsstyle=*[2]{\color{Blue1}},%
  texcsstyle=*[3]{\color{ForestGreen}},%
  commentstyle={\color{FireBrick}},%
  escapechar=`,}
\newcommand*{\CTAN}[1]{\href{http://ctan.org/tex-archive/#1}{\nolinkurl{CTAN:#1}}}
%%

\begin{document}

\title{The Art, Science, and Engineering of Programming}
\subtitle{Preparing Articles for Programming}% optional
\titlerunning{Preparing Articles for Programming} %optional, in case that the title is too long; the running title should fit into the top page column

\author[a]{Tobias Pape}
\authorinfo{is the author of this {LaTeX} class. Contact him at
  \email{tobias.pape@hpi.uni-potsdam.de}.}
\affiliation[a]{Hasso Plattner Institute, University of Potsdam, Germany}
\author{Cristina V. Lopes}
\authorinfo{is associate editor for the first two issues of The Art, Science,
  and Engineering of Programming. Contact her at \email{lopes@ics.uci.edu}.}
\affiliation{University of California, Irvine, USA}
\author[a]{Robert Hirschfeld}
\authorinfo{is chair of the AOSA steering committee. The Art, Science,
  and Engineering of Programming is published by AOSA. Contact Robert at \email{hirschfeld@hpi.uni-potsdam.de}.}

% \authorrunning{T. Pape, C. Lopes, R. Hirschfeld} % Optional, for long author lists

\keywords{programming journal, paper formatting, submission preparation} % please provide 1--5 keywords

%%%%%%%%%%%%%%%%%%
%% These data MUST be filled for your submission. (see 5.3)
\paperdetails{
  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
  %% Choose exactly the one that best describes this work. (see 2.1)
  perspective=art,
  %% State one or more areas, separated by a comma. (see 2.2)
  %% Please see list of areas in http://programming-journal.org/cfp/
  %% The list is open-ended, so use other areas if yours is/are not listed.
  area={Social Coding, General-purpose programming},
  %% You may choose the license for your paper (see 3.)
  %% License options include: cc-by (default), cc-by-nc
  % license=cc-by,
}
%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%
%% These data are provided by the editors. May be left out on submission.
%\paperdetails{
%  submitted=2016-08-10,
%  published=2016-10-11,
%  year=2016,
%  volume=1,
%  issue=1,
%  articlenumber=1,
%}
%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please go to https://dl.acm.org/ccs/ccs.cfm and generate your Classification
% System [view CCS TeX Code] stanz and copy _all of it_ to this place.
%% From HERE
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002944.10011122.10003459</concept_id>
<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10010405.10010476.10010477</concept_id>
<concept_desc>Applied computing~Publishing</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}
\ccsdesc[500]{Applied computing~Publishing}

% To HERE
%%%%%%%%%%%%%%%%%%%%%%%

\maketitle

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
  Many academic journals and conferences exist that publish
  research related to programming, starting with programming languages,
  software engineering, and expanding to the whole Computer Science field. Yet,
  many of us feel that, as the field of Computer Science expanded, programming,
  in itself, has been neglected to a secondary role not worthy of scholarly
  attention.

  We posit that a scholarly journal focusing primarily on programming
  will help close this gap, and foster a more meaningful bridge
  between research and the practice of software development.

  The Art, Science, and Engineering of Programming accepts papers
  under several perspectives: the art itself (programming styles,
  pearls, models, languages), the emerging science of understanding what works
  and what doesn’t work in general and in specific contexts, as well as more
  established engineering and mathematical perspectives.

  This journal aims at placing the wonderful art of programming in the
  map of scholarly works.

  We present this short paper as an example of, and a guide to,
  writing articles for The Art, Science, and Engineering of
  Programming journal. The complete template is provided: the class
  file (\texttt{programming.cls}), an example paper (\texttt{example-programming.tex}, this file), and an example
  bibliography file (\texttt{example.bib}).

  Much of the progress in Computer Science lies on the basis of
  computer programs, the people who write them, and the concepts and
  tools available to them to express computational tasks. The
  existence of a journal focusing on programming is of utmost
  importance for strengthening our understanding of these activities.
\end{abstract}

\section{A Guided Tour}
\label{sec:guided-tour}

The content of a paper is its most important ingredient. That being said,
papers for The Art, Science, and Engineering of Programming should adhere to a
common style, which is achieved with the \lstinline|programming| {LaTeX}
class. You are to use this class as outlined here when submitting articles to
the journal. This document acts as an example of what this can look like and a
short references for all its parts.

\subsection{Document Parts}
\label{sec:document-parts}

\begin{lstlisting}[float,label=lst:tour,caption={A very minimal article.}]
\documentclass[english,submission]{programming}`\label{l:doc}`
\usepackage[backend=biber]{biblatex} % Use Biblatex `\label{l:bib}`
\addbibresource{example.bib}

\begin{document}

\paperdetails{`\label{l:details}`
  perspective=engineering,
  area={General-purpose programming}
}

\title{The importance of why and how to do work}`\label{l:meta}`
\author{Anna Author}
\affiliation{The Unseen University, Ankh-Morpork}
\author{Bert Betatester}
\affiliation{Carolingian Minuschool Academy, Bodoni, San Serriffe}

\keywords{paper, showcase, lorem ipsum}`\label{l:keywords}`

\begin{CCSXML}`\label{l:ccs:start}`
<ccs2012>
<concept>
<concept_id>10002944.10011122.10003459</concept_id>
<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}`\label{l:ccs:end}`

\maketitle

\begin{abstract}
  This paper shows...
\end{abstract}

\section{Introduction}
The art of computer programming~\cite{DBLP:journals/cacm/Knuth74}...
\section{...}
...
\acks`\label{l:acks}`
I want to thank ....

\printbibliography
\end{document}
\end{lstlisting}

{\lstset{language=}%
  We have included a most minimal article in \autoref{lst:tour} that you can
  adapt to the concrete work to be submitted. First, you
  have to select the \lstinline|programming| class as in \autoref{l:doc}. The
  parameter \lstinline|english| is for language selection (optional, but
  recommended). The other parameter, \lstinline|submission|, is crucial for the
  reviewing process and must be included for submission. More information on
  document parameters can be found in \autoref{sec:class-options}. After that,
  we recommend that you load the \lstinline|biblatex| package. While it is %
  perfectly fine to use plain {BibTeX}, we recommend
  \lstinline|biblatex|, as it has native Unicode support and can handle
  \textsc{url}s and \textsc{doi}s better. In both cases, we have preloaded a
  bibliography style.

  At that point, you can load necessary {LaTeX} packages. We already load some
  important packages with the \lstinline|programming| class and that there are
  certain packages that are not recommended or incompatible with this class.
  Refer to \autoref{sec:pack-cons} for which packages are recommended. The you
  start the document with \lstinline|\begin{document}|.

  You have to provide details to your submission that will be used during the
  review process. For that, use the \lstinline|paperdetails| command and
  specify \emph{perspective} and \emph{area} of your submission. For
  perspective, choose form one of the \emph{Art}, \emph{theoretical} or
  \emph{empirical} \emph{Science}, or \emph{Engineering} and provide it as
  argument to the \lstinline|perspective| keyword; they can be found in
  \autoref{sec:perspective}. For area, specify an argument to the
  \lstinline|area| keyword (in braces). Find a list of suggested areas in
  \autoref{sec:area-submission}.

We need you to provide certain metadata which you can do with the following
commands, as form \autoref{l:meta}. You can specify the title and maybe the subtitle with
\lstinline|\title| (and \lstinline|\subtitle|, respectively). Then please
list all authors. For that, use one \lstinline|\author| command per author
and give an affiliation for each with \lstinline|\affiliation|. When several
consecutive authors share the same affiliation, you should only use one
\lstinline|\affiliation| command after the last of them. More information can
be found in \autoref{sec:document-metadata}.

Please include one to five keywords, separated by commas, that are important to
your work with the \lstinline|\keywords| command as in \autoref{l:keywords}. We
kindly ask you to classify your work using the \emph{ACM Computing
  Classification System}\footnote{\url{https://dl.acm.org/ccs/ccs.cfm}.}.
Please visit the ACM's website and generate the {TeX} code that matches
your classification. The code should consist of a \lstinline|CCSXML|
environment and several \lstinline|\ccsdesc| commands, as in lines
\ref{l:ccs:start} to \ref{l:ccs:end}.

With \lstinline|\maketitle| and the \lstinline|abstract| environment, you can
produce the article's title page. Then you can start with your content. If
applicable, you can put acknowledgements at the end after an \lstinline|\acks|
command, as in \autoref{l:acks}.}

\subsection{Running {LaTeX}}
\label{sec:running}

This class has been tested with the popular {LaTeX} programs available.
It should work with {pdfLaTeX}, {LuaLaTeX}, and {XeLaTeX}.
We suggest that you use {LuaLaTeX}, which is included in all major
{TeX} system installations, has a good Unicode support, and is well portable.
Please do not use the LaTeX/dvips combination to produce output files.
To get all metadata right, several LaTeX runs will be necessary. Hence, a
typical run looks like this:
\begin{lstlisting}[language=]
lualatex myarticle
biber myarticle
lualatex myarticle
lualatex myarticle
\end{lstlisting}

You should replace \lstinline|lualatex| with the LaTeX program of your choice,
and \lstinline|biber| with \lstinline|bibtex| if you do not want to use
Biblatex (or use Biblatex's BibTeX backend).

Your \textsc{pdf} file for your article should then be ready for submission.

\section{Submission Classification}

Almost anything about programming is in scope for The Art, Science, and
Engineering of Programming. However, you should classify your article according
to the following aspects.

\subsection{Perspective}
\label{sec:perspective}

We accept descriptions of work under different perspectives:

\emph{The Art.} This perspective is about knowledge and technical skills acquired through
practice and personal experiences. Examples include libraries, frameworks,
languages, APIs, programming models and styles, programming pearls, and essays
about programming.

\emph{Science (theoretical).} This perspective is about knowledge and technical skills acquired
through mathematical formalisms. Examples include formal programming models and
proofs.

\emph{Science (empirical).} This perspective is about knowledge and technical skills acquired
through experiments and systematic observations. Examples include user studies
and programming-related data mining.

\emph{Engineering.} This perspective is about knowledge and technical skills acquired through
designing and building large systems and through calculated application of
principles in building those systems. Examples include measurements of
artifacts’ properties, development processes and tools, and quality assurance
methods.

\bigskip
To classify your work as any of the above perspectives, please set the
\lstinline|perspective| keyword of  the \lstinline|\paperdetails| command to
one of the following values:

\begin{description}
\item[art] for submissions concerning \emph{The Art } (\textsf{theart} is a
  proper alias);
\item[sciencetheoretical] For submissions concerning \emph{Science
    (theoretical)}, with the aliases \textsf{sciencetheoretical},
  \textsf{theoreticalscience}, \textsf{theoretical},
  \textsf{science-theoretical}, and finally \textsf{theoretical-science};
\item[scienceempirical] for submissions concerning \emph{Science (empirical)},
  the valid aliases are \textsf{scienceempirical}, \textsf{empiricalscience},
  \textsf{empirical}, \textsf{science-empirical}, and
  \textsf{empirical-science}; and
\item[engineering] for submissions concerning \emph{Engineering}.
\end{description}

\subsection{Area of Submission}
\label{sec:area-submission}

Independent of the type of work, the journal accepts submissions covering
several expertise areas. Expertise areas include, but are not limited to:

\begin{itemize}
\item General-purpose programming
\item Distributed systems programming
\item Parallel and multi-core programming
\item Graphics and GPU programming
\item Security programming
\item User interface programming
\item Database programming
\item Visual and live programming
\item Data mining and machine learning programming, and for programming
\item Interpreters, virtual machines and compilers
\item Modeling and modularity
\item Testing and debugging
\item Program verification
\item Programming education
\item Programming environments
\item Social coding
\end{itemize}

To specify the area of expertise for your work, please set the
\lstinline|area| keyword of  the \lstinline|\paperdetails| command to
one of the preceding suggested areas, or provide your own. Please use curly
braces \lstinline|{}| around the value.

\section{Choice of License (optional)}

All work published by The Art, Science, and Engineering of Programming will be
available as open-access. Therefore, authors submitting a paper to The Art,
Science, and Engineering of Programming should grant a Creative Commons
license. We support two licenses (description adapted from
Wikipedia\footnote{\url{https://en.wikipedia.org/w/index.php?title=Creative_Commons_license&oldid=754498632}}):
\begin{description}
\item[\ccby] \href{https://creativecommons.org/licenses/by/4.0}{Attribution 4.0
    \textsmaller{\textsc{cc by 4.0}}}\\
  {\small
    Licensees may copy, distribute, display and perform the work and make
    derivative works and remixes based on it only if they give the author or
    licensor the credits (attribution) in the manner specified by these.}
  \item[\ccbync]
  \href{https://creativecommons.org/licenses/by-nc-sa/4.0}{Attribution-NonCommercial
    4.0 \textsmaller{\textsc{cc by-nc 4.0}}}\\
  {\small
  Licensees may copy, distribute, display and perform the work and make
  derivative works and remixes based on it only if they give the author or
  licensor the credits (attribution) in the manner specified by these and only
  for non-commercial purposes.}
\end{description}
%
To choose a license for your work, set the
\lstinline|license| keyword of  the \lstinline|\paperdetails| command to
either \lstinline|cc-by| or \lstinline|cc-by-sa|. Please be aware that, in absence of a choice, we assume \textsc{cc by}.

\section{Typographical and Technical Aspects}

We have carefully chosen the presentation for articles published in The Art, Science, and
Engineering of Programming to fit on-screen reading and occasional printouts.
It is important to maintain a uniform look of all articles appearing in the
journal. We kindly ask you therefore to \emph{not} change the presentation. Most
importantly,
%
\begin{itemize}
\item please do not change the font or the font size for that matter. The text
  font remains fixed as \emph{Charter} (in the XCharter variant) and emphasis
  font is \emph{Fira Sans}. The only exception is the typewriter font, where
  you can choose from two variants (please refer to
  \autoref{sec:typewriter-fonts}). Moreover;
\item please do not change the margins or the line spacing;
\item please refrain from using the \lstinline|\sloppy| command, especially for
  the whole document;
\item please avoid the use of the \lstinline|[H]| or \lstinline|[h!]| modifiers
  for figures, tables, and similar elements;
\item please use a consistent formatting  (for example, use the
  \lstinline|siunitx| package~\cite{siunitx}).
\end{itemize}
%
This class has been tested on several LaTeX systems. We suggest that you use an
up-to-date system, for example
TeX\,Live\footnote{\url{https://www.tug.org/texlive/}.}, which is available for
Linux, Windows, and Apple systems, among others. We tested with the 2015 and
2016 systems and will make sure that it will work with upcoming systems, too.
Please first upgrade your system if problems with this class should occur.
Regarding the LaTeX system, please keep in mind to
%
\begin{itemize}
\item use LaTeX packages only if required, as fewer packages lower the
  chance of conflicts (refer to \autoref{sec:pack-cons} for details);
\item keep the use of custom macros low to not interfere with provided
  packages and use LaTeX commands to do so (for example, \lstinline|\newcommand| and
  similar instead of TeX's \lstinline|\def|);
\item watch out for unsuitable line- and pagebreaks;
\item when using pixel graphics, provide at least 300\,dpi imagery,
  \textsc{pdf} files preferred;
\item minimize the use of LaTeX comments, especially via block constructs;
\item consult the \emph{l2tabu} documentation~\cite{l2tabu} about outdated packages;
\item provide \textsc{doi}s for all your references, unless inaccessible.
\end{itemize}
%
We suggest the use of a spellchecker to avoid typos.

Please note that all article source must be \textsc{utf-8} encoded, no
exceptions.

\section{Detailed Reference}

Several aspects of this class as outlined above provide for some fine tuning.

\subsection{Class Options}
\label{sec:class-options}

As explained in~\autoref{sec:guided-tour}, submissions to the journal must use
the \lstinline|submission| package option. For the accepted final, to-be-published
version, this option should be omitted or replaced by  \lstinline|crc| for
“Camera-ready Copy”. Actually, both are shortcuts for the
\lstinline|phase| keyword, which could be used, instead. 
Hence, the following statements in either column are equivalent:\par\noindent
{\lstset{texcsstyle=*{\color{Purple}\mdseries\smaller}}
\begin{minipage}[t]{.45\linewidth}
\emph{Final phase}
\begin{lstlisting}
\documentclass[phase=final]{programming}
\documentclass[crc]{programming}
\documentclass{programming}
\end{lstlisting}
\end{minipage}%
\hfill%
\begin{minipage}[t]{.5\linewidth}
\emph{Submission phase}
\begin{lstlisting}[numbers= none]
\documentclass[phase=submission]{programming}
\documentclass[submission]{programming}
``
\end{lstlisting}
\end{minipage}}

\paragraph{Code Fonts}
\label{sec:typewriter-fonts}

The standard font for code, as selected with \lstinline|\texttt|,
\lstinline|\ttfamily|, or a listing, is set to be the same as the sans serif
font, \emph{Fira Sans}. This matches the rest of the document style very well.
However, it may be strictly necessary to have a non-proportional (or
monospaced) font for the typewriter style. We therefore provide a class option \lstinline|code|
to change the typewriter font. 
It takes one the following values:
\begin{description}
\item[code=sf] Sets the code font to the sans serif font (Fira Sans). This is the
  default. Aliases for this option are \lstinline|code=sans| and
  \lstinline|code=sansserif|.
\item[code=tt] Sets the code font to Fira Sans' monospaced equivalent, Fira Mono. Aliases for this option are \lstinline|code=mono| and
  \lstinline|code=monospace|. This is the preferred choice if monospaced
  typewriter font is necessary. Note that this font does not have italics.
\end{description}

\paragraph{Other Options}

All other options give in the \lstinline|\documentclass| command are not
processed directly by the class but rather passed on to other packages as
``global options''. For example, the \lstinline|english| option from the
example above does not influence the \lstinline|programming| class directly,
but rather will be picked up by Babel~\cite{babel} to set up language options.

\subsection{Document Metadata}
\label{sec:document-metadata}

The presentation of the document metadata can be fine tuned, as well.
Besides the \lstinline|\title| and \lstinline|\subtitle| commands, there is the
\lstinline|\titlerunning| command to specify a shorter title when the normal
title does not fit into the page header. Similarly, the
\lstinline|\authorrunning| command can be used to specify the presentation of
the authors in the header.

Sometime listing the affiliations of the authors can be tricky when they are
shared. An optional argument to the \lstinline|\author| and
\lstinline|\affiliation| command can be used to associate authors and
affiliations explicitly. Four variants for affiliations are outlined in the
beginning of \autoref{lst:affiliation}.

For the final publication, we ask you to provide a short information of each
author, typically a very short biography, and a photo. You can use the
\lstinline|\authorinfo| command right after an \lstinline|\author| command to
associate that author information with an author. See \autoref{l:authorinfo} in
\autoref{lst:affiliation} for example. The photo is given as optional argument
as file name in brackets (without file extension). Omitting the photo will
produce a blank space next to the author information. The actual information
follows as an argument. Note that the author's \emph{name} is prepended to the
information automatically and is not to be repeated.

An author's email address can be given in either the affiliation or author
information block. Please use the \lstinline|\email| command for that purpose.

\begin{lstlisting}[float,label=lst:affiliation,caption={Four
ways to specify affiliations\,/\,Author information example\,/\,Sample publication paper details}]
% Independent affiliations 
\author{Anna Author}
\affiliation{The Unseen University, Ankh-Morpork}
\author{Bert Betatester}
\affiliation{Carolingian Minuschool Academy, Bodoni, San Serriffe}

% Shared affiliations 
\author{Anna Author}
\author{Bert Betatester}
\affiliation{Carolingian Minuschool Academy, Bodoni, San Serriffe}

% Explicit affiliations 
\author[a]{Anna Author}
\author[b]{Bert Betatester}
\author[a]{Claire Cadence}
\affiliation[a]{The Unseen University, Ankh-Morpork}
\affiliation[b]{Carolingian Minuschool Academy, Bodoni, San Serriffe}

% Mixed affiliations 
\author[a]{Anna Author}
\author[a,b]{Bert Betatester}
\affiliation[a]{The Unseen University, Ankh-Morpork}
\affiliation[b]{Carolingian Minuschool Academy, Bodoni, San Serriffe}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Author information`\label{l:authorinfo}`
\author{Anna Author}
\authorinfo[anna]{is a researcher at the Unseen University in Ankh-Morpork and
a Ph\,D student to Rincewind. Contact her at
\email{anna.the.author@univ-unseen.edu.pratchett}}  
\affiliation[a]{The Unseen University, Ankh-Morpork}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Sample publication paper details
\paperdetails{
  submitted=2017-02-18,
  published=2017-08-01,
  year=2017,
  volume=2,
  issue=1,
  articlenumber=4,
}
\end{lstlisting}

\subsection{Paper Details}

Besides specifying the perspective and area of the submission, the
\lstinline|\paperdetails| command is also used to provide publication
information for the final, camera-ready-copy phase (class option
\lstinline|phase=final|). In this case, perspective and area are removed from
the title page and the Digital Object Identifier (\textsc{doi}), as well as the
dates of submission and publications are presented instead. The information
necessary for this is provided \emph{by the publisher} after acceptance. It
includes the following information:
%
\begin{description}
\item[year] The year of the articles publication.
\item[volume] The volume the article appears in.
\item[issue] The issue the article appears in.
\item[articlenumber] The number of the article within the volume.
\item[submitted] The date the article was submitted to (and received by) the
  publisher. Please use an \textsc{iso} date (\textsc{yyyy-mm-dd}).
\item[published] The date the article was published and made public by the
  publisher. Please use an \textsc{iso} date (\textsc{yyyy-mm-dd}).
\end{description}
%
See the end of \autoref{lst:affiliation} for an example.

\subsection{Figures, Tables, Listings}

Sometimes it is necessary to pay special attention to the \emph{floating
  elements} of your document, that is figures, tables, listings, and similar. 
While it is tempting to force those elements to the specific position where
they are written down, this is not always the best choice, especially since it
can interrupt the reading flow. Therefore please refrain form using the
\lstinline|[h!]| and  \lstinline|[H]| specifiers for such elements. If you
\emph{must} have the element at the current position, it maybe is not a
\emph{floating element}, in the first place. It is perfectly fine to use
graphics \emph{without} a \lstinline|\begin{figrue}...\end{figure}| environment
or use a \lstinline|\begin{tabular}...\end{tabular}| \emph{without} an
enclosing \lstinline|\begin{table}...\end{table}|.

\begin{table}
  \centering
  \caption{Differences between things projected and things achieved}
  \label{tbl:things}
  \begin{threeparttable}
    \footnotesize
    \begin{tabular}{>{}p{.33\linewidth}@{}cSs}
      \toprule
      Part           & done & {Value}   & Unit                      \\
      \midrule
      Title          & yes  &    2.3456 & \dB                       \\
      Abstract   &  yes  &       e3  & \kilo\hertz               \\
      \midrule
      \multicolumn{4}{c}{Rest is not entirely true}                 \\
      \midrule
      Context        & yes  &   90.473  & \percent                  \\
      Problem        & no\tnote{a} &    &                           \\
      Solution       & yes  & 5642.5    & \mega\byte                \\
      Implementation & yes  &    1.2e-3 & \metre\squared\per\second \\
      Evaluation     & no   &           &                           \\
      Related Work   & no   &           &                           \\
      Conclusion     & yes  &    4955.3 & \per\kilogram             \\
      \bottomrule
    \end{tabular}
    \begin{tablenotes}
    \item [a] Just a few things missing
    \end{tablenotes}
  \end{threeparttable}
\end{table}

The caption position of a floating element should match its content. For
tables, listings, algorithms, or similar list-like content, please use captions
\emph{above} its contents, as those elements are typically read from top to
bottom. Find examples in the listings already presented and
\autoref{tbl:things}. For figures, graphs, graphics, or similar, please us
captions \emph{below} the imagery, as in \autoref{fig:test}.

For tabular content, we suggest to avoid vertical lines altogether and keep
horizontal lines to a minimum. Refer to the \lstinline|booktabs| package
documentation for more information~\cite{booktabs}. Also, the
\lstinline|threeparttable| package is valuable if you need footnotes in your
table~\cite{threeparttable}.

\subsection{Bibliography}

As outlined, the \lstinline|programming| class supports both plain BibTeX as
well as the newer Biblatex bibliography package to present your references.
Please note that you may not change the bibliography style, it is pre-selected
by the class. It is a simple, numbered style with all references sorted
alphabetically. In text, it should look like
this:~\cite{DBLP:journals/cacm/Knuth74}. Note that the reference has the
\textsc{doi} included. We would like to encourage you to always provide the
\textsc{doi}s of the works you cite, if possible. However, please make sure in
your bibliography file, that the \lstinline|doi| entry is a plain entry without
a ``resolver'' (for example, \textsf{http://dx.doi.org/}) prepended. A typical
example for an entry format should look like in \autoref{lst:bib}. Please
shorten neither names nor journal articles, if possible.

The bibliography commands for your article should look like this (for Biblatex
and plain BibTeX, respectively):\\*
\begin{minipage}[t]{.45\linewidth}
\emph{Biblatex}
\begin{lstlisting}
\documentclass{programming}
\usepackage[backend=biber]{biblatex}
\addbibresource{example.bib}
\begin{document}
...
\printbibliography
\end{document}
\end{lstlisting}
\end{minipage}%
\hfill%
\begin{minipage}[t]{.45\linewidth}
\emph{Plain  BibTeX}
\begin{lstlisting}[numbers= none]
\documentclass{programming}

\begin{document}
...
\bibliography{example}
\end{document}
\end{lstlisting}
\end{minipage}

\begin{lstlisting}[float,label=lst:bib,language=,caption={A typical BibTeX entry for \cite{DBLP:journals/cacm/Knuth74}}]
@article{DBLP:journals/cacm/Knuth74,
  Author = {Donald E. Knuth},
  Doi = {10.1145/361604.361612},
  Journal = {Communications of the {ACM}},
  Number = {12},
  Pages = {667--673},
  Title = {{Computer Programming as an Art}},
  Volume = {17},
  Year = {1974}}
\end{lstlisting}

\subsection{{LaTeX} Packages Considerations }
\label{sec:pack-cons}

The following packages are automatically loaded by the \lstinline|programming|
class and need not to be loaded manually.

\begin{multicols}{4}
{\color{DarkGreen}
\begin{itemize}
\item \path{accsupp}
\item \path{amsmath}
\item \path{amstext}
\item \path{array}
\item \path{atbegshi}
\item \path{babel}
\item \path{booktabs}
\item \path{calc}
\item \path{caption}
\item \path{colortbl}
\item \path{comment}
\item \path{csquotes}
\item \path{doclicense}
\item \path{expl3}
\item \path{FiraSans}
\item \path{fnpct}
\item \path{fontspec}
\item \path{fontenc}
\item \path{gettitlestring}
\item \path{graphics}
\item \path{graphicx}
\item \path{grfext}
\item \path{hypcap}
\item \path{hyperref}
\item \path{hyperxmp}
\item \path{listings}
\item \path{mathdesign}
\item \path{microtype}
\item \path{morewrites}
\item \path{multirow}
\item \path{relsize}
\item \path{rotating}
\item \path{siunitx}
\item \path{subcaption}
\item \path{tabularx}
\item \path{textcase}
\item \path{textcomp}
\item \path{threeparttable}
\item \path{typearea}
\item \path{url}
\item \path{verbatim}
\item \path{xcolor}
\item \path{xparse}
\item \path{xspace}
\item \path{xunicode}
\end{itemize}}
\end{multicols}

\par\noindent The following packages \emph{must not} be loaded.

\begin{multicols}{4}
{\color{DarkRed}
\begin{itemize}
\item \path{SIstyle}
\item \path{SIunits}
\item \path{a4wide}
\item \path{a4}
\item \path{aecompl}
\item \path{ae}
\item \path{caption2}
\item \path{courier}
\item \path{doublespace}
\item \path{epsfig}
\item \path{epsf}
\item \path{euler}
\item \path{fancyhdr}
\item \path{fancyheadings}
\item \path{fourier}
\item \path{geometry}
\item \path{glossary}
\item \path{helvet}
\item \path{isolatin}
\item \path{mathpple}
\item \path{mathptmx}
\item \path{mathptm}
\item \path{newtxmath}
\item \path{newtxtext}
\item \path{palatino}
\item \path{psfig}
\item \path{pslatex}
\item \path{scrpage}
\item \path{subfigure}
\item \path{subfig}
\item \path{t1enc}
\item \path{times}
\item \path{umlaut}
\item \path{utopia}
\item \path{zefonts}
\end{itemize}}
\end{multicols}
%
\noindent 
Additionally, any package that changes the font must not be loaded.

\acks
We want to thank you for using this class to prepare articles for The Art, Science, and Engineering of Programming.
\printbibliography

\appendix
\section{A Famous Filler Text}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor
incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis
nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu
fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in
culpa qui officia deserunt mollit anim id est laborum.

\section{Submission Checklist}
\begingroup
\smaller
\begin{itemize}
\renewcommand*\labelitemi{\ensuremath{\square}}
\item All authors are listed, with affiliation
\item The abstract is not longer than 500 words.
\item The abstract states, in this order, \emph{Context},
  \emph{Inquiry}, \emph{Approach}, \emph{Knowledge}, \emph{Grounding}, and
  \emph{Importance} of the submission.
\item The ACM CCS 2012 classification is included.
\item A list of relevant keywords is provided.
\item One \emph{perspective} has been chosen.
\item One or more \emph{area(s) of submission} have been chosen.
\item The license has been acknowledged or changed to fit.
\item The title page does not spill onto the second page.
\item The running title on even pages' heads is not wider than the text block.
\item The running list of author on odd pages' heads is not wider than the text
  block.
\item The only text encoding used is \textsc{utf-8}.
\item The fonts, margins, and spacings are unchanged.
\item All \textsc{url}s are marked using the \lstinline|\url| command.
\item All units are separated from their values with a half space, consistently
  (see ~\cite{siunitx}).
\item Bold font is not used (or only in exceptional circumstances).
\item Underlining is not used (no exceptions).
\item Bold or bright colors are only ever used sparingly.
\item The captions of (a) tables, (b) listings, and (c) algorithms are
  \emph{above} the content, flush left.
\item The captions of (a) figures, (b) charts, (c) combined content, and
  similar are \emph{below} the content, flush left.
\item %
  \begin{minipage}[t]{.4\linewidth}
    Tables adhere to best practice layout~\cite{booktabs}. The do not contain
  vertical lines and only few horizontal lines.
  \end{minipage}
  {\renewcommand{\arraystretch}{.6}
    \hspace{2em} \textcolor{DarkGreen}{Yes} \hspace{1em}
      \begin{tabular}[t]{ll}\noalign{\vspace*{-2.2ex}}
        \toprule
        \emph{a} & \emph{b}\\
        \midrule
        x & abc \\
        y & def \\
        \bottomrule
      \end{tabular}
      \hspace{2em} \textcolor{DarkRed}{No} \hspace{1em}
      \begin{tabular}[t]{|l|l|}\noalign{\vspace*{-1.6ex}}
        \hline
        \emph{a} & \emph{b}\\
        \hline\hline
        x & abc \\\hline
        y & def \\\hline
      \end{tabular}}
\item All fonts are embedded. This includes the fonts for any included
  (\textsc{pdf}-)graphics.
\item All graphics are either vector graphics or have at least 300\,dpi.
  \item Acknowledgements, including statement of funding bodies or projects, is
  placed before the references, using the \lstinline|\acks| command.
\item Citations in the text are preceded by a non-breaking space.
  (\(\rightarrow\) Citations do not appear at the beginning of a line)
\item The bibliography lists all cited works, all authors are given (no
  \emph{et. al.}), the full names of the authors are used.
\item All bibliography entries have a \textsc{doi} (unless not applicable, such as for
  slides or unpublished work).
\item All bibliography entires show an \textsc{isbn}, if they have one assigned.
\item All bibliography entires that represent web sites have a \textsc{url} and
  an \emph{access date}.
\item No bibliography entry have duplicate information in \textsc{url} and
  \textsc{doi}.
\end{itemize}
% 
For the final phase (or camera-ready copy), additionally:
\begin{itemize}
\renewcommand*\labelitemi{\ensuremath{\square}}
\renewcommand*\labelitemii{\ensuremath{\square}}
\item The \lstinline|\paperdetails| are included as provided by the journal.
\item The \textsc{doi}, and submission and publication date appear on the title
  page.
\item Neither the \emph{perspective} nor the \emph{area of submission} appear
  on the title.
\end{itemize}

\endgroup

\end{document}

% Local Variables:
% TeX-engine: luatex
% End:

### 2019-12-27 compile-html.sh

@@ -1,0 +1,3 @@
+set -e
+
+cat html.sections | xargs pandoc --filter pandoc-citeproc -s -o paper.html --css base.css

### 2019-12-27 compile-latex.sh

@@ -1,0 +1,7 @@
+set -e
+
+cat latex.sections | xargs pandoc --filter pandoc-citeproc -s -o paper.tex --template=templates/pandoc-template-programming.pandoc --biblatex
+pdflatex -interaction=batchmode paper.tex
+biber paper
+pdflatex -interaction=batchmode paper.tex
+pdflatex -interaction=batchmode paper.tex

### 2019-12-27 watch.sh

@@ -1,0 +1,7 @@
+set -e
+
+ls *.md | entr ./compile-html.sh &
+browser-sync start --server --files paper.html --no-notify --no-open --port 9000 &
+
+open "http://localhost:9000/paper.html"
+

### 2019-12-27 paper.tex
basic format set up

@@
 
-\title{Wildcard: Browser Extensions for All}
+\title{Wildcard: a tool for modifying websites without programming}
 \subtitle{}% optional
@@
 \begin{abstract}
+Browser extensions and user scripts can modify websites in useful
+ways---ranging from blocking ads to adding entire new features to
+Gmail---but many people have unique needs that aren't met by existing
+extensions. Today, most of those people are stuck. They can't build
+their own browser extensions without learning how to program, so they
+have no choice but to accept the way the software was built. What if
+things were different?
 
+Wildcard is a platform that empowers anyone to build browser extensions
+and modify websites to meet their own specific needs. Wildcard shows a
+simplified view of the data in a web page as a familiar table view.
+People can directly manipulate the table to sort/filter content, add
+annotations, and even use spreadsheet-style formulas to pull in data
+from other websites. The key idea is that a table is a powerful, simple,
+and familiar paradigm for modifying a website.
 \end{abstract}
@@
 
-This is the introduction of the paper.
+People have complaints about web apps they use, but they rarely modify
+those apps to meet their needs. Why not? Some guesses:
 
-\hypertarget{demos}{%
-\section{Demos}\label{demos}}
+\begin{itemize}
+\tightlist
+\item
+  Technical skill: Most people don't know how to use Javascript and
+  manipulate the DOM.
+\item
+  Low ROI: It takes a long time to reverse engineer a site, and it's
+  usually not worth it
+\end{itemize}
+
+These seem reasonable but they might not tell the whole story.
+
+There have been many research projects that try to address these
+hurdles. For example, Chickenfoot allows for people to more quickly
+modify sites without using Javascript or dealing with the DOM, directly
+addressing both of these barriers. But, a decade later, not many people
+have ended up using systems like this.
+
+I know how to do web programming, and yet I rarely modify my apps.
+Sometimes it actually turns out to be pretty easy to hack on a site once
+I start doing it so the ROI is actually pretty high. These reasons don't
+seem to fully explain my behavior.
+
+A disheartening explanation might be that most people just don't care
+enough to make changes. This claim might be true in the current context,
+but it's also important to remember that motivation is connected to
+culture and to the space of possibilities provided by our tools. Most
+people probably didn't want to write letters before mass literacy.
+
+One interesting explanation is that \textbf{people can't estimate the
+difficulty of a change}. It's not motivating to think about what changes
+I might want when I don't know whether they would take a few minutes or
+are completely impossible. For me, an engineer, this gets at the heart
+of the issue. I know I could probably implement the change, but it's so
+hard to estimate how long it would take (minutes? months?) that I don't
+even bother trying. In general, when something seems expensive to do (or
+even possibly expensive), it can discourage casual lightweight
+experimentation.
+
+So, perhaps to encourage people to casually modify software,
+\textbf{apps need to provide more consistent affordances indicating what
+changes are possible and easy.} Maybe if it were more obvious that
+certain types of changes could be achieved in mere minutes, programmers
+and non-programmers would end up modifying our software more.
+
+\hypertarget{the-wildcard-platform}{%
+\subsection{The Wildcard platform}\label{the-wildcard-platform}}
+
+Wildcard adds a panel to the bottom of a web page that shows a
+structured table view of some of the core data in the page. When the
+user manipulates the table, the original page also gets modified. We aim
+to make the mapping between the table and the page as direct and
+intuitive as possible.
+
+There's somewhat of a tension here between directness and structure.
+Manipulating the original page itself might seem most ``direct.'' But
+the whole problem we're dealing with is that the original web page
+doesn't provide affordances for end-user modification, and there's no
+consistent structure that people can learn to work with across many
+sites.
+
+The table view is perhaps one hop less ``direct,'' but in turn provides
+other benefits. Because many people are already familiar with
+spreadsheets, they can quickly intuit which changes are easy to make in
+this system. Also, people can learn to work with the same consistent
+spreadsheet view across many sites. This consistency is very
+important---learning a generic tool that can be applied to many
+different specific cases is very powerful, and connected to the idea of
+literacy in a medium (just ask anyone who has used vim, or spreadsheets,
+or a pencil and paper). Wildcard aims to balance this tension, with a
+workflow that involves both the table view and the original webpage.
+
+Wildcard is fairly general and can support many useful changes to
+websites, which will be demoed later:
+
+\begin{itemize}
+\tightlist
+\item
+  sorting and filtering data: eg sorting shopping results
+\item
+  using 3rd-party APIs and performing small computations to add new
+  data, in the style of ``web mashups'': eg adding walkability scores to
+  hotel listings
+\item
+  adding private user annotations to the page: eg taking notes on
+  different options
+\item
+  using alternate UI widgets to enter data into a page: eg using a
+  personal datepicker widget with private calendar data, to enter the
+  right dates for taking a flight
+\end{itemize}
+
+The overall goal is to provide generic tools that fit well with the
+table paradigm and enable many specific useful changes. But it's
+important to note that Wildcard doesn't aim to provide maximum coverage
+all the possible ways someone might want to modify a web page. Rather,
+it aims to provide a useful, simple subset of modifications, and to
+provide consistent affordances so that users confidently understand
+which modifications they can make.
+
+Eventually web apps might provide the structured data table themselves.
+In the meantime, we need some sort of adapter to make this system work
+with existing sites.
+
+Wildcard provides a system for creating a wrapper on top of existing
+websites. This wrapper defines how structured data can be extracted out
+of the page, and also how manipulating the table should modify the page.
+
+The most basic way of building these wrappers is for skilled programmers
+to manually build and maintain them for popular sites. This approach
+beats the status quo because many people, including end users, can
+benefit from the generic wrapper and use it in many ways. This is
+different from the current world where programmers build
+use-case-specific browser extensions, and each extension has to
+implement its own interactions with the low-level DOM of a page. There's
+also a greater incentive for many people to collectively maintain a
+wrapper if it's shared.
+
+A more advanced way would be to make these wrappers partially or totally
+automated, and enable end users to create them. This future work could
+leverage existing research on wrapper induction but isn't the focus of
+the current work.
+
+Again, ultimately we hope that first-party sites would find it
+beneficial and straightforward to provide a structured data view
+themselves. Wildcard doesn't require sites to expose some complex
+Semantic Web schema; it merely asks for a simple structured data view.
+
+\hypertarget{example-use-cases}{%
+\section{Example use cases}\label{example-use-cases}}
 

### 2019-12-27 compile-html.sh

@@ -1,3 +1,4 @@
 set -e
 
-cat html.sections | xargs pandoc --filter pandoc-citeproc -s -o paper.html --css base.css
+cat html.sections | xargs pandoc \
+  --filter pandoc-citeproc -s -o paper.html --css base.css --toc --variable=toc-title:"Contents" --template=templates/pandoc-template-html.html

### 2019-12-27 watch.sh

@@ -1,5 +1,8 @@
 set -e
 
+# kill any existing servers
+ps aux | grep node | grep browser-sync | awk '{ print $2}' | xargs kill
+
 ls *.md | entr ./compile-html.sh &
 browser-sync start --server --files paper.html --no-notify --no-open --port 9000 &
 

### 2019-12-28 paper.tex
rework deployment for my website

@@
 
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
 \begin{document}
 
-\title{Wildcard: a tool for modifying websites without programming}
+\title{Wildcard: a tool for modifying websites by directly manipulating
+data tables}
 \subtitle{}% optional
@@
 
+Todos in intro:
+
+\begin{itemize}
+\tightlist
+\item
+  hook with Airbnb sort
+\item
+  add more reasons
+\item
+  make this shorter
+\end{itemize}
+
 People have complaints about web apps they use, but they rarely modify
@@
 There have been many research projects that try to address these
-hurdles. For example, Chickenfoot allows for people to more quickly
-modify sites without using Javascript or dealing with the DOM, directly
-addressing both of these barriers. But, a decade later, not many people
-have ended up using systems like this.
+hurdles. For example, Chickenfoot \autocite{miller2010} allows for
+people to more quickly modify sites without using Javascript or dealing
+with the DOM, directly addressing both of these barriers. But, a decade
+later, not many people have ended up using systems like this.
+
+\hypertarget{why-dont-people-tweak-websites}{%
+\subsubsection{Why don't people tweak
+websites?}\label{why-dont-people-tweak-websites}}
 
@@
 \hypertarget{the-wildcard-platform}{%
-\subsection{The Wildcard platform}\label{the-wildcard-platform}}
+\subsubsection{The Wildcard platform}\label{the-wildcard-platform}}
 
@@
 
+For example, here we open up a table view that corresponds to search
+results on the Airbnb travel site.
+
+Todos here:
+
+\begin{itemize}
+\tightlist
+\item
+  a better demo is actually sorting too at this point too, to show the
+  bidirectional nature
+\item
+  hint at future possibilities
+\item
+  introduce an image fallback for the PDF version
+\end{itemize}
+
 There's somewhat of a tension here between directness and structure.
@@
 sites.
+
+\hypertarget{directness}{%
+\subsubsection{Directness}\label{directness}}
 
@@
 
+\hypertarget{the-big-picture}{%
+\subsubsection{The big picture}\label{the-big-picture}}
+
 Eventually web apps might provide the structured data table themselves.
@@
 
-\hypertarget{example-use-cases}{%
-\section{Example use cases}\label{example-use-cases}}
+\hypertarget{demo}{%
+\section{Demo}\label{demo}}
+
+\begin{itemize}
+\tightlist
+\item
+  Sorting search results:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Airbnb took away search
+  \item
+    once you see the tabular view, there's an obvious interaction
+    available to anyone familiar with tables
+  \end{itemize}
+\item
+  Injecting new data into the page
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Take own custom notes, saved in the browser. Maybe shared in the
+    future
+  \end{itemize}
+\item
+  Formulas
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Compute values to inject into the page
+  \item
+    TBD: styling
+  \end{itemize}
+\item
+  Using a custom date picker
+\end{itemize}
+
+\hypertarget{design-principles}{%
+\section{Design principles}\label{design-principles}}
+
+\hypertarget{expose-generic-structure}{%
+\subsection{Expose generic structure}\label{expose-generic-structure}}
+
+\begin{itemize}
+\tightlist
+\item
+  Consistent across all sites, can gain familiarity
+\item
+  Nightmare bike: users learn the structure they can exploit
+\end{itemize}
+
+\hypertarget{no-api-needed}{%
+\subsection{No API needed}\label{no-api-needed}}
+
+\begin{itemize}
+\tightlist
+\item
+  3p-only is enough. 1p help is optional.
+\item
+  Can even do things the 1p didn't want to expose.
+\end{itemize}
+
+\hypertarget{in-place-toolchain}{%
+\subsection{In-place toolchain}\label{in-place-toolchain}}
+
+\begin{itemize}
+\tightlist
+\item
+  meet the user where they are, in the page
+\item
+  very native to the web style (dev tools is kinda like this)
+\end{itemize}
+
+\hypertarget{decouple-ui-from-data}{%
+\subsection{Decouple UI from data}\label{decouple-ui-from-data}}
+
+\begin{itemize}
+\tightlist
+\item
+  bring your own views of the data
+\item
+  bring your own widgets for data entry
+\item
+  instrumental interaction
+\end{itemize}
+
+\hypertarget{analysis}{%
+\section{Analysis}\label{analysis}}
+
+Cognitive dimensions?
+
+\hypertarget{related-work}{%
+\section{Related work}\label{related-work}}
+
+\begin{itemize}
+\tightlist
+\item
+  Instrumental interaction
+\item
+  Web automation
+\item
+  Wrapper induction
+\item
+  Personal data
+\item
+  Extension helpers
+\end{itemize}
+
+\hypertarget{future-work-open-questions}{%
+\section{Future work / Open
+Questions}\label{future-work-open-questions}}
+
+\begin{itemize}
+\tightlist
+\item
+  How far does functionality go?
+
+  \begin{itemize}
+  \tightlist
+  \item
+    workflows? triggers?
+  \item
+    what can and can't be done?
+
+    \begin{itemize}
+    \tightlist
+    \item
+    \end{itemize}
+  \end{itemize}
+\item
+  automated wrappers?
+
+  \begin{itemize}
+  \tightlist
+  \item
+    lean on existing tech
+  \end{itemize}
+\item
+  usability studies
+\end{itemize}
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+\hypertarget{todos}{%
+\section{Todos}\label{todos}}
+
+\begin{itemize}
+\tightlist
+\item
+  remove bold text for the proceedings version
+\end{itemize}
 

### 2019-12-28 compile-html.sh

@@ -1,4 +1,14 @@
 set -e
 
 cat html.sections | xargs pandoc \
-  --filter pandoc-citeproc -s -o paper.html --css base.css --toc --variable=toc-title:"Contents" --template=templates/pandoc-template-html.html
+  --filter pandoc-citeproc \
+  -s \
+  -o paper.html \
+  --css wildcard/basic.css \
+  --toc \
+  --toc-depth=2 \
+  --variable=toc-title:"Contents" \
+  --template=templates/pandoc-template-html.html
+
+cp ./paper.html ~/dev/homepage/source/wildcard/index.html
+cp ./wildcard/basic.css ~/dev/homepage/source/wildcard

### 2019-12-28 compile-latex.sh

@@ -5,3 +5,4 @@ pdflatex -interaction=batchmode paper.tex
 biber paper
 pdflatex -interaction=batchmode paper.tex
 pdflatex -interaction=batchmode paper.tex
+open paper.pdf

### 2019-12-28 compile-html.sh
more tweaks

@@ -4,7 +4,7 @@ cat html.sections | xargs pandoc \
   --filter pandoc-citeproc \
   -s \
   -o paper.html \
-  --css wildcard/basic.css \
+  --css basic.css \
   --toc \
   --toc-depth=2 \
   --variable=toc-title:"Contents" \

### 2019-12-28 paper.tex
more tweaks
add a conditional rendering plugin for html and pdf

@@
 
-\title{Wildcard: a tool for modifying websites by directly manipulating
-data tables}
+\title{{[}DRAFT, please don't distribute{]} Wildcard: a tool for
+modifying websites by directly manipulating data tables}
 \subtitle{}% optional
@@
 
+Do somes tuff
+
 \begin{itemize}
@@
 
-For example, here we open up a table view that corresponds to search
-results on the Airbnb travel site.
-
 Todos here:
@@
 \end{itemize}
+
+For example, in Fig.~\ref{fig:table} we open up a table view that
+corresponds to search results on the Airbnb travel site.
+
+\begin{figure}
+\hypertarget{fig:table}{%
+\centering
+\includegraphics{media/opentable.png}
+\caption{Opening a table corresponding to search results on
+Airbnb}\label{fig:table}
+}
+\end{figure}
 

### 2019-12-28 compile-html.sh

@@ -2,6 +2,8 @@ set -e
 
 cat html.sections | xargs pandoc \
   --filter pandoc-citeproc \
+  --filter conditional-render \
+  --metadata=format:html \
   -s \
   -o paper.html \
   --css basic.css \
@@ -11,4 +13,4 @@ cat html.sections | xargs pandoc \
   --template=templates/pandoc-template-html.html
 
 cp ./paper.html ~/dev/homepage/source/wildcard/index.html
-cp ./wildcard/basic.css ~/dev/homepage/source/wildcard
+cp ./basic.css ~/dev/homepage/source/wildcard

### 2019-12-28 compile-latex.sh

@@ -1,6 +1,15 @@
 set -e
 
-cat latex.sections | xargs pandoc --filter pandoc-citeproc -s -o paper.tex --template=templates/pandoc-template-programming.pandoc --biblatex
+cat latex.sections | xargs pandoc \
+  --filter conditional-render \
+  --filter pandoc-crossref \
+  --filter pandoc-citeproc \
+  -s \
+  -o \
+  paper.tex \
+  --template=templates/pandoc-template-programming.pandoc \
+  --biblatex \
+  --metadata=format:pdf
 pdflatex -interaction=batchmode paper.tex
 biber paper
 pdflatex -interaction=batchmode paper.tex

### 2019-12-28 paper.tex
done for the day

@@
 
-\title{{[}DRAFT, please don't distribute{]} Wildcard: a tool for
-modifying websites by directly manipulating data tables}
+\title{Wildcard: Modifying web applications by directly manipulating a
+data table}
 \subtitle{}% optional
@@
 \begin{abstract}
-Browser extensions and user scripts can modify websites in useful
-ways---ranging from blocking ads to adding entire new features to
-Gmail---but many people have unique needs that aren't met by existing
-extensions. Today, most of those people are stuck. They can't build
-their own browser extensions without learning how to program, so they
-have no choice but to accept the way the software was built. What if
-things were different?
+Browser extensions and user scripts can modify web applications in
+useful ways, but many people have unique needs that aren't met by
+existing extensions. Today, most of them are stuck---they can't build
+their own browser extensions without learning how to program in
+Javascript.
 
 Wildcard is a platform that empowers anyone to build browser extensions
-and modify websites to meet their own specific needs. Wildcard shows a
-simplified view of the data in a web page as a familiar table view.
-People can directly manipulate the table to sort/filter content, add
-annotations, and even use spreadsheet-style formulas to pull in data
-from other websites. The key idea is that a table is a powerful, simple,
-and familiar paradigm for modifying a website.
+without doing traditional programming. Wildcard shows the main data from
+a web page in a table with a bidirectional connection to the original
+page. People can directly manipulate the table to sort/filter content,
+add private annotations, and use spreadsheet formulas to fetch data from
+other websites.
 \end{abstract}
@@
 
-Todos in intro:
-
-Do somes tuff
+Todos for the introduction:
 
@@
 \item
-  add more reasons
+  add more bigger picture reasons, less focus on the affordances
 \item
@@
 There have been many research projects that try to address these
-hurdles. For example, Chickenfoot \autocite{miller2010} allows for
-people to more quickly modify sites without using Javascript or dealing
-with the DOM, directly addressing both of these barriers. But, a decade
+hurdles. For example, Chickenfoot \autocite{bolin2005} allows for people
+to more quickly modify sites without using Javascript or dealing with
+the DOM, directly addressing both of these barriers. But, a decade
 later, not many people have ended up using systems like this.
-
-\hypertarget{why-dont-people-tweak-websites}{%
-\subsubsection{Why don't people tweak
-websites?}\label{why-dont-people-tweak-websites}}
 
@@
 
-There's somewhat of a tension here between directness and structure.
-Manipulating the original page itself might seem most ``direct.'' But
-the whole problem we're dealing with is that the original web page
-doesn't provide affordances for end-user modification, and there's no
-consistent structure that people can learn to work with across many
-sites.
-
-\hypertarget{directness}{%
-\subsubsection{Directness}\label{directness}}
-
-The table view is perhaps one hop less ``direct,'' but in turn provides
-other benefits. Because many people are already familiar with
-spreadsheets, they can quickly intuit which changes are easy to make in
-this system. Also, people can learn to work with the same consistent
-spreadsheet view across many sites. This consistency is very
-important---learning a generic tool that can be applied to many
-different specific cases is very powerful, and connected to the idea of
-literacy in a medium (just ask anyone who has used vim, or spreadsheets,
-or a pencil and paper). Wildcard aims to balance this tension, with a
-workflow that involves both the table view and the original webpage.
-
 Wildcard is fairly general and can support many useful changes to
@@
 
-\hypertarget{the-big-picture}{%
-\subsubsection{The big picture}\label{the-big-picture}}
-
-Eventually web apps might provide the structured data table themselves.
-In the meantime, we need some sort of adapter to make this system work
-with existing sites.
-
-Wildcard provides a system for creating a wrapper on top of existing
-websites. This wrapper defines how structured data can be extracted out
-of the page, and also how manipulating the table should modify the page.
-
-The most basic way of building these wrappers is for skilled programmers
-to manually build and maintain them for popular sites. This approach
-beats the status quo because many people, including end users, can
-benefit from the generic wrapper and use it in many ways. This is
-different from the current world where programmers build
-use-case-specific browser extensions, and each extension has to
-implement its own interactions with the low-level DOM of a page. There's
-also a greater incentive for many people to collectively maintain a
-wrapper if it's shared.
-
-A more advanced way would be to make these wrappers partially or totally
-automated, and enable end users to create them. This future work could
-leverage existing research on wrapper induction but isn't the focus of
-the current work.
-
-Again, ultimately we hope that first-party sites would find it
-beneficial and straightforward to provide a structured data view
-themselves. Wildcard doesn't require sites to expose some complex
-Semantic Web schema; it merely asks for a simple structured data view.
-
-\hypertarget{demo}{%
-\section{Demo}\label{demo}}
+\hypertarget{demo-booking-a-trip-with-wildcard}{%
+\section{Demo: Booking a trip with
+Wildcard}\label{demo-booking-a-trip-with-wildcard}}
 
@@
 
+\hypertarget{implementation}{%
+\section{Implementation}\label{implementation}}
+
+Eventually web apps might provide the structured data table themselves.
+In the meantime, we need some sort of adapter to make this system work
+with existing sites.
+
+Wildcard provides a system for creating a wrapper on top of existing
+websites. This wrapper defines how structured data can be extracted out
+of the page, and also how manipulating the table should modify the page.
+
+The most basic way of building these wrappers is for skilled programmers
+to manually build and maintain them for popular sites. This approach
+beats the status quo because many people, including end users, can
+benefit from the generic wrapper and use it in many ways. This is
+different from the current world where programmers build
+use-case-specific browser extensions, and each extension has to
+implement its own interactions with the low-level DOM of a page. There's
+also a greater incentive for many people to collectively maintain a
+wrapper if it's shared.
+
+A more advanced way would be to make these wrappers partially or totally
+automated, and enable end users to create them. This future work could
+leverage existing research on wrapper induction but isn't the focus of
+the current work.
+
 \hypertarget{design-principles}{%
@@
 
-\hypertarget{expose-generic-structure}{%
-\subsection{Expose generic structure}\label{expose-generic-structure}}
+Here are some of the ideas behind the design of Wildcard. We hope these
+can be useful principles not only for Wildcard, but also for other tools
+that enable end users to modify software.
+
+\hypertarget{expose-unified-structure-across-applications}{%
+\subsection{Expose unified structure across
+applications}\label{expose-unified-structure-across-applications}}
+
+In \emph{Changing Minds} \autocite{disessa2000}, Andrea diSessa
+critiques the design of modern software with a story about a
+hypothetical ``nightmare bike.'' Each gear on the nightmare bike is
+labeled not with a number, but with an icon describing its intended use:
+smooth pavement uphill, smooth pavement downhill, gravel, etc. By some
+logic, this is more ``user-friendly'' than numbered gears, but in fact,
+hiding orderly structure from the user makes it more difficult to
+operate the bike. Many modern software designs fall into this trap,
+teaching users to use isolated modes rather than coherent structure, and
+the problem gets far worse when operating across multiple applications.
+Unlike the UNIX philosophy of small tools interoperating through shared
+abstractions, in modern computing each application is in its own silo of
+data and functionality.
+
+Wildcard helps people understand and modify the behavior of applications
+through the lens of a consistent abstraction: a data table. This
+abstraction strikes a balance between being both simple and generic. A
+data table is simpler than the DOM tree that describes the details of
+the UI, but is also generic enough to describe the essence of many
+different kinds of applications.
+
+Exposing a unified higher-level abstraction is a deliberate choice, and
+is not the only way to enable users to program without directly using
+the DOM. Chickenfoot \autocite{bolin2005} and CoScripter
+\autocite{leshed2008} allow users to create scripts that resemble
+natural language and ``sloppily'' match queries to elements in the DOM.
+These designs allow for a wide range of operations, but don't explicitly
+indicate what operations are possible---the user must look at the
+original page and imagine the possibilities. In contrast, Wildcard
+provides affordances that clearly suggest the availability of certain
+types of modifications. In addition to giving users more certainty about
+whether a modification is possible, these affordances might give users
+new ideas for things to try. Because people are not used to modifying
+web applications, providing inspiration is an important goal. (Todo:
+Perhaps something to cite here, re: discoverability in GUIs vs CLIs?)
+
+\hypertarget{direct-manipulation-by-proxy}{%
+\subsection{Direct manipulation by
+proxy}\label{direct-manipulation-by-proxy}}
+
+In Wildcard, users don't interact with the original page to modify it;
+instead they manipulate an alternate representation of the data in the
+page. The tool uses visual cues to indicate the mapping between the
+table representation and the original page. This is similar to the way
+that browser developer tools use in-page highlighting to indicate the
+mapping between HTML and the original page.
+
+\begin{figure}
+\hypertarget{fig:devtools}{%
+\centering
+\includegraphics{media/devtools.png}
+\caption{The Chrome Dev Tools use highlighting to show the mapping
+between HTML code and the page}\label{fig:devtools}
+}
+\end{figure}
+
+We considered alternate approaches where the user would interact more
+directly with the original page, but decided that the table view
+provided a consistent
 
@@
 \item
-  Consistent across all sites, can gain familiarity
+  True DM might be right on the page itself
+
+  \begin{itemize}
+  \tightlist
   \item
-  Nightmare bike: users learn the structure they can exploit
+    There are good UIs for this, eg Chrome DevTools element picker
+  \end{itemize}
+\item
+  Considered for Wildcard, but went with an indirect proxy. Still
+  maintain a close mapping.
+\item
+  Pros: consistency across sites. Avoid site-specific styling issues.
+\item
+  Cons: maybe less intuitive? More work to do the mapping, might be a
+  source of confusion.
 \end{itemize}
 
-\hypertarget{no-api-needed}{%
-\subsection{No API needed}\label{no-api-needed}}
+\hypertarget{first-party-optional}{%
+\subsection{First party optional}\label{first-party-optional}}
 
@@
   Can even do things the 1p didn't want to expose.
+\item
+  make it easy for 1p: just ask for a data table, no fancy schema.
 \end{itemize}
 
-\hypertarget{in-place-toolchain}{%
-\subsection{In-place toolchain}\label{in-place-toolchain}}
+\hypertarget{built-for-the-web}{%
+\subsection{Built for the web}\label{built-for-the-web}}
 
@@
 \item
-  meet the user where they are, in the page
+  web has the right foundations:
+
+  \begin{itemize}
+  \tightlist
   \item
-  very native to the web style (dev tools is kinda like this)
+    the right platform
+  \item
+    open platform
+  \item
+    dev tools
+  \item
+    originally intended to be
+  \end{itemize}
+\item
+  in-place toolchain: meet the user where they are, in the page
 \end{itemize}
@@
 \end{itemize}
-
-\hypertarget{analysis}{%
-\section{Analysis}\label{analysis}}
-
-Cognitive dimensions?
 
@@
 
-\hypertarget{future-work-open-questions}{%
-\section{Future work / Open
-Questions}\label{future-work-open-questions}}
+\hypertarget{future-work}{%
+\section{Future work}\label{future-work}}
 
@@
   \item
-    what can and can't be done?
-
-    \begin{itemize}
-    \tightlist
-    \item
-    \end{itemize}
+    what can and can't be done
   \end{itemize}
@@
 
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-\hypertarget{todos}{%
-\section{Todos}\label{todos}}
+\hypertarget{meta-todos}{%
+\section{meta: todos}\label{meta-todos}}
 
@@
   remove bold text for the proceedings version
+\item
+  go through the Programming proceedings template checklist
 \end{itemize}

### 2019-12-28 compile-html.sh

@@ -1,8 +1,8 @@
 set -e
 
 cat html.sections | xargs pandoc \
-  --filter pandoc-citeproc \
   --filter conditional-render \
+  --filter pandoc-citeproc \
   --metadata=format:html \
   -s \
   -o paper.html \

### 2019-12-29 paper.tex
intro

@@
 %% Second parameter: use 'submission' for initial submission, remove it for camera-ready (see 5.1)
+
+\usepackage{longtable,booktabs}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 
@@
 
-\title{Wildcard: Modifying web applications by directly manipulating a
+\title{Wildcard: End user modification of web applications through a
 data table}
@@
 useful ways, but many people have unique needs that aren't met by
-existing extensions. Today, most of them are stuck---they can't build
-their own browser extensions without learning how to program in
-Javascript.
+existing extensions. Today, most of those people can't build their own
+browser extensions, since they can't program in Javascript.
 
@@
 without doing traditional programming. Wildcard shows the main data from
-a web page in a table with a bidirectional connection to the original
-page. People can directly manipulate the table to sort/filter content,
-add private annotations, and use spreadsheet formulas to fetch data from
-other websites.
+a web page in a table that maintains a bidirectional connection to the
+original page. By directly manipulating the table, people can
+sort/filter content, add private annotations, and use spreadsheet
+formulas to fetch data from other websites.
 \end{abstract}
@@
 
-Todos for the introduction:
+In 2012, the travel site Airbnb removed the ability to sort listings by
+price. Users could still filter down to a price range, but could no
+longer view the cheapest listings first. Many users complained on online
+message boards that the change seemed hostile to users---``It's so
+frustrating!..What is the logic behind not having this function?'' says
+one user on the
+\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
+support forum}. Alas, the feature remains missing to this day.
 
-\begin{itemize}
-\tightlist
-\item
-  hook with Airbnb sort
-\item
-  add more bigger picture reasons, less focus on the affordances
-\item
-  make this shorter
-\end{itemize}
+This is a familiar situation in a world of web applications that are
+frequently updated without user consent. When web software doesn't quite
+meet users' needs, their only recourse is to complain to the developers
+and hope someone listens. Sometimes there is a way to use a browser
+extension or user script to fix the issue, but making an extension
+requires sophisticated programming skills. We see this as a waste of the
+pliability of software and the openness of the Web. If users could make
+tweaks to web applications without programming, a vast number of people
+would gain the agency to modify software to meet their unique
+preferences.
 
-People have complaints about web apps they use, but they rarely modify
-those apps to meet their needs. Why not? Some guesses:
+In this paper, we introduce a system called Wildcard\footnote{Wildcard
+  was the original internal name for Hypercard, which pioneered both
+  user modifiable software and the modern Web.} that aims to meet this
+need. Wildcard adds a panel to the bottom of a web page that shows a
+structured table view of some of the main data in the page. The table
+maintains a bidirectional connection to the original page---when the
+user manipulates the table, the original page also gets modified, and
+vice versa.
 
-\begin{itemize}
-\tightlist
-\item
-  Technical skill: Most people don't know how to use Javascript and
-  manipulate the DOM.
-\item
-  Low ROI: It takes a long time to reverse engineer a site, and it's
-  usually not worth it
-\end{itemize}
+In Wildcard, sorting Airbnb listings by price takes just one click on a
+table header, with no programming required by the end user. Beyond
+sorting and filtering data, Wildcard also supports a broad set of other
+use cases like accessing third party APIs, performing small
+computations, recording private user annotations to the page, and using
+alternate UI widgets to enter data.
 
-These seem reasonable but they might not tell the whole story.
+Wildcard currently requires a programmer to manually define an adapter
+between the original page and the data table using web scraping
+techniques. Even though programming is still required for part of the
+process, this forms an effective collaboration between the programmer
+and the end user, since the end user is able to perform a wide variety
+of modifications to a site once a single adapter is defined. Eventually,
+we envision adapter creation being less of a burden by making it
+automated or making it possible for end users to build adapters.
 
-There have been many research projects that try to address these
-hurdles. For example, Chickenfoot \autocite{bolin2005} allows for people
-to more quickly modify sites without using Javascript or dealing with
-the DOM, directly addressing both of these barriers. But, a decade
-later, not many people have ended up using systems like this.
+\hypertarget{demo-booking-a-trip-with-wildcard}{%
+\section{Demo: Booking a trip with
+Wildcard}\label{demo-booking-a-trip-with-wildcard}}
 
-I know how to do web programming, and yet I rarely modify my apps.
-Sometimes it actually turns out to be pretty easy to hack on a site once
-I start doing it so the ROI is actually pretty high. These reasons don't
-seem to fully explain my behavior.
-
-A disheartening explanation might be that most people just don't care
-enough to make changes. This claim might be true in the current context,
-but it's also important to remember that motivation is connected to
-culture and to the space of possibilities provided by our tools. Most
-people probably didn't want to write letters before mass literacy.
-
-One interesting explanation is that \textbf{people can't estimate the
-difficulty of a change}. It's not motivating to think about what changes
-I might want when I don't know whether they would take a few minutes or
-are completely impossible. For me, an engineer, this gets at the heart
-of the issue. I know I could probably implement the change, but it's so
-hard to estimate how long it would take (minutes? months?) that I don't
-even bother trying. In general, when something seems expensive to do (or
-even possibly expensive), it can discourage casual lightweight
-experimentation.
-
-So, perhaps to encourage people to casually modify software,
-\textbf{apps need to provide more consistent affordances indicating what
-changes are possible and easy.} Maybe if it were more obvious that
-certain types of changes could be achieved in mere minutes, programmers
-and non-programmers would end up modifying our software more.
-
-\hypertarget{the-wildcard-platform}{%
-\subsubsection{The Wildcard platform}\label{the-wildcard-platform}}
-
-Wildcard adds a panel to the bottom of a web page that shows a
-structured table view of some of the core data in the page. When the
-user manipulates the table, the original page also gets modified. We aim
-to make the mapping between the table and the page as direct and
-intuitive as possible.
-
-Todos here:
-
-\begin{itemize}
-\tightlist
-\item
-  a better demo is actually sorting too at this point too, to show the
-  bidirectional nature
-\item
-  hint at future possibilities
-\item
-  introduce an image fallback for the PDF version
-\end{itemize}
+Let's see an example of how an end user might modify a website with
+Wildcard in practice.
 
@@
 \end{figure}
-
-Wildcard is fairly general and can support many useful changes to
-websites, which will be demoed later:
-
-\begin{itemize}
-\tightlist
-\item
-  sorting and filtering data: eg sorting shopping results
-\item
-  using 3rd-party APIs and performing small computations to add new
-  data, in the style of ``web mashups'': eg adding walkability scores to
-  hotel listings
-\item
-  adding private user annotations to the page: eg taking notes on
-  different options
-\item
-  using alternate UI widgets to enter data into a page: eg using a
-  personal datepicker widget with private calendar data, to enter the
-  right dates for taking a flight
-\end{itemize}
-
-The overall goal is to provide generic tools that fit well with the
-table paradigm and enable many specific useful changes. But it's
-important to note that Wildcard doesn't aim to provide maximum coverage
-all the possible ways someone might want to modify a web page. Rather,
-it aims to provide a useful, simple subset of modifications, and to
-provide consistent affordances so that users confidently understand
-which modifications they can make.
-
-\hypertarget{demo-booking-a-trip-with-wildcard}{%
-\section{Demo: Booking a trip with
-Wildcard}\label{demo-booking-a-trip-with-wildcard}}
 
@@
 
-\hypertarget{implementation}{%
-\section{Implementation}\label{implementation}}
+\hypertarget{system-implementation}{%
+\section{System Implementation}\label{system-implementation}}
 
-Eventually web apps might provide the structured data table themselves.
-In the meantime, we need some sort of adapter to make this system work
-with existing sites.
+\emph{Goal of this section: briefly explain the current implementation.
+Enough detail to ground further discussion, but don't dwell on it. It's
+not the main point, and will evolve a lot.}
 
-Wildcard provides a system for creating a wrapper on top of existing
-websites. This wrapper defines how structured data can be extracted out
-of the page, and also how manipulating the table should modify the page.
+\begin{itemize}
+\tightlist
+\item
+  Built as a Greasemonkey script for now (\emph{todo: convert to a full
+  broser extension?})
+\item
+  Describe the adapter API
 
-The most basic way of building these wrappers is for skilled programmers
-to manually build and maintain them for popular sites. This approach
-beats the status quo because many people, including end users, can
-benefit from the generic wrapper and use it in many ways. This is
-different from the current world where programmers build
-use-case-specific browser extensions, and each extension has to
-implement its own interactions with the low-level DOM of a page. There's
-also a greater incentive for many people to collectively maintain a
-wrapper if it's shared.
+  \begin{itemize}
+  \tightlist
+  \item
+    programmer creates adapters for now
+  \item
+    maybe end users can create in the future, or automated
+  \item
+    show a snippet of adapter code
+  \end{itemize}
+\item
+  Mention the technique of scraping data from AJAX requests
+\item
+  future implementation goals
 
-A more advanced way would be to make these wrappers partially or totally
-automated, and enable end users to create them. This future work could
-leverage existing research on wrapper induction but isn't the focus of
-the current work.
+  \begin{itemize}
+  \tightlist
+  \item
+    make it easy for programmers to add adapters + plugins, and
+    distribute them to users. (Currently all adapters + plugins are part
+    of the main Wildcard codebase)
+  \end{itemize}
+\end{itemize}
 
@@
 
-Here are some of the ideas behind the design of Wildcard. We hope these
-can be useful principles not only for Wildcard, but also for other tools
-that enable end users to modify software.
+Below are some of the ideas behind the design of Wildcard. We hope that
+these principles are also useful for thinking more broadly about
+enabling end users to modify software.
 
@@
 applications}\label{expose-unified-structure-across-applications}}
+
+\emph{Todo: add a diagram here?}
 
@@
 
-In Wildcard, users don't interact with the original page to modify it;
-instead they manipulate an alternate representation of the data in the
-page. The tool uses visual cues to indicate the mapping between the
-table representation and the original page. This is similar to the way
-that browser developer tools use in-page highlighting to indicate the
-mapping between HTML and the original page.
+In Wildcard, users manipulate a data table added to the page, rather
+than the original page. Although the interaction with the table is
+direct like using a spreadsheet, the interaction with the page is
+indirectly mediated through the table.
+
+We considered other approaches where the user would interact more
+directly with the original UI, e.g.~injecting sort controls into the
+page, but decided that the consistency and affordances of the table view
+outweighed the potential confusion of a new layer of indirection.
+
+Making this design successful requires maintaining a close mapping in
+the user's mind between the new representation and the original page
+(cite Norman? Cognitive Dimensions?). Wildcard provides live visual cues
+as the user navigates the data table, similar to the highlighting
+provided by browser developer tools to indicate the mapping between HTML
+and the original page.
 
@@
 
-We considered alternate approaches where the user would interact more
-directly with the original page, but decided that the table view
-provided a consistent
-
-\begin{itemize}
-\tightlist
-\item
-  True DM might be right on the page itself
-
-  \begin{itemize}
-  \tightlist
-  \item
-    There are good UIs for this, eg Chrome DevTools element picker
-  \end{itemize}
-\item
-  Considered for Wildcard, but went with an indirect proxy. Still
-  maintain a close mapping.
-\item
-  Pros: consistency across sites. Avoid site-specific styling issues.
-\item
-  Cons: maybe less intuitive? More work to do the mapping, might be a
-  source of confusion.
-\end{itemize}
-
-\hypertarget{first-party-optional}{%
-\subsection{First party optional}\label{first-party-optional}}
-
-\begin{itemize}
-\tightlist
-\item
-  3p-only is enough. 1p help is optional.
-\item
-  Can even do things the 1p didn't want to expose.
-\item
-  make it easy for 1p: just ask for a data table, no fancy schema.
-\end{itemize}
-
-\hypertarget{built-for-the-web}{%
-\subsection{Built for the web}\label{built-for-the-web}}
-
-\begin{itemize}
-\tightlist
-\item
-  web has the right foundations:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    the right platform
-  \item
-    open platform
-  \item
-    dev tools
-  \item
-    originally intended to be
-  \end{itemize}
-\item
-  in-place toolchain: meet the user where they are, in the page
-\end{itemize}
-
 \hypertarget{decouple-ui-from-data}{%
@@
 
-\begin{itemize}
-\tightlist
-\item
-  bring your own views of the data
-\item
-  bring your own widgets for data entry
-\item
-  instrumental interaction
-\end{itemize}
+Most software does not provide much user choice for interface elements,
+even for common data types: e.g., the website provides a datepicker
+widget, and you have no ability to provide your own replacement
+datepicker. This forces users to learn many different interfaces for
+similar tasks, some of them worse than others. It also prevents
+combining data across multiple applications within a single user
+interface.
+
+Wildcard exposes the data underlying an application's UI, allowing the
+user to use interfaces of their choice to view and modify the data. The
+Expedia datepicker demo above showed one example of how this can be
+useful, but we also envision creating other widgets for visualizing and
+editing data. Some examples would be showing geographic data in a custom
+map that includes the user's own annotations, or editing a blog post in
+a rich text editor of the user's choice.
+
+One benefit of decoupling data from interfaces is improved UI quality.
+When UI widgets can compete on their merits rather than based on network
+effects from the data they have access to, it creates much stronger
+competition at the interface level. For example, there is competition
+among email clients (which consume an open protocol), but not among
+Facebook or Twitter clients. This benefit relates to the SOLID project
+led by Tim Berners-Lee \autocite{berners-lee2018}, which envisions
+user-controlled data as a mechanism for decoupling data from interfaces,
+e.g.~giving users a choice of which client to use to consume a given
+social media feed. Wildcard has overlapping goals with SOLID, but does
+not require decentralized user control of data---the data can remain on
+a centralized server, as long as the interface can be tweaked by end
+users. \emph{Note: maybe this paragraph belongs in related work?}
+
+There is also value in using the same consistent interface across many
+applications. For example, many programmers become deeply familiar with
+one text editor and use it for many different kinds of tasks. This usage
+even extends beyond editing permanent files; shell programs can use the
+user's preferred text editor as an interactive input mechanism (e.g.~for
+editing git commit messages). This ability to generically reuse the text
+editor in many contexts makes it more worthwhile to invest time
+mastering the tool. Beaudouin-Lafon and Mackay refer to this type of
+reuse as \emph{polymorphism} in interaction
+\autocite{beaudouin-lafon2000}, noting that it is a key technique for
+keeping interfaces simple while increasing their power.
+
+\emph{Note: Maybe could relate this section to Concept Reuse?}
+
+\emph{Note: Does this section overlap with ``expose unified
+structure''?}
+
+\hypertarget{in-place-toolchain}{%
+\subsection{In-place toolchain}\label{in-place-toolchain}}
+
+\begin{longtable}[]{@{}lll@{}}
+\toprule
+\endhead
+& \textbf{In-place} & \textbf{Not in-place}\tabularnewline
+\textbf{End user friendly} & Wildcard & IFTTT\tabularnewline
+\textbf{For developers} & browser dev tools & editing OSS\tabularnewline
+\bottomrule
+\end{longtable}
+
+We believe that people are more motivated to modify software when the
+tools for modification are close at hand while using the original
+software, referred to by Ink and Switch as an ``in-place toolchain''
+\autocite{inkandswitch2019}.
+
+Browser developer tools exemplify an in-place toolchain, since they are
+built in to browsers and available in the context of viewing a page; in
+contrast, creating a browser extension requires moving out of the
+browser to a separate development environment.
+
+Wildcard creates an in-place toolchain for modifying web applications.
+Once installed, if an adapter is available for a site, it can be
+immediately modified from within the browser. Hopefully, this can help
+make switching from using to modifying software into a smooth
+transition, rather than a chasm to cross.
+
+\hypertarget{first-party-cooperation-optional}{%
+\subsection{First party cooperation
+optional}\label{first-party-cooperation-optional}}
+
+The Web is an unusually extensible platform. On many other platforms
+(e.g.~smartphone operating systems), software is locked down unless
+first-party developers explicitly provide hooks for plugins and
+interoperability, but on the Web, all client-side code is available for
+browser extensions to modify. Application authors can use practices that
+make it easier to modify their apps (e.g.~clean semantic markup), or
+more difficult (e.g.~code obfuscation), but the default state is
+openness. This gives extensions freedom to modify applications in
+creative ways that the original developers did not plan for.
+
+Wildcard takes advantage of this openness, and does not depend on
+cooperation from first-party website developers. Any programmer can add
+support for any website to Wildcard by building a third party adapter.
+This design acknowledges the pragmatic reality that most sites today
+will not have Wildcard support built in, and more broadly do not
+prioritize client side extensibility.
+
+However, we also hope that first party website developers eventually
+build in Wildcard support to their applications, which would reduce the
+burden of maintaining adapters and make Wildcard plugins more stable.
+While this might seem optimistic, implementing the Wildcard adapter API
+could be straightforward in a typical client-side application that
+already has access to a structured version of the data in the page.
+There is also precedent for first parties implementing an official
+client extension API in response to user demand: for several years,
+Google exposed an extension API in Gmail for Greasemonkey scripts to
+use. (Since then, third parties have created well-maintained
+alternatives \autocite{streak,talwar2019}, illustrating the value of
+sharing even third party adapters.)
 
@@
 \section{Related work}\label{related-work}}
+
+\emph{Note: a lot of this was already covered above; how to deal with
+that?}
 
@@
 \item
-  Personal data
+  Personal data ownsership / SOLID
 \item
-  Extension helpers
+  Extension helper libraries, e.g.~Gmail.js.
 \end{itemize}
@@
 \item
-  How far does functionality go?
+  still in early development; note the beta release plan (tentative:
+  target public beta availability at the workshop in March?)
+\item
+  What are the limits of functionality? What can't it do?
 
@@
   \item
-    workflows? triggers?
-  \item
-    what can and can't be done
+    Should we extend further? Automation workflows? Triggers?
   \end{itemize}
 \item
-  automated wrappers?
-
-  \begin{itemize}
-  \tightlist
+  Could explore automated wrapper induction, building on prior work
 \item
-    lean on existing tech
-  \end{itemize}
+  Usability studies
 \item
-  usability studies
-\end{itemize}
-
-\hypertarget{meta-todos}{%
-\section{meta: todos}\label{meta-todos}}
-
-\begin{itemize}
-\tightlist
-\item
-  remove bold text for the proceedings version
-\item
-  go through the Programming proceedings template checklist
+  Link to sign up for future updates
 \end{itemize}

### 2019-12-29 compile-latex.sh

@@ -10,7 +10,7 @@ cat latex.sections | xargs pandoc \
   --template=templates/pandoc-template-programming.pandoc \
   --biblatex \
   --metadata=format:pdf
-pdflatex -interaction=batchmode paper.tex
+pdflatex paper.tex
 biber paper
 pdflatex -interaction=batchmode paper.tex
 pdflatex -interaction=batchmode paper.tex

### 2019-12-29 paper.tex
lotsa changes

@@
 longer view the cheapest listings first. Many users complained on online
-message boards that the change seemed hostile to users---``It's so
+message boards that the change seemed hostile to users. ``It's so
 frustrating!..What is the logic behind not having this function?'' says
@@
 requires sophisticated programming skills. We see this as a waste of the
-pliability of software and the openness of the Web. If users could make
-tweaks to web applications without programming, a vast number of people
-would gain the agency to modify software to meet their unique
-preferences.
+openness of the Web platform, and of the pliability of software in
+general. If users could make tweaks to web applications without
+programming, a vast number of people would gain the agency to modify
+software to meet their unique preferences.
 
@@
 table header, with no programming required by the end user. Beyond
-sorting and filtering data, Wildcard also supports a broad set of other
+sorting and filtering data, Wildcard also supports a variety of other
 use cases like accessing third party APIs, performing small
-computations, recording private user annotations to the page, and using
-alternate UI widgets to enter data.
+computations, recording private user annotations, and using alternate UI
+widgets. While Wildcard does not support all changes someone might want
+to make to a website, the goal of the project is to make a broad subset
+of changes easily accessible to end users.
 
-Wildcard currently requires a programmer to manually define an adapter
-between the original page and the data table using web scraping
-techniques. Even though programming is still required for part of the
-process, this forms an effective collaboration between the programmer
-and the end user, since the end user is able to perform a wide variety
-of modifications to a site once a single adapter is defined. Eventually,
-we envision adapter creation being less of a burden by making it
-automated or making it possible for end users to build adapters.
+Under the hood, there is no magic---Wildcard currently requires a
+programmer to manually define an adapter between the original page and
+the data table using web scraping techniques. While programming is
+required for part of the process, this forms a useful collaboration
+between the programmer and the end user, since the end user is able to
+perform many different modifications once an adapter is defined. We also
+plan to eventually reduce the burden of making adapters (discussed in
+more detail later).
 
@@
 
-Let's see an example of how an end user might modify a website with
-Wildcard in practice.
+To get a sense of the user experience of Wildcard, let's see an example
+of how an end user can modify a website with Wildcard in practice.
 

### 2020-01-03 compile-html.sh
filled in demos, sketched out system architecture

@@ -14,3 +14,4 @@ cat html.sections | xargs pandoc \
 
 cp ./paper.html ~/dev/homepage/source/wildcard/index.html
 cp ./basic.css ~/dev/homepage/source/wildcard
+cp -r ./media ~/dev/homepage/source/wildcard

### 2020-01-03 paper.tex
updated pdf

--- 2019-12-29 paper.tex
@@
   %% The list is open-ended, so use other areas if yours is/are not listed.
-  area={Social Coding, General-purpose programming},
+  area={Programming environments, Visual and live programming},
   %% You may choose the license for your paper (see 3.)
   %% License options include: cc-by (default), cc-by-nc
-  % license=cc-by,
+  % license=cc-by-sa,
 }
@@
 \begin{abstract}
-Browser extensions and user scripts can modify web applications in
-useful ways, but many people have unique needs that aren't met by
-existing extensions. Today, most of those people can't build their own
-browser extensions, since they can't program in Javascript.
+Many people use Web applications that do not exactly meet their unique
+needs. While the Web platform supports client-side modification through
+user scripts and browser extensions, most people do not have the
+programming skills to implement such modifications.
 
-Wildcard is a platform that empowers anyone to build browser extensions
-without doing traditional programming. Wildcard shows the main data from
-a web page in a table that maintains a bidirectional connection to the
-original page. By directly manipulating the table, people can
-sort/filter content, add private annotations, and use spreadsheet
-formulas to fetch data from other websites.
+In this paper, we present a prototype of a browser extension called
+Wildcard, that empowers users to casually tweak web applications without
+programming. Wildcard shows the main data from a web page in a table,
+and maintains a bidirectional connection between the table and the
+original page. By directly manipulating the table, people can perform a
+wide variety of modifications: sorting/filtering content, adding private
+annotations, using spreadsheet formulas to fetch data from other web
+services, using custom UI elements to edit form data, and more.
+
+We present examples of using Wildcard to solve real world problems, and
+explain the design principles behind the prototype. In the future, we
+envision continuing to build Wildcard into a fully deployed system that
+makes the web into a more malleable medium.
 \end{abstract}
@@
 longer view the cheapest listings first. Many users complained on online
-message boards that the change seemed hostile to users---``It's so
-frustrating!..What is the logic behind not having this function?'' says
+message boards that the change seemed hostile to users. ``It's so
+frustrating!..What is the logic behind not having this function?'' said
 one user on the
@@
 This is a familiar situation in a world of web applications that are
-frequently updated without user consent. When web software doesn't quite
-meet users' needs, their only recourse is to complain to the developers
-and hope someone listens. Sometimes there is a way to use a browser
-extension or user script to fix the issue, but making an extension
-requires sophisticated programming skills. We see this as a waste of the
-pliability of software and the openness of the Web. If users could make
-tweaks to web applications without programming, a vast number of people
-would gain the agency to modify software to meet their unique
-preferences.
+frequently updated without user consent. For most people, when web
+software does not quite meet their needs, their only recourse is to
+complain to the developers and hope someone listens. If they know how to
+program in Javascript, perhaps they can implement a user script or a
+browser extension to patch the issue, but most people do not have these
+programming skills. While many have become accustomed to this status
+quo, we see it as a waste of the openness of the Web platform and the
+general pliability of software. In \emph{Personal Dynamic Media}, Alan
+Kay envisioned personal computing as a medium that let a user ``mold and
+channel its power to his own needs,'' but today's software is far from
+this vision.
 
-In this paper, we introduce a system called Wildcard\footnote{Wildcard
-  was the original internal name for Hypercard, which pioneered both
-  user modifiable software and the modern Web.} that aims to meet this
-need. Wildcard adds a panel to the bottom of a web page that shows a
-structured table view of some of the main data in the page. The table
-maintains a bidirectional connection to the original page---when the
-user manipulates the table, the original page also gets modified, and
+In this paper, we introduce Wildcard\footnote{Wildcard was the internal
+  pre-release name for Hypercard, which served as an inspiration for
+  this project since it promoted software modification by end users and
+  was a precursor to the modern Web.}, a browser extension that aims to
+make software more malleable by enabling users to tweak web applications
+without programming. Wildcard adds a panel to the bottom of a web page
+that shows a structured table view of the main data in the page. The
+table maintains a bidirectional connection to the original page---when
+the user manipulates the table, the original page gets modified, and
 vice versa.
 
-In Wildcard, sorting Airbnb listings by price takes just one click on a
-table header, with no programming required by the end user. Beyond
-sorting and filtering data, Wildcard also supports a broad set of other
-use cases like accessing third party APIs, performing small
-computations, recording private user annotations to the page, and using
-alternate UI widgets to enter data.
+In Wildcard, a user can sort Airbnb listings with just one intuitive
+click on a table header, with no programming required. Beyond sorting
+and filtering data, Wildcard also supports accessing third party APIs,
+performing small computations, recording private user annotations, using
+alternate UI widgets, and other useful changes. While Wildcard does not
+support all changes someone might want to make to a website, it makes
+broad subset of changes easily accessible to end users.
 
-Wildcard currently requires a programmer to manually define an adapter
-between the original page and the data table using web scraping
-techniques. Even though programming is still required for part of the
-process, this forms an effective collaboration between the programmer
-and the end user, since the end user is able to perform a wide variety
-of modifications to a site once a single adapter is defined. Eventually,
-we envision adapter creation being less of a burden by making it
-automated or making it possible for end users to build adapters.
+Under the hood, the implementation is straightforward, because a
+programmer must manually write an adapter for each individual website,
+which uses web scraping techniques to map the web page to the table.
+\emph{todo: mention the other bits of extension code too} While
+programming is required for part of the process, this is still very
+different from traditional browser extensions---instead of the
+programmer defining a narrow use case, the end user is able to make many
+different changes on top of a single site-specific adapter. Programmers
+can extend Wildcard with plugins which provide various bits of
+functionality including connecting it to specific websites and web APIs,
+but the end user never needs to do any traditional programming.
 
-\hypertarget{demo-booking-a-trip-with-wildcard}{%
-\section{Demo: Booking a trip with
-Wildcard}\label{demo-booking-a-trip-with-wildcard}}
+\emph{note the current prototype stage}
 
-Let's see an example of how an end user might modify a website with
-Wildcard in practice.
+In this paper, we present examples of using Wildcard to solve real world
+problems, and explain the design principles behind the prototype:
+
+\emph{todo: bullet point the design principles here}
+
+In the future, we envision building Wildcard into a fully deployed
+system that makes the web into a more malleable medium. \emph{todo: this
+needs something more}
+
+\hypertarget{demos}{%
+\section{Demos}\label{demos}}
+
+To get a sense of how it feels to use Wildcard, let's see an example of
+using it to help with booking a trip using the travel sites Airbnb and
+Expedia.
+
+\hypertarget{sorting-and-filtering}{%
+\subsection{Sorting and filtering}\label{sorting-and-filtering}}
+
+We start by opening up the Airbnb search listings page to look for a
+place to stay. As mentioned before, this page annoyingly doesn't let us
+sort by price, so we'll use Wildcard to fix that. First, we open up the
+Wildcard panel, which shows a table corresponding to the search results
+in the page. As we click around in the table, the corresponding row in
+the page is highlighted so we can see the connection between the views.
+
+To sort the page in ascending order by price, all we need to do is click
+on a table header to sort the table, and the original page gets sorted
+in the same order. We can also filter to only listings with a rating
+above 4.5 using a filtering UI on the column header. (Filtering by
+rating is another feature not offered in the Airbnb site.)
+
+Note how after finishing the sorting and filtering, we can close the
+table view and continue using the website in its original design. The
+Wildcard interface is better for flexibly manipulating the underlying
+data, but the original site will usually offer a more polished and
+pleasant experience for viewing the data after it's in the desired form.
+
+\hypertarget{row-actions}{%
+\subsection{Row actions}\label{row-actions}}
+
+Most websites that show tables of data also offer various actions that
+can be taken on a row in the table, like adding an item to a shopping
+cart. Wildcard has the ability to make these actions available in the
+data table, passed through by the site-specific adapter. The main
+advantage this provides to users is the ability to easily perform an
+action in bulk across multiple rows.
+
+For example, it's tedious on Airbnb to click on listings one by one to
+add them to a list of favorites, or open the listings in a new tab.
+Using Wildcard, we can just select multiple rows, right click, and then
+choose an action from the context menu to apply to all the rows.
+
+Within the site adapter, each action is implemented as a web automation
+that can do anything available in Javascript running in the context of
+the page: clicking on buttons in the UI, launching AJAX requests,
+navigating to a new page, etc.
+
+\hypertarget{user-annotations}{%
+\subsection{User annotations}\label{user-annotations}}
+
+It's often useful to annotate a web page with notes. Users can use
+Wildcard to annotate a page by adding data into a new column, which is
+then shown in context of the original page.
+
+Here, we use this feature to jot down notes on pros and cons of various
+listings:
+
+When we come back later to the site, the annotations will still be
+visible. To support this, the site adapter saves the user annotations to
+browser local storage for persistence, and associates each table row
+with a stable identifier from the original site.
+
+\hypertarget{computation-with-formulas}{%
+\subsection{Computation with formulas}\label{computation-with-formulas}}
+
+The demos so far have shown small, straightforward tweaks that provide
+useful conveniences while requiring very little effort. But Wildcard
+also supports adding more sophisticated functionality through a formula
+system.
+
+When traveling without a car, it's nice to evaluate potential places to
+stay based on how walkable the surroundings are. There's an online
+service called Walkscore that can rate the walkability of any location
+on a 1-100 scale. It would take way too much work to manually
+cross-reference the scores with Airbnb locations, but we can use
+Wildcard formulas to automatically integrate Walkscore values into the
+page.
+
+Wildcard includes a formula that uses the Walkscore API to fetch the
+score for any latitude and longitude. When we call the formula, it
+fetches the associated score and populates it into the page. By copy
+pasting the formula into all the rows on the page, we can add Walkscore
+data to all the search listings:
+
+Programmers can extend Wildcard with new formulas, which are just
+Javascript functions that take table data as input and return new data.
+Formulas can access external APIs to fetch data, but cannot have side
+effects that directly manipulate the page. Instead, they return data,
+which the user can choose to add into the page by showing a column of
+data in the page.
+
+\emph{todo: demo of composing formulas (requires a real formula
+system\ldots)}
+
+\hypertarget{custom-ui-elements}{%
+\subsection{Custom UI elements}\label{custom-ui-elements}}
+
+It might seem that Wildcard is only useful on websites that display
+lists of tabular data like search results. But in fact, the table
+metaphor is flexible enough to represent many types of data. For
+example, a flight search form on Expedia.com can be represented as a
+table with a single row:
+
+This alone doesn't provide additional value on top of the original form,
+but it becomes useful when combined with two other features of Wildcard.
+First, Wildcard offers writable cells, where edits in either the table
+or the original site propagate in both directions. Second, Wildcard
+offers UI widgets that allow users to graphically edit the value of a
+cell.
+
+Here's an example of using those features to help with filling in a
+form. When filling in dates for a flight search, it's important to
+correctly remember the planned dates for the trip. This often requires
+opening up a separate calendar application to look up the dates, and
+manually entering the dates into the form. In Wildcard, we can do this
+without manual copying, by editing the date cell directly using a
+datepicker widget that has access to our calendar information. Notice
+how the dates in the original form update when we update the table
+cells.
+
+Custom UIs enable people to use a consistent UI to enter common types of
+data across the web. They also allow a user to access their own private
+data as part of a web interface, without needing to expose that private
+data to the website.
+
+The PDF version of the output will go here.
+
+Probably make one giant full-page figure that explains the whole thing?
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Sorting search results:
+Overall, we think an interactive data table is a natural computation
+model that presents a surprisingly large range of possibilities for end
+user modification of websites. While we've presented just a small
+sampling of use cases here to concretely illustrate some of these
+possibilities, we plan to continue developing site adapters, formulas,
+and UI elements to explore more use cases, and to eventually publicly
+release the tool to allow real end users to discover their own uses.
+
+\hypertarget{system-architecture}{%
+\section{System Architecture}\label{system-architecture}}
+
+\begin{figure}
+\centering
+\includegraphics{media/architecture.png}
+\caption{The architecture of the Wildcard system}
+\end{figure}
+
+Wildcard is written in Typescript, and is injected into pages using the
+\href{https://www.tampermonkey.net/}{Tampermonkey} userscript manager
+(although in the future we plan to deploy it as a standalone browser
+extension.)
+
+In order to maximize extensibility, Wildcard is implemented as a small
+core program along with several types of plugins: site adapters,
+formulas, and cell renderers/editors. The core contains functionality
+for displaying the data table and handling user interactions, and the
+table implementation is currently built on the
+\href{https://handsontable.com/}{Handsontable} Javascript library to
+enable rapid prototyping.
+
+Site adapters\ldots.
+
+Formula functions\ldots{}
+
+Cell viewers\ldots{}
 
@@
 \item
-    Airbnb took away search
-  \item
-    once you see the tabular view, there's an obvious interaction
-    available to anyone familiar with tables
-  \end{itemize}
-\item
-  Injecting new data into the page
+  Extension points
 
@@
   \item
-    Take own custom notes, saved in the browser. Maybe shared in the
-    future
-  \end{itemize}
-\item
-  Formulas
+    site adapters
 
@@
     \item
-    Compute values to inject into the page
+      show a snippet of adapter code
     \item
-    TBD: styling
+      discuss how easy it is for programmers to make adapters
+    \item
+      discuss possible future automation of adapter creation
+    \item
+      Mention the technique of scraping data from AJAX requests
     \end{itemize}
   \item
-  Using a custom date picker
+    formula functions
+  \item
+    cell viewer/editors
   \end{itemize}
-
-\hypertarget{system-implementation}{%
-\section{System Implementation}\label{system-implementation}}
-
-\emph{Goal of this section: briefly explain the current implementation.
-Enough detail to ground further discussion, but don't dwell on it. It's
-not the main point, and will evolve a lot.}
-
-\begin{itemize}
-\tightlist
-\item
-  Built as a Greasemonkey script for now (\emph{todo: convert to a full
-  broser extension?})
-\item
-  Describe the adapter API
-
-  \begin{itemize}
-  \tightlist
-  \item
-    programmer creates adapters for now
-  \item
-    maybe end users can create in the future, or automated
-  \item
-    show a snippet of adapter code
-  \end{itemize}
-\item
-  Mention the technique of scraping data from AJAX requests
 \item
@@
 
-Below are some of the ideas behind the design of Wildcard. We hope that
-these principles are also useful for thinking more broadly about
-enabling end users to modify software.
+The design of Wildcard is grounded in several principles, informed by
+prior work and our own experimentation. We hope you find these
+principles helpful not only for understanding our prototype, but also
+for designing other systems for end user programming.
 
-\hypertarget{expose-unified-structure-across-applications}{%
-\subsection{Expose unified structure across
-applications}\label{expose-unified-structure-across-applications}}
-
-\emph{Todo: add a diagram here?}
-
-In \emph{Changing Minds} \autocite{disessa2000}, Andrea diSessa
-critiques the design of modern software with a story about a
-hypothetical ``nightmare bike.'' Each gear on the nightmare bike is
-labeled not with a number, but with an icon describing its intended use:
-smooth pavement uphill, smooth pavement downhill, gravel, etc. By some
-logic, this is more ``user-friendly'' than numbered gears, but in fact,
-hiding orderly structure from the user makes it more difficult to
-operate the bike. Many modern software designs fall into this trap,
-teaching users to use isolated modes rather than coherent structure, and
-the problem gets far worse when operating across multiple applications.
-Unlike the UNIX philosophy of small tools interoperating through shared
-abstractions, in modern computing each application is in its own silo of
-data and functionality.
-
-Wildcard helps people understand and modify the behavior of applications
-through the lens of a consistent abstraction: a data table. This
-abstraction strikes a balance between being both simple and generic. A
-data table is simpler than the DOM tree that describes the details of
-the UI, but is also generic enough to describe the essence of many
-different kinds of applications.
-
-Exposing a unified higher-level abstraction is a deliberate choice, and
-is not the only way to enable users to program without directly using
-the DOM. Chickenfoot \autocite{bolin2005} and CoScripter
-\autocite{leshed2008} allow users to create scripts that resemble
-natural language and ``sloppily'' match queries to elements in the DOM.
-These designs allow for a wide range of operations, but don't explicitly
-indicate what operations are possible---the user must look at the
-original page and imagine the possibilities. In contrast, Wildcard
-provides affordances that clearly suggest the availability of certain
-types of modifications. In addition to giving users more certainty about
-whether a modification is possible, these affordances might give users
-new ideas for things to try. Because people are not used to modifying
-web applications, providing inspiration is an important goal. (Todo:
-Perhaps something to cite here, re: discoverability in GUIs vs CLIs?)
-
-\hypertarget{direct-manipulation-by-proxy}{%
-\subsection{Direct manipulation by
-proxy}\label{direct-manipulation-by-proxy}}
-
-In Wildcard, users manipulate a data table added to the page, rather
-than the original page. Although the interaction with the table is
-direct like using a spreadsheet, the interaction with the page is
-indirectly mediated through the table.
-
-We considered other approaches where the user would interact more
-directly with the original UI, e.g.~injecting sort controls into the
-page, but decided that the consistency and affordances of the table view
-outweighed the potential confusion of a new layer of indirection.
-
-Making this design successful requires maintaining a close mapping in
-the user's mind between the new representation and the original page
-(cite Norman? Cognitive Dimensions?). Wildcard provides live visual cues
-as the user navigates the data table, similar to the highlighting
-provided by browser developer tools to indicate the mapping between HTML
-and the original page.
-
-\begin{figure}
-\hypertarget{fig:devtools}{%
-\centering
-\includegraphics{media/devtools.png}
-\caption{The Chrome Dev Tools use highlighting to show the mapping
-between HTML code and the page}\label{fig:devtools}
-}
-\end{figure}
+\emph{Todo: add diagrams illustrating principles}
 
@@
 
-Most software does not provide much user choice for interface elements,
-even for common data types: e.g., the website provides a datepicker
-widget, and you have no ability to provide your own replacement
-datepicker. This forces users to learn many different interfaces for
-similar tasks, some of them worse than others. It also prevents
-combining data across multiple applications within a single user
-interface.
+Most software does not allow users to choose their own UI elements, even
+for common data types. If a website provides a datepicker widget, you
+have no ability to provide your own replacement datepicker, with your
+preferred design or with privileged access to your private calendar
+data. This forces users to learn many different interfaces of varying
+quality for similar tasks. Some websites have APIs to allow users to
+access the data underlying the UI, but in addition to requiring
+programming, these are heavyweight tools more fit for batch exports or
+building entire new clients than for casual UI modification.
 
-Wildcard exposes the data underlying an application's UI, allowing the
-user to use interfaces of their choice to view and modify the data. The
-Expedia datepicker demo above showed one example of how this can be
+In Wildcard, a user gets access to a view of the underlying data in the
+page, and can choose their own interfaces to view and modify the data.
+The Expedia datepicker demo showed one example of how this can be
 useful, but we also envision creating other widgets for visualizing and
@@
 a centralized server, as long as the interface can be tweaked by end
-users. \emph{Note: maybe this paragraph belongs in related work?}
+users.
 
-There is also value in using the same consistent interface across many
-applications. For example, many programmers become deeply familiar with
-one text editor and use it for many different kinds of tasks. This usage
-even extends beyond editing permanent files; shell programs can use the
-user's preferred text editor as an interactive input mechanism (e.g.~for
-editing git commit messages). This ability to generically reuse the text
-editor in many contexts makes it more worthwhile to invest time
-mastering the tool. Beaudouin-Lafon and Mackay refer to this type of
-reuse as \emph{polymorphism} in interaction
-\autocite{beaudouin-lafon2000}, noting that it is a key technique for
-keeping interfaces simple while increasing their power.
+Another benefit of decoupling data from UI is that it becomes possible
+to use the same consistent interface across many applications. For
+example, many programmers become deeply familiar with one text editor
+and use it for many different kinds of tasks, even as an interactive
+input mechanism in the shell (e.g.~for editing git commit messages). The
+ability to generically reuse the text editor in many contexts makes it
+worth investing time in mastering the tool. Beaudouin-Lafon and Mackay
+refer to this ability to use a UI tool in many contexts as
+\emph{polymorphic} interaction \autocite{beaudouin-lafon2000}, noting
+that it is a useful technique for keeping interfaces simple while
+increasing their power. diSessa also \autocite{disessa2000} notes that
+there is a connection between polymorphism and the idea of literacy in a
+medium: textual literacy rests on a single rich medium of writing which
+can be adapted to many different genres and uses.
 
@@
 
-\emph{Note: Does this section overlap with ``expose unified
-structure''?}
+\hypertarget{expose-structure}{%
+\subsection{Expose structure}\label{expose-structure}}
 
-\hypertarget{in-place-toolchain}{%
-\subsection{In-place toolchain}\label{in-place-toolchain}}
+In \emph{Changing Minds} \autocite{disessa2000}, Andrea diSessa
+critiques the design of modern software with a story about a
+hypothetical ``nightmare bike.'' Each gear on the nightmare bike is
+labeled not with a number, but with an icon describing its intended use:
+smooth pavement uphill, smooth pavement downhill, gravel, etc. This
+might seem more ``user-friendly'' than numbered gears, but in fact, it
+makes it harder to operate the bike. A numerical sequence allows the
+user to develop intuition for the structure of the system, but isolated
+modes provide a superficial understanding with no grounding in
+structure. This understanding might be sufficient for the most common
+cases but breaks down in unfamiliar situations. If someone needs to go
+uphill on gravel, do they need to try every mode at random?
+
+Many modern software designs fall into this trap, teaching users to use
+isolated modes rather than coherent structure, and the problem gets far
+worse when operating across multiple applications. Unlike the UNIX
+philosophy of small tools interoperating through shared abstractions, in
+modern computing each application is in its own silo of data and
+functionality.
+
+Wildcard helps people understand and modify the behavior of applications
+through the lens of a consistent abstraction: a data table. This
+abstraction strikes a balance between being both simple and generic. A
+data table is simpler than the DOM tree that describes the details of
+the UI, but is also generic enough to describe the essence of many
+different kinds of applications.
+
+Creating a structured abstraction to represent a web page is a
+deliberate choice, and is not the only way to enable users to modify
+websites without directly accessing the DOM. Systems like Chickenfoot
+\autocite{bolin2005} and CoScripter \autocite{leshed2008} allow users to
+create scripts in an informal language and then perform fuzzy pattern
+matching to find elements in the DOM. For example, to find a button
+after a textbox in Chickenfoot, the user could type
+\texttt{click(find(“button\ just\ after\ textbox”))}. These designs
+allow for expressing a wide range of operations, but they don't
+explicitly indicate what operations are possible---the user can only see
+the original page and imagine the possibilities. In contrast, Wildcard
+provides affordances that clearly suggest the availability of certain
+actions (e.g.~sorting, editing a cell, adding a column with a derived
+value), especially to users who are familiar with spreadsheets. In
+addition to giving users more certainty about whether a modification is
+possible, these affordances might give users new ideas for things to
+try. Just as graphical interfaces better communicate the space of
+possible actions than command line interfaces, Wildcard aims to clearly
+communicate the space of possible modifications.
+
+\hypertarget{direct-manipulation-of-an-alternate-representation}{%
+\subsection{Direct manipulation of an alternate
+representation}\label{direct-manipulation-of-an-alternate-representation}}
+
+In Wildcard, users manipulate an alternate representation of a web page.
+The interaction with the data table is direct like using a spreadsheet,
+but the interaction with the page is indirectly mediated through the
+table.
+
+We considered other approaches where the user would interact more
+directly with the original UI, e.g.~injecting sort controls into the
+page, but decided that the table view had advantages that justified the
+cost of adding a layer of indirection:
+
+\begin{itemize}
+\tightlist
+\item
+  \emph{Consistency}: Even across different websites, the table view
+  always has the same layout, making it easier to learn to use.
+\item
+  \emph{Affordances}: The table view suggests possible actions like
+  adding a new column, which are challenging to suggest in the context
+  of the original page.
+\item
+  \emph{Blank slate for UI}: When a custom UI element is used to
+  manipulate a cell in the data table, there are no conflicts with the
+  existing interface of the site.
+\end{itemize}
+
+The main challenge of making this design successful is maintaining a
+close mapping in the user's mind between the new representation and the
+original page (\emph{note: cite Norman? Cognitive Dimensions `closeness
+of mapping'?}). Wildcard provides live visual cues as the user navigates
+the data table, similar to the highlighting provided by browser
+developer tools to indicate the mapping between HTML and the original
+page. In practice in our own usage, we have found that this live
+highlighting is sufficient to make it clear how the two representations
+map to each other.
+
+\hypertarget{encourage-casual-tweaking}{%
+\subsection{Encourage casual tweaking}\label{encourage-casual-tweaking}}
 
@@
 \endhead
-& \textbf{In-place} & \textbf{Not in-place}\tabularnewline
-\textbf{End user friendly} & Wildcard & IFTTT\tabularnewline
-\textbf{For developers} & browser dev tools & editing OSS\tabularnewline
+& \emph{Casual} & \emph{Not casual}\tabularnewline
+\emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
+\emph{Requires programming} & browser dev tools & editing open source
+desktop applications\tabularnewline
 \bottomrule
@@
 
-We believe that people are more motivated to modify software when the
-tools for modification are close at hand while using the original
-software, referred to by Ink and Switch as an ``in-place toolchain''
+We can evaluate a system for modifying software along two dimensions.
+First, the technical capability required of the user: is programming
+knowledge needed? Second, the level of effort required: how far out of
+their way the user must go to make a change? Can they casually make a
+small change, or do they need to make a larger project out of it? These
+dimensions are not orthogonal, but they are distinct. For example,
+setting up a workflow trigger in an end user programming system like
+\href{https://ifttt.com/}{IFTTT} does not require much technical skill,
+but it does require leaving the user's normal software and entering a
+separate environment. On the other hand, running a Javascript snippet in
+the browser console requires programming skills, but can be done
+immediately and casually in the flow of using a website.
+
+In addition to requiring no programming skills, Wildcard also aims to
+support frequent, small modifications. The Wildcard table appears in the
+course of normal web browsing, to ensure that the tools for modification
+are close at hand while using the original software. Ink and Switch
+refers to this property as having an ``in-place toolchain''
 \autocite{inkandswitch2019}.
 
-Browser developer tools exemplify an in-place toolchain, since they are
-built in to browsers and available in the context of viewing a page; in
-contrast, creating a browser extension requires moving out of the
-browser to a separate development environment.
-
-Wildcard creates an in-place toolchain for modifying web applications.
-Once installed, if an adapter is available for a site, it can be
-immediately modified from within the browser. Hopefully, this can help
-make switching from using to modifying software into a smooth
-transition, rather than a chasm to cross.
+We also try to make simple changes possible with particularly low
+effort, like being able to sort a table in a single click. This property
+is inspired by spreadsheets, which can be useful even to someone who has
+learned only a small part of their functionality. In contrast, many
+traditional programming systems require someone to learn many complex
+concepts just to perform a simple task (e.g., needing to learn what
+\texttt{public\ static\ void\ main} means to write a Hello World program
+in Java).
 
@@
 support for any website to Wildcard by building a third party adapter.
-This design acknowledges the pragmatic reality that most sites today
-will not have Wildcard support built in, and more broadly do not
-prioritize client side extensibility.
+This design decision acknowledges the pragmatic need to interoperate
+with current websites, but we hope that eventually first party website
+developers will build in Wildcard support to their applications, since
+this would reduce the burden of maintaining adapters and make Wildcard
+plugins more stable.
 
-However, we also hope that first party website developers eventually
-build in Wildcard support to their applications, which would reduce the
-burden of maintaining adapters and make Wildcard plugins more stable.
-While this might seem optimistic, implementing the Wildcard adapter API
-could be straightforward in a typical client-side application that
-already has access to a structured version of the data in the page.
-There is also precedent for first parties implementing an official
-client extension API in response to user demand: for several years,
-Google exposed an extension API in Gmail for Greasemonkey scripts to
-use. (Since then, third parties have created well-maintained
-alternatives \autocite{streak,talwar2019}, illustrating the value of
-sharing even third party adapters.)
+Implementing the Wildcard adapter API could help developers by allowing
+users to fix some of their own issues, particularly idiosyncratic use
+cases that the first party developer would never plan to prioritize.
+Supporting Wildcard could be straightforward in a typical client-side
+application that already has access to a structured version of the data
+in the page. And while some developers might hesitate to promote
+extensibility in their clients to avoid unwanted changes, the most
+common problem of users blocking ads is already ground well trod by
+existing browser extensions. There is also precedent for first parties
+implementing an official client extension API in response to user
+demand: for several years, Google maintained an official extension API
+in Gmail for Greasemonkey scripts to use. (Incidentally, since then,
+third parties have continued to maintain Gmail extension APIs used by
+many Gmail browser extensions \autocite{streak,talwar2019}, illustrating
+the value of collaboratively maintaining third party adapters.)
+
+\hypertarget{no-magic}{%
+\subsection{No magic}\label{no-magic}}
+
+\begin{itemize}
+\tightlist
+\item
+  an ecosystem of programmers collaboratively building a platform for
+  end users
+\item
+  but still empowering end users to do the creative parts, the final
+  glue
+\item
+  less brittle, less confusing and surprising (maybe Chickenfoot or
+  Coscripter had this problem? Find a nugget in those papers to support.
+  I remember Helena had something about auto-scraping breaking\ldots{}
+  contrast with Gmail.Js stability?)
+\item
+  still leaves the door open for future automation
+\item
+  maybe ``first party optional'' can be included here?
+\end{itemize}
 
@@
 \item
-  Instrumental interaction
+  Malleable software: Kay, Webstrates
 \item
-  Web automation
+  Instrumental interaction, polymorphic UI
 \item
-  Wrapper induction
+  Web automation: Chickenfoot, CoScripter
 \item
-  Personal data ownsership / SOLID
+  Wrapper induction: Thresher, Helena
+\item
+  Personal data ownership: SOLID
 \item
   Extension helper libraries, e.g.~Gmail.js.
+\item
+  Doctorow:
+  \href{https://www.eff.org/deeplinks/2019/10/adversarial-interoperability/}{adversarial
+  interoperability}
+\item
+  Marmite
 \end{itemize}
@@
 \item
-  still in early development; note the beta release plan (tentative:
-  target public beta availability at the workshop in March?)
-\item
-  What are the limits of functionality? What can't it do?
+  limitations / future ideas
 
@@
   \item
-    Should we extend further? Automation workflows? Triggers?
+    the spreadsheet language is very primitive, can we make it more
+    expressive? what are the exact semantics of fetching data from APIs?
+    What about making API requests that do mutation, not just fetching
+    data?
+  \item
+    only works when the user is browsing. Should we explore triggers,
+    scheduled scraping?
+  \item
+    only has spreadsheet-style functional transformations. Should we
+    explore imperative workflows? Injecting buttons into pages that do
+    things? (Eg, imagine a ``save to google maps'' button that you can
+    inject into a page)
+  \item
+    limited options for how to style injected content, could explore
+    styling (eg, maybe you can restyle a table cell and the styling is
+    reflected when it's injected into the page?)
+  \item
+    how much do adapters generalize to many sites? We've built X
+    adapters but need to make more
+  \item
+    page boundaries, eg scraping multi page results
+  \item
+    how do people save scripts and share them with each other? TBD
+  \item
+    nested data? lean on SIEUFERD?
+  \item
+    reduce the number of primitives?
   \end{itemize}
+\item
+  still in early development; note the beta release plan (tentative:
+  target public beta availability at the workshop in March?)
 \item
@@
 \item
-  Usability studies
+  Want to get more real usage of the tool and run usability studies
 \item
-  Link to sign up for future updates
+  Include a link to sign up for future updates
 \end{itemize}

### 2020-01-04 paper.tex
finished draft 1; edited the PDF credentials too

@@
 
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\newenvironment{Shaded}{}{}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+
 \usepackage[backend=biber]{biblatex}
@@
 \author{Geoffrey Litt}
-\authorinfo{is bla bla bla}
+\authorinfo{is a PhD student at MIT, researching programming tools for end users and developers.}
 \author{Daniel Jackson}
-\authorinfo{is bla bla bla}
+\authorinfo{is a professor in the Department of Electrical Engineering and Computer Science at MIT, associate director of CSAIL, and a MacVicar Fellow. He leads the Software Design Group.}
 \affiliation{Massachusetts Institute of Technology}
 
-\keywords{programming journal, paper formatting, submission preparation} % please provide 1--5 keywords
+\keywords{end-user programming, software customization, web browser extensions} % please provide 1--5 keywords
 
@@
 
-In this paper, we introduce Wildcard\footnote{Wildcard was the internal
-  pre-release name for Hypercard, which served as an inspiration for
-  this project since it promoted software modification by end users and
-  was a precursor to the modern Web.}, a browser extension that aims to
+In this paper, we introduce Wildcard, a browser extension that aims to
 make software more malleable by enabling users to tweak web applications
@@
 
-\begin{figure}
-\centering
-\includegraphics{media/architecture.png}
-\caption{The architecture of the Wildcard system}
-\end{figure}
-
 Wildcard is written in Typescript, and is injected into pages using the
@@
 
-In order to maximize extensibility, Wildcard is implemented as a small
-core program along with several types of plugins: site adapters,
-formulas, and cell renderers/editors. The core contains functionality
-for displaying the data table and handling user interactions, and the
-table implementation is currently built on the
-\href{https://handsontable.com/}{Handsontable} Javascript library to
-enable rapid prototyping.
+In order to maximize extensibility by other programmers, Wildcard is
+implemented as a small core program along with several types of plugins:
+site adapters, formulas, and cell renderers/editors. The core contains
+functionality for displaying the data table and handling user
+interactions, and the table implementation is built using the
+\href{https://handsontable.com/}{Handsontable} Javascript library.
 
-Site adapters\ldots.
+\begin{figure}
+\centering
+\includegraphics{media/architecture-clean.png}
+\caption{The architecture of the Wildcard system}
+\end{figure}
 
-Formula functions\ldots{}
+Site adapters specify the bidirectional connection between the web page
+and its structured data representation.
 
-Cell viewers\ldots{}
+For extracting data from the page and getting it into structured form,
+Wildcard provides ways to concisely and declaratively express scraping
+logic. For example, here is a code snippet for extracting the name of an
+Airbnb listing. It specifies the field name and some metadata
+properties, and then a function that extracts the name div from a row
+div.
 
-\begin{itemize}
-\tightlist
-\item
-  Extension points
+\begin{Shaded}
+\begin{Highlighting}[]
+  \OperatorTok{\{}
+\NormalTok{    fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,}
+\NormalTok{    readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}
+\NormalTok{    type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}
+\NormalTok{    el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
+  \OperatorTok{\}}
+\end{Highlighting}
+\end{Shaded}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    site adapters
+The site adapter also needs to support the reverse direction: sending
+updates from the table to the original page. Most DOM manipulation is
+not performed directly by the site adapter. To support sorting and
+filtering, the adapter specifies how to find the divs representing data
+rows, and the core platform sorts and filters the DOM to reflect the
+table state. User annotations are handled similarly. Finally, site
+adapters can also optionally implement site-specific actions (like
+favoriting in Airbnb) in normal Javascript code that controls the page.
 
-    \begin{itemize}
-    \tightlist
-    \item
-      show a snippet of adapter code
-    \item
-      discuss how easy it is for programmers to make adapters
-    \item
-      discuss possible future automation of adapter creation
-    \item
-      Mention the technique of scraping data from AJAX requests
-    \end{itemize}
-  \item
-    formula functions
-  \item
-    cell viewer/editors
-  \end{itemize}
-\item
-  future implementation goals
+So far we have implemented X (\emph{fill in latest number}) site
+adapters built into Wildcard. We plan to implement more in the near
+future to continue to test the limits of the API, and to encourage third
+party programmers to build, maintain, and distribute adapters for
+popular sites.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    make it easy for programmers to add adapters + plugins, and
-    distribute them to users. (Currently all adapters + plugins are part
-    of the main Wildcard codebase)
-  \end{itemize}
-\end{itemize}
+Formulas and cell editors are also implemented as separate modules, so
+that additional ones could be contributed by third party programmers.
 
@@
 The design of Wildcard is grounded in several principles, informed by
-prior work and our own experimentation. We hope you find these
-principles helpful not only for understanding our prototype, but also
-for designing other systems for end user programming.
-
-\emph{Todo: add diagrams illustrating principles}
+prior work and our own experimentation. We think these principles can
+also more generally inform the design of end user programming,
+particularly systems enabling users to modify GUI apps.
 
@@
 openness. This gives extensions freedom to modify applications in
-creative ways that the original developers did not plan for.
+creative ways that the original developers did not plan for. At the most
+extreme, this can even include modifications like ad blocking that are
+hostile to the desires of the original developers (a notion Cory
+Doctorow calls
+\href{https://www.eff.org/deeplinks/2019/10/adversarial-interoperability/}{adversarial
+interoperability}), but there are also many other changes which are not
+actively hostile, merely falling outside the range of possibilities
+originally imagined by the developer.
 
@@
 
-\hypertarget{no-magic}{%
-\subsection{No magic}\label{no-magic}}
+\hypertarget{collaboration-over-automation}{%
+\subsection{Collaboration over
+automation}\label{collaboration-over-automation}}
+
+The architecture of Wildcard requires site adapters to abstract the low
+level representation of a web page into a higher-level, structured
+representation that is easier for end users to work with. Many web
+scraping research systems have either automated this process entirely
+using heuristics, or created interactive workflows for end users to
+specify the translation. In Wildcard, we take a different approach:
+programmers manually code site adapters.
+
+This decision is partially a pragmatic one---we haven't had time to
+incorporate sophisticated automation yet, and we may choose to do so in
+the future. But it is also a principled decision. Automated approaches
+tend to work much of the time but not all of the time
+\autocite{bolin2005} \autocite{little2010}. In contrast, human
+programmers have been able to maintain very stable scraping libraries,
+even for complex sites like Gmail \autocite{streak,talwar2019}. We are
+willing to trade off lower site coverage in return for more reliable
+results on the sites that are covered, because this builds confidence
+for users about where they can expect Wildcard to work.
+
+We see this architecture as a form of collaboration between programmers
+and end users. In the wild, spreadsheets are built collaboratively by
+users with different levels of technical expertise: some users write
+simple formulas, while others help write more complex formulas or even
+macros \autocite{nardi1990}. Even though some parts of the collaboration
+require more expertise, even users with little expertise are still
+empowered to perform much of the work themselves, which differs
+radically from custom software development where coding expertise is
+table stakes for making any contributions. Similarly, we envision
+multiple layers of Wildcard usage depending on technical expertise:
 
@@
 \item
-  an ecosystem of programmers collaboratively building a platform for
-  end users
+  only using sorting/filtering/annotation, with no computation of any
+  kind
 \item
-  but still empowering end users to do the creative parts, the final
-  glue
+  writing simple formulas to fetch new data
 \item
-  less brittle, less confusing and surprising (maybe Chickenfoot or
-  Coscripter had this problem? Find a nugget in those papers to support.
-  I remember Helena had something about auto-scraping breaking\ldots{}
-  contrast with Gmail.Js stability?)
+  composing complex formulas that perform more advanced computation
 \item
-  still leaves the door open for future automation
-\item
-  maybe ``first party optional'' can be included here?
+  using Javascript to create or maintain site adapters, formulas, or
+  cell editors
 \end{itemize}
@@
 
-\emph{Note: a lot of this was already covered above; how to deal with
-that?}
+Many projects have explored end user customization of web applications,
+but it has not become a common activity in the wild. We suspect that so
+far no system has found the right combination of utility, ease of use,
+and robustness required for widespread adoption. Wildcard aims to build
+on lessons learned from this body of work, employing a novel approach of
+using a familiar spreadsheet table to go about the task of modifying the
+UI of an existing web application. Here we describe some of the areas of
+related work, and how Wildcard builds on them.
+
+\hypertarget{malleable-software}{%
+\subsection{Malleable software}\label{malleable-software}}
+
+In the broadest sense, Wildcard is inspired by systems aiming to make
+software into a dynamic medium where end users frequently create and
+modify software to meet their own needs, rather than accepting pre-built
+applications built by programmers. These systems include Smalltalk
+\autocite{kay1977}, Hypercard \autocite{hypercard2019}, Boxer
+\autocite{disessa1986}, Webstrates \autocite{klokmose2015}, and
+Dynamicland \autocite{victor}. (The project's name Wildcard comes from
+the internal pre-release name for Hypercard, which promoted software
+modification by end users and was a precursor to the modern Web.)
+
+These projects generally require building new software from scratch for
+that environment. This provides a great deal of flexibility, but also
+means they are not compatible with existing software and require
+wholesale adoption. In contrast, Wildcard takes a pragmatic approach of
+enabling people to tweak the software they already use.
+
+\hypertarget{web-augmentation}{%
+\subsection{Web augmentation}\label{web-augmentation}}
+
+Wildcard's goals are closely shared with other browser extensions which
+provide panels to augment and modify a website in the context of using
+the site.
+
+Sifter \autocite{huynh2006} enables users to sort and filter lists of
+data on web pages by converting the data into a structured format,
+providing a result similar to Wildcard's sort and filter functionality.
+The mechanism is different, though: Sifter uses a combination of
+automated heuristics and interactive user feedback to extract data,
+whereas Wildcard currently relies on programmers creating wrappers for
+extracting structured data, probably leading to higher quality
+extraction but on fewer sites. Also, Wildcard aims for much broader
+functionality than just sorting and filtering, and shows the structured
+table view of the data to facilitate these other interactions, whereas
+Sifter just shows sort controls and never reveals the underlying data
+table.
+
+``Sloppy programming'' \autocite{little2010} tools like Chickenfoot
+\autocite{bolin2005} and Coscripter \autocite{leshed2008} enable users
+to create scripts that perform actions like filling in text boxes and
+clicking buttons, without directly interacting with the DOM. Users
+express the desired page elements in natural, informal terms
+(e.g.~writing ``the username box'' to represent the textbook closest to
+the label ``username''), and then using heuristics to determine which
+elements most likely match the user's intent. This approach allows for
+expressing a wide variety of commands with minimal training, but it also
+has downsides. It is difficult to know whether a command will
+consistently work over time (in addition to changes to the website,
+changes to the heuristics can also cause problems), and it is not easy
+for users to discover the space of possible commands.
+
+Wildcard offers a sharp contrast to sloppy programming, instead choosing
+to expose a high degree of structure through the familiar spreadsheet
+table, which offers the reverse set of tradeoffs: more robust behavior,
+but in a narrower domain. Wildcard offers more consistency (e.g.,
+clicking a sort header will always work correctly as long as the site
+adapter is maintained) and offers clearer affordances for what types of
+actions are possible (or, crucially, what actions are \emph{not}
+possible, which is very useful for users to know). On the other hand,
+Wildcard cannot offer coverage of all websites, and has a narrower range
+of possible actions than sloppy tools. Still, we hope that with enough
+site adapters and formulas, these downsides can be mitigated, and that
+the benefits of structure outweigh these costs.
+
+\hypertarget{spreadsheet-based-app-builders}{%
+\subsection{Spreadsheet-based app
+builders}\label{spreadsheet-based-app-builders}}
+
+It is a powerful realization to notice that a spreadsheet can serve as
+an end-user-friendly backing data store and computation layer for an
+interactive web application. Research projects like Quilt
+\autocite{benson2014}, Gneiss \autocite{chang2014}, and Marmite
+\autocite{wong2007}, as well as commercial tools like Airtable Blocks
+\autocite{zotero-79} and Glide \autocite{zotero-81} allow users to view
+data in a spreadsheet table, compute over the data using formulas, and
+connect the table to a GUI. Because many users are already familiar with
+using spreadsheets, this way of creating applications tends to be much
+easier than traditional software methods; for example, in a user study
+of Quilt, many people were able to create applications in under 10
+minutes, even if they expected it would take them many hours.
+
+Wildcard builds on this powerful idea, but applies it to modifying
+existing applications, rather than building new applications from
+scratch. For many people, we suspect that tweaking existing applications
+provides more motivation as a starting point for programming than
+creating a new application from scratch.
+
+An important design decision for tools in this space is how far to
+deviate from traditional spreadsheets, and in what ways. Quilt and Glide
+use existing web spreadsheet software as a backend, providing maximum
+familiarity and compatibility with existing spreadsheets, but other
+projects branch out in various ways. Gneiss has its own spreadsheet with
+additional features useful for building GUIs. Marmite provides a live
+data view that resembles a spreadsheet, but programming is actually done
+using a separate data flow pane rather than spreadsheet formulas, which
+led to some confusion in a user study among users who expected behavior
+more similar to spreadsheets \autocite{wong2007}. Airtable offers a
+relational database with formula support in a spreadsheet-style direct
+manipulation interface, a combination which has found commercial
+success. Wildcard's table is most similar to Airtable, aiming to offer
+the structure of a relational table, but to keep the notion of pure
+reactive formulas from spreadsheets rather than introducing a separate
+programming model outside of the table.
+
+Quilt also raised the hypothetical idea of ``web worksheets'' in
+spreadsheets: small HTML UIs inside of a spreadsheet table to help users
+edit data or formulas in a more domain-specific interface. This is
+similar to Wildcard's notion of cell editors.
+
+\hypertarget{web-scraping-data-extraction}{%
+\subsection{Web scraping / data
+extraction}\label{web-scraping-data-extraction}}
+
+Web scraping tools focus on extracting structured data out of
+unstructured web pages. Web scraping is closely related to the
+implementation of Wildcard, but has very different goals: web scraping
+extracts static data for processing in another environment, whereas
+Wildcard supports modification of the original page by maintaining a
+bidirectional connection between the extracted data and the page.
+
+Web scraping tools differ in how much structure they attempt to map onto
+the data. Some tools like Rousillon \autocite{chasins2018} extract data
+in a minimally structured relational format; other tools like Piggy Bank
+\autocite{huynh2005} or Thresher \autocite{hogue2005} more ambitiously
+map the data to a Semantic Web schema. In Wildcard, we chose to avoid
+rich schemas, in order to reduce the work associated with creating a
+site adapter.
+
+\hypertarget{limitations-and-future-work}{%
+\section{Limitations and future
+work}\label{limitations-and-future-work}}
+
+Wildcard is still in early development; there are still many limitations
+to resolve and open questions to explore.
+
+The most important question is whether the combination of features
+provided by Wildcard can actually be combined in enough useful ways to
+make the system worth using in practice. We plan to test this by first
+continuing to privately test the system with our own needs, and
+eventually by shipping the system publicly, once the API is stable
+enough and we have developed enough site adapters and formulas to
+support a critical mass of sites and use cases.
+
+Here are a few of the most pressing limitations in the current system:
 
@@
 \item
-  Malleable software: Kay, Webstrates
+  Wildcard only extracts data on the page, which means that subsequent
+  pages of results in paginated lists are not included in the table.
+  (Sifter explains \autocite{huynh2006} techniques that might help get
+  around this.)
 \item
-  Instrumental interaction, polymorphic UI
+  There is no mechanism for end users to express imperative workflows
+  (e.g.~``click this button, and then\ldots{}''); they can only write
+  formulas that return data and then inject the resulting data into the
+  page. While this makes the system simpler, it also may exclude many
+  valuable use cases.
 \item
-  Web automation: Chickenfoot, CoScripter
+  Wildcard's data model expresses a single table at a time, without any
+  notion of relationships between tables. A UI like SIEUFERD
+  \autocite{bakke2016} might help to visualize joins between tables.
 \item
-  Wrapper induction: Thresher, Helena
-\item
-  Personal data ownership: SOLID
-\item
-  Extension helper libraries, e.g.~Gmail.js.
-\item
-  Doctorow:
-  \href{https://www.eff.org/deeplinks/2019/10/adversarial-interoperability/}{adversarial
-  interoperability}
-\item
-  Marmite
+  Only allowing site adapters to be manually coded means that the number
+  of supported sites is limited. We may explore also building wrappers
+  using automated heuristics.
 \end{itemize}
 
-\hypertarget{future-work}{%
-\section{Future work}\label{future-work}}
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
-\begin{itemize}
-\tightlist
-\item
-  limitations / future ideas
+We live in a time of pessimism about the effects of the internet and
+software. At times it seems like there is no way out of the mess, but we
+think that one promising solution is to enable more people to have
+deeper control over the software they use every day. The Web offers an
+open foundation that browser extensions have capitalized on, but even
+this has only empowered users within the limited confines of the ideas
+that programmers have been able to think of.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    the spreadsheet language is very primitive, can we make it more
-    expressive? what are the exact semantics of fetching data from APIs?
-    What about making API requests that do mutation, not just fetching
-    data?
-  \item
-    only works when the user is browsing. Should we explore triggers,
-    scheduled scraping?
-  \item
-    only has spreadsheet-style functional transformations. Should we
-    explore imperative workflows? Injecting buttons into pages that do
-    things? (Eg, imagine a ``save to google maps'' button that you can
-    inject into a page)
-  \item
-    limited options for how to style injected content, could explore
-    styling (eg, maybe you can restyle a table cell and the styling is
-    reflected when it's injected into the page?)
-  \item
-    how much do adapters generalize to many sites? We've built X
-    adapters but need to make more
-  \item
-    page boundaries, eg scraping multi page results
-  \item
-    how do people save scripts and share them with each other? TBD
-  \item
-    nested data? lean on SIEUFERD?
-  \item
-    reduce the number of primitives?
-  \end{itemize}
-\item
-  still in early development; note the beta release plan (tentative:
-  target public beta availability at the workshop in March?)
-\item
-  Could explore automated wrapper induction, building on prior work
-\item
-  Want to get more real usage of the tool and run usability studies
-\item
-  Include a link to sign up for future updates
-\end{itemize}
+In this paper, we have presented Wildcard, a general-purpose browser
+extension that gives end users more freedom to mold web applications to
+meet their needs. To receive future updates on Wildcard and
+notifications about a public release, sign up for the email newsletter.
 

### 2020-01-07 paper.tex
starting to add dnj edits

--- 2019-12-29 paper.tex
@@
 \begin{abstract}
-Browser extensions and user scripts can modify web applications in
-useful ways, but many people have unique needs that aren't met by
-existing extensions. Today, most of those people can't build their own
-browser extensions, since they can't program in Javascript.
+Many Web applications do not meet the precise needs of their users.
+Browser extensions and user scripts offer a way to customize web
+applications, but most people do not have the programming skills to
+implement their own customizations.
 
-Wildcard is a platform that empowers anyone to build browser extensions
-without doing traditional programming. Wildcard shows the main data from
-a web page in a table that maintains a bidirectional connection to the
-original page. By directly manipulating the table, people can
-sort/filter content, add private annotations, and use spreadsheet
-formulas to fetch data from other websites.
+In this paper, we present a prototype of Wildcard, a live programming
+system embedded in the browser that empowers users to casually tweak web
+applications without needing to program in Javascript or interact with
+the DOM.
+
+Wildcard shows the data from a web page in a table, and maintains a
+bidirectional connection between the table and the original page. By
+directly manipulating the table, users can perform a wide variety of
+modifications: sorting/filtering content, adding private annotations,
+using spreadsheet formulas to fetch data from other web services, using
+custom UI elements to edit form data, and more.
+
+We present examples of using Wildcard to solve real world problems, and
+explain the design principles behind the prototype. In the future, we
+envision continuing to build Wildcard into a fully deployed system that
+helps make the web into a more malleable medium for all users.
 \end{abstract}
@@
 In 2012, the travel site Airbnb removed the ability to sort listings by
-price. Users could still filter down to a price range, but could no
-longer view the cheapest listings first. Many users complained on online
-message boards that the change seemed hostile to users. ``It's so
-frustrating!..What is the logic behind not having this function?'' says
+price. Users could still filter by price range, but could no longer view
+the cheapest listings first. Many users complained on online message
+boards that the change seemed hostile to users. ``It's so
+frustrating!..What is the logic behind not having this function?'' said
 one user on the
@@
 This is a familiar situation in a world of web applications that are
-frequently updated without user consent. When web software doesn't quite
-meet users' needs, their only recourse is to complain to the developers
-and hope someone listens. Sometimes there is a way to use a browser
-extension or user script to fix the issue, but making an extension
-requires sophisticated programming skills. We see this as a waste of the
-openness of the Web platform, and of the pliability of software in
-general. If users could make tweaks to web applications without
-programming, a vast number of people would gain the agency to modify
-software to meet their unique preferences.
+frequently updated without user consent. Sometimes there is a browser
+extension or user script that fixes an issue, and if the user is a
+motivated, programmer they might even be able to implement their own
+fix. But for most people in most situations, the only recourse is to
+complain to the developers and pray that someone listens---or more
+likely, to simply give up. While many have become used to this status
+quo, we see it as a tremendous waste of the openness of the Web
+platform. In \emph{Personal Dynamic Media} \autocite{kay1977}, Alan Kay
+originally envisioned personal computing as a medium that let a user
+``mold and channel its power to his own needs,'' but today's software is
+more like concrete than clay.
 
-In this paper, we introduce a system called Wildcard\footnote{Wildcard
-  was the original internal name for Hypercard, which pioneered both
-  user modifiable software and the modern Web.} that aims to meet this
-need. Wildcard adds a panel to the bottom of a web page that shows a
-structured table view of some of the main data in the page. The table
-maintains a bidirectional connection to the original page---when the
-user manipulates the table, the original page also gets modified, and
-vice versa.
+In this paper, we introduce Wildcard, a live programming system embedded
+in the browser that enables users to tweak web applications without
+programming in Javascript or interacting with the DOM. Wildcard adds a
+panel to the bottom of a web page that shows a structured table view of
+the main data in the page. The table maintains a bidirectional
+connection to the original page---when the user manipulates the table,
+the original page gets modified, and vice versa.
 
-In Wildcard, sorting Airbnb listings by price takes just one click on a
-table header, with no programming required by the end user. Beyond
-sorting and filtering data, Wildcard also supports a variety of other
-use cases like accessing third party APIs, performing small
-computations, recording private user annotations, and using alternate UI
-widgets. While Wildcard does not support all changes someone might want
-to make to a website, the goal of the project is to make a broad subset
-of changes easily accessible to end users.
+In Wildcard, a user can sort Airbnb listings with just one intuitive
+click on a table header, with no programming required. Beyond sorting
+and filtering data, Wildcard also supports accessing third party APIs,
+performing small computations, recording private user annotations, and
+using alternate UI widgets to control a page. While Wildcard does not
+support any change someone might want to make to a website, it makes a
+variety of common and useful changes more accessible to end users, as
+well as to programmers who want to avoid tedious low-level programming.
 
-Under the hood, there is no magic---Wildcard currently requires a
-programmer to manually define an adapter between the original page and
-the data table using web scraping techniques. While programming is
-required for part of the process, this forms a useful collaboration
-between the programmer and the end user, since the end user is able to
-perform many different modifications once an adapter is defined. We also
-plan to eventually reduce the burden of making adapters (discussed in
-more detail later).
+Under the hood, the implementation of Wildcard is straightforward,
+because a programmer must manually write an adapter for each website
+which uses common web scraping techniques to map the data from the web
+page into the Wildcard table format. Although there is still manual
+programming involved, the result is very different from traditional
+single-use browser extensions, because instead of the programmer
+defining a narrow use case, the end user is able to make many different
+changes on top of a single site-specific adapter.
 
-\hypertarget{demo-booking-a-trip-with-wildcard}{%
-\section{Demo: Booking a trip with
-Wildcard}\label{demo-booking-a-trip-with-wildcard}}
-
-To get a sense of the user experience of Wildcard, let's see an example
-of how an end user can modify a website with Wildcard in practice.
-
-For example, in Fig.~\ref{fig:table} we open up a table view that
-corresponds to search results on the Airbnb travel site.
-
-\begin{figure}
-\hypertarget{fig:table}{%
-\centering
-\includegraphics{media/opentable.png}
-\caption{Opening a table corresponding to search results on
-Airbnb}\label{fig:table}
-}
-\end{figure}
+In this paper, we present examples of using Wildcard to solve real world
+problems, and then explain the system architecture and design principles
+behind the prototype:
 
@@
 \item
-  Sorting search results:
-
-  \begin{itemize}
-  \tightlist
+  \emph{Expose consistent structure}: turning isolated applications into
+  a structured medium
 \item
-    Airbnb took away search
+  \emph{Encourage casual tweaking}: using frequent software modification
+  as a starting point for end user programming
 \item
-    once you see the tabular view, there's an obvious interaction
-    available to anyone familiar with tables
-  \end{itemize}
-\item
-  Injecting new data into the page
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Take own custom notes, saved in the browser. Maybe shared in the
-    future
-  \end{itemize}
-\item
-  Formulas
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Compute values to inject into the page
-  \item
-    TBD: styling
-  \end{itemize}
-\item
-  Using a custom date picker
+  \emph{Build for an ecosystem}: designing for first party developers,
+  third party programmers, and non-programmer end users
 \end{itemize}
 
-\hypertarget{system-implementation}{%
-\section{System Implementation}\label{system-implementation}}
+We also explore related work on web augmentation tools,
+spreadsheet-based app builders, and web scraping tools. Our novel
+contribution in this paper is to combine ideas from each of these areas
+into a new kind of tool: a programming environment which maps data from
+web pages onto a live table view, enabling end users to modify the data.
 
-\emph{Goal of this section: briefly explain the current implementation.
-Enough detail to ground further discussion, but don't dwell on it. It's
-not the main point, and will evolve a lot.}
+Wildcard is currently an early research prototype, with incomplete
+features and limited coverage of sites. We plan to continue privately
+testing the system with our own use cases, and then to release the tool
+publicly, enabling other programmers to contribute site adapters and
+other plugins. Ultimately, we hope that Wildcard makes modifying
+websites into a natural part of using the web, as mundane and ubiquitous
+as using spreadsheets.
 
-\begin{itemize}
-\tightlist
-\item
-  Built as a Greasemonkey script for now (\emph{todo: convert to a full
-  broser extension?})
-\item
-  Describe the adapter API
+\hypertarget{demos}{%
+\section{Demos}\label{demos}}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    programmer creates adapters for now
-  \item
-    maybe end users can create in the future, or automated
-  \item
-    show a snippet of adapter code
-  \end{itemize}
-\item
-  Mention the technique of scraping data from AJAX requests
-\item
-  future implementation goals
+To get a sense of how it feels to use Wildcard, let's see an example of
+using it to help with booking a trip using the travel sites Airbnb and
+Expedia.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    make it easy for programmers to add adapters + plugins, and
-    distribute them to users. (Currently all adapters + plugins are part
-    of the main Wildcard codebase)
-  \end{itemize}
-\end{itemize}
+\hypertarget{sorting-and-filtering}{%
+\subsection{Sorting and filtering}\label{sorting-and-filtering}}
+
+We start by opening up the Airbnb search listings page to look for a
+place to stay. As mentioned before, this page doesn't let us sort by
+price, but we can use Wildcard to fix that. First, we open up the
+Wildcard panel, which shows a table corresponding to the search results
+in the page. As we click around in the table, the corresponding row in
+the page is highlighted so we can see the connection between the views.
+
+When we click the Price column header in the table, both the table and
+the original page become sorted by price. We can also filter listings by
+rating, another feature not offered in the Airbnb site.
+
+After finishing the sorting and filtering, we can close the table view
+and continue using the website in its original polished design.
+
+\hypertarget{row-actions}{%
+\subsection{Row actions}\label{row-actions}}
+
+Most websites that show lists of data also offer actions that can be
+taken on a row in the table, like adding an item to a shopping cart.
+Wildcard has the ability to make these actions available in the data
+table if the site adapter implements them. The main advantage this
+provides is the ability to easily perform an action in bulk across
+multiple rows.
+
+For example, it's tedious on Airbnb to click on listings one by one to
+add them to a list of favorites. Using Wildcard, we can just select
+multiple rows and favorite all of them with one click. Similarly, we can
+also open the detailed pages for many listings in new tabs.
+
+Within the site adapter, each action is implemented as a regular
+Javascript function running in the context of the page, which can click
+on buttons in the UI, launch AJAX requests, navigate to a new page, etc.
+
+\hypertarget{user-annotations}{%
+\subsection{User annotations}\label{user-annotations}}
+
+It's a common practice in spreadsheets to add a column next to a table
+for recording notes about the item in each row. In Wildcard, this
+pattern can be used to create private annotations in a website. Here, we
+use this feature to jot down pros and cons of various listings:
+
+In order to match the annotations with the existing design, the site
+adapter specifies a reasonable location in each row for injecting new
+data.
+
+Annotations are persisted in browser local storage so the user can come
+back and view them later. The site adapter associates each table row
+with a stable identifier from the original site so that annotations can
+be displayed with the appropriate item.
+
+Currently annotations are private to the browser of the user that
+created them, but it could be useful in the future to allow users to
+share annotations.
+
+\hypertarget{computation-with-formulas}{%
+\subsection{Computation with formulas}\label{computation-with-formulas}}
+
+The demos so far have shown straightforward tweaks that provide useful
+conveniences with little effort. But Wildcard also supports more
+sophisticated tweaks that fetch external data and perform computations,
+using a formula language.
+
+When traveling without a car, it's nice to evaluate potential places to
+stay based on how walkable the surroundings. Using Wildcard formulas, we
+can integrate Airbnb with Walkscore, an API that can rate the
+walkability of any location on a 1-100 scale. When we call the
+\texttt{WALKSCORE} formula with the latitude and longitude of the
+listing, it returns the score as the cell value. Because the cell is
+shown in the page, the score also shows up in the page body.
+
+Wildcard formulas are just Javascript functions that take table data as
+input and return new data. Formulas can have limited side effects like
+accessing external APIs to fetch data, but do not directly manipulate
+the page. Instead, they return data into the table, which is then
+injected into the page with the same mechanism used for injecting user
+annotations.
+
+\hypertarget{custom-ui-elements}{%
+\subsection{Custom UI elements}\label{custom-ui-elements}}
+
+It might seem that Wildcard is only useful on websites that display
+lists of tabular data like search results. But in fact, the table
+metaphor is flexible enough to represent many types of data. For
+example, a form can be represented as a single row, with a column for
+each input.
+
+In previous examples the data extracted from the site was marked as
+read-only (it's not allowed to change the name of an Airbnb listing),
+but in this case, the cells are marked as writable, so that changes in
+the table are reflected in the original form. This becomes particularly
+useful when combined with custom cell editing UIs, which allow users to
+graphically edit the value of a cell.
+
+Here's an example of using those features to help with filling in a
+form. Filling in dates for a flight search typically requires opening up
+a separate calendar app to look up the correct dates, and then manually
+copying them into the form. In Wildcard, we can make this easier by
+editing the date cell using a datepicker widget that has privileged
+access to our calendar information.
+
+Custom UIs enable people to use a consistent UI to enter common types of
+data across the web. They also allow a user to access their own private
+data as part of a web interface, without needing to expose it to the
+website server.
+
+Overall, an interactive data table is a computation model that presents
+a surprisingly large range of useful possibilities for end user
+modification of websites, while also remaining familiar and easy to use.
+While we've presented concretely illustrated here just a few of these
+possibilities, we plan to continue developing site adapters, formulas,
+and UI elements to explore more use cases, and to eventually publicly
+release the tool to allow real end users to discover their own uses.
+
+\hypertarget{system-architecture}{%
+\section{System Architecture}\label{system-architecture}}
+
+Wildcard is written in Typescript. It is currently injected into pages
+using the \href{https://www.tampermonkey.net/}{Tampermonkey} userscript
+manager, but in the future we plan to deploy it as a standalone browser
+extension to make it easier to install.
+
+In order to maximize extensibility, Wildcard is implemented as a small
+core program along with several types of plugins: site adapters,
+formulas, and cell renderers/editors. The core contains functionality
+for displaying the data table and handling user interactions, and the
+table implementation is built using the
+\href{https://handsontable.com/}{Handsontable} Javascript library.
+
+\begin{figure}
+\centering
+\includegraphics{media/architecture-clean.png}
+\caption{The architecture of the Wildcard system}
+\end{figure}
+
+Site adapters specify the bidirectional connection between the web page
+and its structured data representation.
+
+For extracting data from the page and getting it into structured form,
+Wildcard provides ways to concisely and declaratively express scraping
+logic. For example, here is a code snippet for extracting the name of an
+Airbnb listing. It specifies the field name and some metadata
+properties, and then a function that extracts the name div from a row
+div.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+  \OperatorTok{\{}
+\NormalTok{    fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,}
+\NormalTok{    readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}
+\NormalTok{    type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}
+\NormalTok{    el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
+  \OperatorTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+The site adapter also needs to support the reverse direction: sending
+updates from the table to the original page. Most DOM manipulation is
+not performed directly by the site adapter: the adapter specifies how to
+find the divs representing data rows, and the core platform mutates the
+DOM to reflect the table state. The only exception is site-specific
+actions (like favoriting in Airbnb) which runs Javascript code that can
+mutate the DOM or perform other arbitrary behaviors.
+
+So far we have only implemented three site adapters in Wildcard, but we
+plan to implement many more in the near future to continue to test the
+limits of the API. Eventually we will also encourage third party
+programmers to build and maintain adapters for popular sites.
 
@@
 
-Below are some of the ideas behind the design of Wildcard. We hope that
-these principles are also useful for thinking more broadly about
-enabling end users to modify software.
+The design of Wildcard is grounded in several principles, informed by
+prior work and our own experimentation. We think these principles extend
+beyond this specific system and can more generally inform the design of
+end user programming tools, particularly systems enabling users to
+modify GUI apps.
 
-\hypertarget{expose-unified-structure-across-applications}{%
-\subsection{Expose unified structure across
-applications}\label{expose-unified-structure-across-applications}}
-
-\emph{Todo: add a diagram here?}
+\hypertarget{expose-consistent-structure}{%
+\subsection{Expose consistent
+structure}\label{expose-consistent-structure}}
 
@@
 labeled not with a number, but with an icon describing its intended use:
-smooth pavement uphill, smooth pavement downhill, gravel, etc. By some
-logic, this is more ``user-friendly'' than numbered gears, but in fact,
-hiding orderly structure from the user makes it more difficult to
-operate the bike. Many modern software designs fall into this trap,
-teaching users to use isolated modes rather than coherent structure, and
-the problem gets far worse when operating across multiple applications.
-Unlike the UNIX philosophy of small tools interoperating through shared
-abstractions, in modern computing each application is in its own silo of
-data and functionality.
+smooth pavement uphill, smooth pavement downhill, gravel, etc. This
+might seem more ``user-friendly'' than numbered gears, but in fact, it
+makes it harder to operate the bike. An ordered numerical sequence
+allows the user to intuitively grasp the structure of the system---``if
+I shift up, it gets harder to pedal''---but isolated modes provide a
+superficial understanding with no grounding in structure. This
+understanding might be sufficient for the most common cases, but breaks
+down in unfamiliar situations. If someone needs to go uphill on gravel,
+do they need to try every mode at random?
 
-Wildcard helps people understand and modify the behavior of applications
-through the lens of a consistent abstraction: a data table. This
-abstraction strikes a balance between being both simple and generic. A
-data table is simpler than the DOM tree that describes the details of
-the UI, but is also generic enough to describe the essence of many
-different kinds of applications.
+Today, most personal computing consists of using distinct apps, each one
+an isolated silo of concepts, interfaces, data, and functionality. While
+there are some shared idioms and limited points of interoperability, the
+fundamental expectation is that each app operates independently. If a
+new need emerges, a user must hope ``there's an app for that''; creating
+a new tool to meet the need is often out of the question.
 
-Exposing a unified higher-level abstraction is a deliberate choice, and
-is not the only way to enable users to program without directly using
-the DOM. Chickenfoot \autocite{bolin2005} and CoScripter
-\autocite{leshed2008} allow users to create scripts that resemble
-natural language and ``sloppily'' match queries to elements in the DOM.
-These designs allow for a wide range of operations, but don't explicitly
-indicate what operations are possible---the user must look at the
-original page and imagine the possibilities. In contrast, Wildcard
-provides affordances that clearly suggest the availability of certain
-types of modifications. In addition to giving users more certainty about
-whether a modification is possible, these affordances might give users
-new ideas for things to try. Because people are not used to modifying
-web applications, providing inspiration is an important goal. (Todo:
-Perhaps something to cite here, re: discoverability in GUIs vs CLIs?)
+Computing does not need to operate this way. As one example, UNIX offers
+a compelling alternate vision of small tools connected through a common
+abstraction of text files. Just like riding a bike with numbered gears,
+UNIX requires users to build up a structured understanding of the
+system, which then allows them to flexibly react to new needs not served
+by existing tools. Users can also deeply master a text editor and use it
+for a vast array of tasks, even as an interactive input mechanism in
+shell programs (e.g.~for editing git commit messages). Beaudouin-Lafon
+and Mackay call this style of interaction \emph{polymorphic}
+\autocite{beaudouin-lafon2000}, since a single interface can be used in
+many contexts. diSessa \autocite{disessa2000} notes the deep connection
+between polymorphism and literacy: textual literacy rests on a single
+rich medium of writing, which once mastered, can be adapted to many
+different genres and uses. If people needed to relearn from scratch when
+switching from writing essays to writing emails, the medium would lose
+most of its potency; this is the situation in software today.
 
-\hypertarget{direct-manipulation-by-proxy}{%
-\subsection{Direct manipulation by
-proxy}\label{direct-manipulation-by-proxy}}
-
-In Wildcard, users manipulate a data table added to the page, rather
-than the original page. Although the interaction with the table is
-direct like using a spreadsheet, the interaction with the page is
-indirectly mediated through the table.
-
-We considered other approaches where the user would interact more
-directly with the original UI, e.g.~injecting sort controls into the
-page, but decided that the consistency and affordances of the table view
-outweighed the potential confusion of a new layer of indirection.
-
-Making this design successful requires maintaining a close mapping in
-the user's mind between the new representation and the original page
-(cite Norman? Cognitive Dimensions?). Wildcard provides live visual cues
-as the user navigates the data table, similar to the highlighting
-provided by browser developer tools to indicate the mapping between HTML
-and the original page.
+With Wildcard, we hope to make isolated Web applications work a little
+bit more like UNIX, by imposing a consistent and simple structure onto
+many isolated applications. Just as UNIX uses text files to represent
+all data, Wildcard uses relational tables. This abstraction strikes a
+balance between being simple and generic: a data table is much simpler
+than the DOM tree that describes all the details of the UI, but is also
+generic enough to describe the essence of many different interfaces.
 
 \begin{figure}
-\hypertarget{fig:devtools}{%
 \centering
-\includegraphics{media/devtools.png}
-\caption{The Chrome Dev Tools use highlighting to show the mapping
-between HTML code and the page}\label{fig:devtools}
-}
+\includegraphics{media/structure.png}
+\caption{Wildcard layers the consistent structure of a data table over
+different isolated applications.}
 \end{figure}
 
-\hypertarget{decouple-ui-from-data}{%
-\subsection{Decouple UI from data}\label{decouple-ui-from-data}}
+Exposing more structured data to users provides improved quality and
+consistency of user interfaces. When UI widgets can compete on their
+merits rather than having a monopoly on a given data silo or
+application, it creates stronger competition at the interface level. For
+example, there is competition among email clients which consume an open
+protocol, but not among Facebook or Twitter clients. Exposing structure
+also leads to improved UI consistency across apps: UNIX users can reuse
+their text editor everywhere, but Web users must use a different rich
+text editor for each site. Tim Berners-Lee has written in more detail
+about these benefits \autocite{berners-lee2018} in the context of the
+SOLID project, which envisions user-controlled data as a mechanism for
+decoupling data from interfaces, e.g.~giving users a choice of which
+client to use to consume a given social media feed. Wildcard has
+overlapping goals with SOLID, but does not require decentralized user
+control of data---the data can remain on a centralized server, as long
+as the interface can be tweaked by end users.
 
-Most software does not provide much user choice for interface elements,
-even for common data types: e.g., the website provides a datepicker
-widget, and you have no ability to provide your own replacement
-datepicker. This forces users to learn many different interfaces for
-similar tasks, some of them worse than others. It also prevents
-combining data across multiple applications within a single user
-interface.
+Layering a structured view on top of an existing application is not the
+only option available. Systems like Chickenfoot \autocite{bolin2005} and
+CoScripter \autocite{leshed2008} forego the intermediate layer of
+structure, instead allowing users to create scripts in an informal
+language and then perform fuzzy pattern matching to find elements in the
+DOM. For example, to find a button after a textbox in Chickenfoot, the
+user could type \texttt{click(find(“button\ just\ after\ textbox”))}.
+These designs allow for expressing a wide range of operations, but they
+don't explicitly indicate what operations are possible---the user can
+only see the original page and imagine the possibilities. In contrast,
+Wildcard provides affordances that clearly suggest the availability of
+certain actions (e.g.~sorting, editing a cell, adding a column with a
+derived value), especially to users who are familiar with spreadsheets.
+In addition to giving users more certainty about whether a modification
+is possible, these affordances might improve discoverability and give
+users new ideas for things to try. We also think that sloppy programming
+does little to solve the problem of thinking of applications as isolated
+worlds, whereas Wildcard contributes to giving users a structured
+understanding of the different apps that they use.
 
-Wildcard exposes the data underlying an application's UI, allowing the
-user to use interfaces of their choice to view and modify the data. The
-Expedia datepicker demo above showed one example of how this can be
-useful, but we also envision creating other widgets for visualizing and
-editing data. Some examples would be showing geographic data in a custom
-map that includes the user's own annotations, or editing a blog post in
-a rich text editor of the user's choice.
+One key challenge of making a structured view successful is ensuring a
+close mapping in the user's mind between the new representation and the
+original page. Wildcard provides live visual cues as the user navigates
+the data table (similar to the highlighting provided by DOM inspectors
+in browser developer tools). In our own usage, we have found that this
+live highlighting makes it very clear how the two representations map to
+each other.
 
-One benefit of decoupling data from interfaces is improved UI quality.
-When UI widgets can compete on their merits rather than based on network
-effects from the data they have access to, it creates much stronger
-competition at the interface level. For example, there is competition
-among email clients (which consume an open protocol), but not among
-Facebook or Twitter clients. This benefit relates to the SOLID project
-led by Tim Berners-Lee \autocite{berners-lee2018}, which envisions
-user-controlled data as a mechanism for decoupling data from interfaces,
-e.g.~giving users a choice of which client to use to consume a given
-social media feed. Wildcard has overlapping goals with SOLID, but does
-not require decentralized user control of data---the data can remain on
-a centralized server, as long as the interface can be tweaked by end
-users. \emph{Note: maybe this paragraph belongs in related work?}
+\hypertarget{encourage-casual-tweaking}{%
+\subsection{Encourage casual tweaking}\label{encourage-casual-tweaking}}
 
-There is also value in using the same consistent interface across many
-applications. For example, many programmers become deeply familiar with
-one text editor and use it for many different kinds of tasks. This usage
-even extends beyond editing permanent files; shell programs can use the
-user's preferred text editor as an interactive input mechanism (e.g.~for
-editing git commit messages). This ability to generically reuse the text
-editor in many contexts makes it more worthwhile to invest time
-mastering the tool. Beaudouin-Lafon and Mackay refer to this type of
-reuse as \emph{polymorphism} in interaction
-\autocite{beaudouin-lafon2000}, noting that it is a key technique for
-keeping interfaces simple while increasing their power.
-
-\emph{Note: Maybe could relate this section to Concept Reuse?}
-
-\emph{Note: Does this section overlap with ``expose unified
-structure''?}
-
-\hypertarget{in-place-toolchain}{%
-\subsection{In-place toolchain}\label{in-place-toolchain}}
+In order to serve the widest possible audience, Wildcard is designed for
+tweaking existing software for one's own needs, not for designing a new
+application from scratch. Creating a new application requires motivation
+and other skills besides programming that not everyone has, but having
+frustrations with some aspect of an app which could be solved by a small
+tweak is a nearly universal experience.
 
@@
 \endhead
-& \textbf{In-place} & \textbf{Not in-place}\tabularnewline
-\textbf{End user friendly} & Wildcard & IFTTT\tabularnewline
-\textbf{For developers} & browser dev tools & editing OSS\tabularnewline
+& \emph{Casual} & \emph{Not casual}\tabularnewline
+\emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
+\emph{Requires programming} & browser dev tools & editing open source
+software\tabularnewline
 \bottomrule
@@
 
-We believe that people are more motivated to modify software when the
-tools for modification are close at hand while using the original
-software, referred to by Ink and Switch as an ``in-place toolchain''
-\autocite{inkandswitch2019}.
+We can evaluate a system for tweaking software along two dimensions.
+First, the technical capability required of the user: is programming
+knowledge needed? Second, the level of effort required: how far out of
+their way the user must go to make a change? Can they casually make a
+small change, or do they need to make a larger project out of it? These
+dimensions are not orthogonal, but they are distinct. For example,
+setting up a workflow trigger in an end user programming system like
+\href{https://ifttt.com/}{IFTTT} does not require much technical skill,
+but it does require leaving the user's normal software and entering a
+separate environment. On the other hand, running a Javascript snippet in
+the browser console requires programming skills, but can be done
+immediately and casually in the flow of using a website.
 
-Browser developer tools exemplify an in-place toolchain, since they are
-built in to browsers and available in the context of viewing a page; in
-contrast, creating a browser extension requires moving out of the
-browser to a separate development environment.
+In addition to requiring no programming skills, Wildcard also aims to
+support frequent, small modifications. The Wildcard table appears in the
+course of normal web browsing, to ensure that the tools for modification
+are close at hand while using the original software. Ink and Switch
+refers to this property as having an ``in-place toolchain''
+\autocite{inkandswitch2019}. Wildcard also provides live feedback,
+immediately editing the page in response to user interaction, which
+gives users confidence in the changes they are making.
 
-Wildcard creates an in-place toolchain for modifying web applications.
-Once installed, if an adapter is available for a site, it can be
-immediately modified from within the browser. Hopefully, this can help
-make switching from using to modifying software into a smooth
-transition, rather than a chasm to cross.
+Wildcard also aims to make some valuable changes possible with
+particularly low effort: for example, sorting a table in a single click,
+or adding an annotation by filling in a cell. We would expect many
+Wildcard users to start by using these bits of functionality, before
+possibly moving on to more complex features like formulas. This property
+is inspired by spreadsheets, which can be used productively even by
+someone who has learned only a tiny part of their functionality,
+e.g.~for storing tables of numbers or computing simple sums
+\autocite{nardi1991}. In contrast, many traditional programming systems
+require someone to spend days learning complex concepts before being
+able to create a useful piece of software---printing ``hello world'' is
+not a genuinely valuable activity.
 
-\hypertarget{first-party-cooperation-optional}{%
-\subsection{First party cooperation
-optional}\label{first-party-cooperation-optional}}
+\hypertarget{build-for-an-ecosystem}{%
+\subsection{Build for an ecosystem}\label{build-for-an-ecosystem}}
 
-The Web is an unusually extensible platform. On many other platforms
-(e.g.~smartphone operating systems), software is locked down unless
-first-party developers explicitly provide hooks for plugins and
-interoperability, but on the Web, all client-side code is available for
-browser extensions to modify. Application authors can use practices that
-make it easier to modify their apps (e.g.~clean semantic markup), or
-more difficult (e.g.~code obfuscation), but the default state is
-openness. This gives extensions freedom to modify applications in
-creative ways that the original developers did not plan for.
+Some end user programming systems are designed with a single user in
+mind---a lone non-programmer, needing to achieve a task through some
+combination of individual skills and a powerful tool. But in fact,
+real-world usage of end user programming systems is far more
+collaborative. Spreadsheets are built together by users with different
+levels of technical expertise: some users just write simple formulas,
+while their officemates help write more complex formulas or even program
+macros in traditional languages \autocite{nardi1990}. Our goal with
+Wildcard is to consider the full ecosystem of users and to design a tool
+that best combines all of their abilities.
+
+This design principle is most salient in our approach to building site
+adapters. Many web augmentation projects aim to make the full process of
+tweaking available to end users with no programmer help needed, through
+a combination of automated heuristics and interactive feedback. This is
+a laudable goal because it makes these tools work with any site, and
+these approaches are often quite successful, but they also inevitably
+break down sometimes \autocite{bolin2005} \autocite{little2010}. In
+contrast, human programmers working together have been able to maintain
+stable scraping libraries, even for complex sites like Gmail
+\autocite{streak,talwar2019}. Our primary approach to building site
+adapters is having programmers manually create them, because we are
+willing to trade off lower site coverage in return for more reliable
+results on the sites that are covered, building confidence for users.
+Like the collaboration between an expert and novice spreadsheet user,
+this architecture leverages more skilled programmers while still
+providing end users with a flexible programming environment to meet
+their own needs. On the other hand, we remain very open to incorporating
+other tools in the future that enable end users to also create their own
+adapters.
+
+\begin{figure}
+\centering
+\includegraphics{media/adapters.png}
+\caption{Having third party programmers create site adapters balances
+site coverage and adapter quality.}
+\end{figure}
+
+In any system for tweaking software, another important consideration is
+how to involve the first party developers of the original software. Our
+approach is to avoid depending on their cooperation, but also to aim for
+eventual cooperation.
+
+This approach takes advantage of the unusual openness of the Web
+platform. On many other platforms (e.g.~smartphone operating systems),
+software is locked down unless first-party developers explicitly provide
+hooks for plugins and interoperability, but on the Web, all client-side
+code is available for browser extensions to modify. Application authors
+can use practices that make it easier to modify their apps (e.g.~clean
+semantic markup), or more difficult (e.g.~code obfuscation), but the
+default state is openness. This gives extensions freedom to modify
+applications in creative ways that the original developers did not plan
+for. At the most extreme, this can even include modifications that are
+hostile to the desires of the original developers, a notion Cory
+Doctorow calls
+\href{https://www.eff.org/deeplinks/2019/10/adversarial-interoperability/}{adversarial
+interoperability}. Although ad blocking extensions make this style of
+extension very prominent, most ideas for extending software merely fall
+outside the range of possibilities originally imagined by the developer,
+rather than being actively hostile.
 
@@
 support for any website to Wildcard by building a third party adapter.
-This design acknowledges the pragmatic reality that most sites today
-will not have Wildcard support built in, and more broadly do not
-prioritize client side extensibility.
-
-However, we also hope that first party website developers eventually
-build in Wildcard support to their applications, which would reduce the
-burden of maintaining adapters and make Wildcard plugins more stable.
-While this might seem optimistic, implementing the Wildcard adapter API
-could be straightforward in a typical client-side application that
-already has access to a structured version of the data in the page.
-There is also precedent for first parties implementing an official
-client extension API in response to user demand: for several years,
-Google exposed an extension API in Gmail for Greasemonkey scripts to
-use. (Since then, third parties have created well-maintained
-alternatives \autocite{streak,talwar2019}, illustrating the value of
-sharing even third party adapters.)
+However, we also hope that eventually first party website developers
+will build in Wildcard support to their applications, since this would
+reduce the burden of maintaining adapters and make Wildcard plugins more
+stable. Implementing the Wildcard adapter API could help developers by
+allowing users to fix some of their own issues, and would not
+necessarily require too much effort---we envision making it
+straightforward to add Wildcard support in a client-side application
+which already has access to the structured data in the page. There is
+also precedent for first parties implementing an official client
+extension API in response to user demand: for several years, Google
+maintained an official extension API in Gmail for Greasemonkey scripts
+to use. (Incidentally, since then, third parties have continued to
+maintain stable Gmail extension APIs used by many browser extensions
+\autocite{streak,talwar2019}, illustrating the potential of
+collaboratively maintaining third party adapters.)
 
@@
 
-\emph{Note: a lot of this was already covered above; how to deal with
-that?}
+In the broadest sense, Wildcard is inspired by systems aiming to make
+software into a dynamic medium where end users frequently create and
+modify software to meet their own needs, rather than only consuming
+applications built by programmers. These systems include Smalltalk
+\autocite{kay1977}, Hypercard \autocite{hypercard2019}, Boxer
+\autocite{disessa1986}, Webstrates \autocite{klokmose2015}, and
+Dynamicland \autocite{victor}. (The project's name Wildcard comes from
+the internal pre-release name for Hypercard, which doubly inspired our
+work by promoting both software modification by end users and the ideas
+behind the Web.)
+
+While similar in high-level goals, Wildcard employs a different solution
+strategy from these projects: whereas they generally require building
+new software from scratch for that environment, Wildcard instead aims to
+maximize the malleability of already existing software. This approach
+has the pragmatic benefit of immediately being useful in a much broader
+context, although it also requires working within rigid constraints. We
+also see Wildcard growing beyond its current bounds in the future and
+becoming more similar to these other systems. With substantial future
+work, Wildcard could grow from merely being a platform for tweaking
+existing software into a platform for building new software from
+scratch, designed from the ground up for end user tweakability.
+
+More narrowly, Wildcard builds on three areas of existing work: web
+augmentation, spreadsheet-based app development, and web scraping. Our
+contribution in this work is to combine these areas in a new way:
+extracting structured data from a website and directly exposing it to
+the user as a live data table, with the end goal of modifying an
+existing application. We think this combination builds on valuable ideas
+and learnings from each of the areas, while also overcoming some of the
+downsides of each.
+
+\hypertarget{web-augmentation}{%
+\subsection{Web augmentation}\label{web-augmentation}}
+
+Wildcard's goals are closely shared with other systems which provide
+interfaces in the browser for end users to augment and modify websites
+while using them.
+
+\hypertarget{structured-augmentation}{%
+\subsubsection{Structured augmentation}\label{structured-augmentation}}
+
+Wildcard's approach is most similar to a class of tools that identify
+structured data in a web page, and use that structure to support end
+user modification of the page.
+
+Sifter \autocite{huynh2006} enables users to sort and filter lists of
+data on web pages, providing a result similar to Wildcard's sort and
+filter functionality. The underlying mechanism is also similar: Sifter
+extracts structured data from the page to enable its user-facing
+functionality. Wildcard aims to extend this approach to support much
+broader functionality than just sorting and filtering. In support of
+this goal, Wildcard also shows the structured data table directly to the
+user, whereas Sifter only shows sort and filter controls, without
+revealing the underlying data table. The extraction mechanism is also
+different: Sifter uses a combination of automated heuristics and
+interactive user feedback to extract data, whereas Wildcard currently
+relies on programmers creating wrappers for extracting structured data,
+likely leading to higher quality extraction but on fewer sites.
+
+Thresher \autocite{hogue2005} enables users to create wrappers which map
+unstructured website content to Semantic Web content. Like Wildcard and
+Sifter, Thresher augments the experience of original page based on
+identifying structure: once semantic content has been identified, it
+creates context menus in the original website which allow users to take
+actions based on that content. Wildcard and Thresher share an overall
+approach but focus on complementary parts of the problem: Thresher aims
+to enable end users to create content wrappers, but the actions
+available on the structured data are created by programmers; conversely,
+Wildcard delegates wrapper creation to programmers but gives end users
+more flexibility to use the structured data in an open-ended way.
+
+\hypertarget{sloppy-augmentation}{%
+\subsubsection{Sloppy augmentation}\label{sloppy-augmentation}}
+
+``Sloppy programming'' \autocite{little2010} tools like Chickenfoot
+\autocite{bolin2005} and Coscripter \autocite{leshed2008} enable users
+to create scripts that perform actions like filling in text boxes and
+clicking buttons, without directly interacting with the DOM. Users
+express the desired page elements in natural, informal terms
+(e.g.~writing ``the username box'' to represent the textbook closest to
+the label ``username''), and then using heuristics to determine which
+elements most likely match the user's intent. This approach allows for
+expressing a wide variety of commands with minimal training, but it also
+has downsides. It is difficult to know whether a command will
+consistently work over time (in addition to changes to the website,
+changes to the heuristics can also cause problems), and it is not easy
+for users to discover the space of possible commands.
+
+Wildcard offers a sharp contrast to sloppy programming, instead choosing
+to expose a high degree of structure through the familiar spreadsheet
+table. Wildcard offers more consistency: for example, clicking a sort
+header will always work correctly as long as the site adapter is
+maintained. Wildcard also offers clearer affordances for what types of
+actions are possible, or, crucially, what actions are \emph{not}
+possible, which is useful to know. On the other hand, Wildcard cannot
+offer coverage of all websites, and has a narrower range of possible
+actions than sloppy tools. We expect that with enough site adapters and
+formulas, these downsides can be mitigated.
+
+\hypertarget{spreadsheet-based-app-builders}{%
+\subsection{Spreadsheet-based app
+builders}\label{spreadsheet-based-app-builders}}
+
+It is a powerful realization to notice that a spreadsheet can serve as
+an end-user-friendly backing data store and computation layer for an
+interactive web application. Research projects like Quilt
+\autocite{benson2014}, Gneiss \autocite{chang2014}, and Marmite
+\autocite{wong2007}, as well as commercial tools like Airtable Blocks
+\autocite{zotero-79} and Glide \autocite{zotero-81} allow users to view
+data in a spreadsheet table, compute over the data using formulas, and
+connect the table to a GUI. Because many users are already familiar with
+using spreadsheets, this way of creating applications tends to be much
+easier than traditional software methods; for example, in a user study
+of Quilt, many people were able to create applications in under 10
+minutes, even if they expected it would take them many hours.
+
+Wildcard builds on this powerful idea, but applies it to modifying
+existing applications, rather than building new applications from
+scratch. For many people, we suspect that tweaking existing applications
+provides more motivation as a starting point for programming than
+creating a new application from scratch.
+
+An important design decision for tools in this space is how to deviate
+from traditional spreadsheets. Quilt and Glide use existing web
+spreadsheet software as a backend, providing maximum familiarity and
+compatibility with existing spreadsheets. Gneiss has its own spreadsheet
+implementation, with additional features useful for building GUIs.
+Marmite provides a live data view that resembles a spreadsheet, but
+programming is actually done using a separate data flow pane rather than
+spreadsheet formulas. (Marmite's approach led to some confusion in a
+user study, because users expected behavior more similar to spreadsheets
+\autocite{wong2007}.) Airtable deviates the furthest from spreadsheets:
+although the user interface has formulas, shows live feedback and is
+easy to use, the underlying structure is a relational database with
+typed columns. Wildcard's table is most similar to Airtable; the
+structure of a relational table is most appropriate for most data in
+websites, and we have not yet found a need for arbitrary untyped cells.
+
+\hypertarget{web-scraping-data-extraction}{%
+\subsection{Web scraping / data
+extraction}\label{web-scraping-data-extraction}}
+
+Web scraping tools focus on extracting structured data out of
+unstructured web pages. Web scraping is closely related to the
+implementation of Wildcard, but has different end goals: web scraping
+generally extracts static data for processing in another environment,
+whereas Wildcard modifies the original page by maintaining a
+bidirectional connection between the extracted data and the page.
+
+Web scraping tools differ in how much structure they attempt to map onto
+the data. Some tools like Rousillon \autocite{chasins2018} extract data
+in a minimally structured relational format; other tools like Piggy Bank
+\autocite{huynh2005} more ambitiously map the data to a rich semantic
+schema. In Wildcard, we chose to avoid schemas, in order to minimize the
+work associated with creating a site adapter.
+
+In the future, we might be able to integrate web scraping tools to help
+create more reliable site adapters for Wildcard with less work, and to
+open up adapter creation to end users. Sifter was built on top of the
+Piggy Bank scraping library, suggesting precisely this type of
+architecture where web scraping tools are used to support interactive
+page modification.
+
+\hypertarget{limitations-and-future-work}{%
+\section{Limitations and future
+work}\label{limitations-and-future-work}}
+
+Wildcard is still in early development; there are still many limitations
+to resolve and open questions to explore.
+
+The most important question is whether the combination of features
+provided by Wildcard can be combined in enough useful ways to make the
+system worth using in practice. While initial demos are promising, we
+need to develop more site adapters and use cases to more fully assess
+this question. We plan to continue privately testing the system with our
+own needs, and to eventually deploy the tool publicly, once the API is
+stable enough and can support a critical mass of sites and use cases. We
+also plan to run usability studies to evaluate and improve the design of
+the tool.
+
+Here are a few of the largest limitations in the current system:
 
@@
 \item
-  Instrumental interaction
+  Wildcard only extracts data visible on the page, which means that
+  subsequent pages of results in paginated lists are not included in the
+  table. (Sifter \autocite{huynh2006} uses techniques that might help
+  get around this.)
 \item
-  Web automation
+  There is no mechanism for end users to express imperative workflows
+  (e.g.~``click this button, and then\ldots{}''); they can only write
+  formulas that return data and then inject the resulting data into the
+  page. While this makes the system simpler, it also may exclude many
+  valuable use cases. We may add a system for expressing workflows like
+  this, although it's not obvious how it would fit together with the
+  existing table view. Gneiss \autocite{chang2014} contains some
+  mechanisms for writing spreadsheet formulas which can handle behaviors
+  like reloading data in response to a user pressing a button, which
+  might prove helpful.
 \item
-  Wrapper induction
+  Wildcard's data model only shows a single table at a time, without any
+  notion of relationships between tables. A richer data model with
+  foreign keys might be necessary to support certain use cases. For
+  designing a tabular interface on top of a richer data model, we could
+  learn from the interface of Airtable which shows related objects in a
+  table cell, or SIEUFERD \autocite{bakke2016} which uses nested rows.
 \item
-  Personal data ownsership / SOLID
-\item
-  Extension helper libraries, e.g.~Gmail.js.
+  Only allowing site adapters to be manually coded means that the number
+  of supported sites is limited. As mentioned previously, we may explore
+  also building wrappers using automated heuristics or end user scraping
+  tools.
 \end{itemize}
 
-\hypertarget{future-work}{%
-\section{Future work}\label{future-work}}
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
-\begin{itemize}
-\tightlist
-\item
-  still in early development; note the beta release plan (tentative:
-  target public beta availability at the workshop in March?)
-\item
-  What are the limits of functionality? What can't it do?
+We live in a time of pessimism about the effects of the internet,
+especially given the degree of centralized control that large
+corporations wield. We think that one promising solution is to enable
+more people to have deeper control over the Web software they use every
+day. The Web offers a promising opportunity of an open foundation, but
+even the success of browser extensions has only empowered users within
+the limited confines of the ideas that programmers have decided to
+build.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Should we extend further? Automation workflows? Triggers?
-  \end{itemize}
-\item
-  Could explore automated wrapper induction, building on prior work
-\item
-  Usability studies
-\item
-  Link to sign up for future updates
-\end{itemize}
+In this paper, we have presented Wildcard, a browser-embedded
+programming system that maps websites to a structured data table,
+enabling end users to modify their behavior. We hope that it contributes
+to making the web into a more dynamic medium that users can mold to
+their own needs.
+
+We plan to continue developing the system and to eventually deploy it as
+an open-source tool. To receive future updates on Wildcard and
+notifications about a public release, \href{}{sign up for the email
+newsletter}.
+
+We are also looking for private beta testers. If you have an idea for
+how you might want to use Wildcard, \href{mailto:glitt@mit.edu}{please
+contact us}. We would love to collaborate on building site adapters and
+formulas to support your use case.
 

### 2020-01-07 compile-html.sh

@@ -1,6 +1,6 @@
 set -e
 
-cat html.sections | xargs pandoc \
+ pandoc \
   --filter conditional-render \
   --filter pandoc-citeproc \
   --metadata=format:html \
@@ -10,7 +10,8 @@ cat html.sections | xargs pandoc \
   --toc \
   --toc-depth=2 \
   --variable=toc-title:"Contents" \
-  --template=templates/pandoc-template-html.html
+  --template=templates/pandoc-template-html.html \
+  paper.md
 
 cp ./paper.html ~/dev/homepage/source/wildcard/index.html
 cp ./basic.css ~/dev/homepage/source/wildcard

### 2020-01-07 compile-latex.sh

@@ -1,6 +1,6 @@
 set -e
 
-cat latex.sections | xargs pandoc \
+pandoc \
   --filter conditional-render \
   --filter pandoc-crossref \
   --filter pandoc-citeproc \
@@ -9,7 +9,9 @@ cat latex.sections | xargs pandoc \
   paper.tex \
   --template=templates/pandoc-template-programming.pandoc \
   --biblatex \
-  --metadata=format:pdf
+  --metadata=format:pdf \
+  paper.md
+
 pdflatex paper.tex
 biber paper
 pdflatex -interaction=batchmode paper.tex

### 2020-01-07 paper.tex
in progress

@@
 
-\title{Wildcard: End user modification of web applications through a
+\title{Wildcard: End user modification of web applications using a live
 data table}
@@
 \begin{abstract}
-Many Web applications do not meet the precise needs of their users.
+Many Web applications do not meet the particular needs of their users.
 Browser extensions and user scripts offer a way to customize web
@@
 
-In this paper, we present a prototype of Wildcard, a live programming
-system embedded in the browser that empowers users to casually tweak web
-applications without needing to program in Javascript or interact with
-the DOM.
+In this paper, we present a prototype of Wildcard, an extension
+framework embedded in the browser that empowers users to casually tweak
+web applications without needing to program in Javascript or directly
+interact with the DOM.
 
-Wildcard shows the data from a web page in a table, and maintains a
+Wildcard shows the data from a web page in a table, and maintains a live
 bidirectional connection between the table and the original page. By
@@
 modifications: sorting/filtering content, adding private annotations,
-using spreadsheet formulas to fetch data from other web services, using
+using spreadsheet formulas to fetch data from other web services, adding
 custom UI elements to edit form data, and more.
@@
 explain the design principles behind the prototype. In the future, we
-envision continuing to build Wildcard into a fully deployed system that
-helps make the web into a more malleable medium for all users.
+envision growing Wildcard into a fully deployed system that helps make
+the web into a more malleable medium for all its users.
 \end{abstract}
@@
 frequently updated without user consent. Sometimes there is a browser
-extension or user script that fixes an issue, and if the user is a
-motivated, programmer they might even be able to implement their own
+extension or user script that fixes an issue, and if the user is both
+motivated and skilled they might even be able to implement their own
 fix. But for most people in most situations, the only recourse is to
@@
 quo, we see it as a tremendous waste of the openness of the Web
-platform. In \emph{Personal Dynamic Media} \autocite{kay1977}, Alan Kay
-originally envisioned personal computing as a medium that let a user
-``mold and channel its power to his own needs,'' but today's software is
-more like concrete than clay.
+platform. Back in 1977, in \emph{Personal Dynamic Media}
+\autocite{kay1977}, Alan Kay had originally envisioned personal
+computing as a medium that let a user ``mold and channel its power to
+his own needs,'' but today's software is more like concrete than clay.
 
-In this paper, we introduce Wildcard, a live programming system embedded
-in the browser that enables users to tweak web applications without
-programming in Javascript or interacting with the DOM. Wildcard adds a
-panel to the bottom of a web page that shows a structured table view of
-the main data in the page. The table maintains a bidirectional
-connection to the original page---when the user manipulates the table,
-the original page gets modified, and vice versa.
+In this paper, we introduce Wildcard, an extension framework embedded in
+the browser that lets users tweak web applications without programming
+in Javascript or interacting with the DOM. Wildcard shows a structured
+table view of the main data in a webpage, which maintains a live
+bidirectional connection to the original UI: when the user manipulates
+the table, the UI is instantly modified, and vice versa. This simple
+framework allows a wide range of possible modifications, ranging from
+sorting lists to fetching related data from other websites.
 
-In Wildcard, a user can sort Airbnb listings with just one intuitive
-click on a table header, with no programming required. Beyond sorting
-and filtering data, Wildcard also supports accessing third party APIs,
-performing small computations, recording private user annotations, and
-using alternate UI widgets to control a page. While Wildcard does not
-support any change someone might want to make to a website, it makes a
-variety of common and useful changes more accessible to end users, as
-well as to programmers who want to avoid tedious low-level programming.
+The system builds on two insights from prior work. The first insight is
+that a spreadsheet can serve as the backend for a GUI application, which
+enables an end user to build applications in an extremely accessible
+environment without doing traditional programming. The second insight is
+that web scraping---extracting structured data from an unstructured web
+page---unlocks value by making it possible to compute over that data.
+Our contribution in this work is to combine those two ideas in a novel
+way. By extracting structured data from an application and showing that
+data in a live spreadsheet table, Wildcard enables a wide class of end
+user modifications to existing applications. Wildcard makes it seem like
+an application is backed by a spreadsheet, even though the actual
+implementation is not. This allows the user to easily manipulate and
+compute over the application's underlying data just as they would in a
+spreadsheet, and then to immediately see the results of their changes in
+the original application's UI.
 
-Under the hood, the implementation of Wildcard is straightforward,
-because a programmer must manually write an adapter for each website
-which uses common web scraping techniques to map the data from the web
-page into the Wildcard table format. Although there is still manual
-programming involved, the result is very different from traditional
-single-use browser extensions, because instead of the programmer
-defining a narrow use case, the end user is able to make many different
-changes on top of a single site-specific adapter.
+Wildcard embodies several principles we propose for designing systems
+that enable end user modification of existing software:
 
-In this paper, we present examples of using Wildcard to solve real world
-problems, and then explain the system architecture and design principles
-behind the prototype:
+\emph{Expose consistent structure}: Most applications only expose a user
+interface, without showing the structure of the data underneath.
+Furthermore, every application has a different interface, requiring
+users to learn many ways of using apps. Wildcard shows data from diverse
+applications in a single consistent structure, so that users are able to
+invest in learning one mechanism for manipulating the data in their
+applications, and for programming with that data.
 
-\begin{itemize}
-\tightlist
-\item
-  \emph{Expose consistent structure}: turning isolated applications into
-  a structured medium
-\item
-  \emph{Encourage casual tweaking}: using frequent software modification
-  as a starting point for end user programming
-\item
-  \emph{Build for an ecosystem}: designing for first party developers,
-  third party programmers, and non-programmer end users
-\end{itemize}
+\emph{Low floor, high ceiling}: We aim to provide a low barrier to entry
+so that non-programmers can casually use the system, but a high ceiling
+on the complexity it can achieve. Small but genuinely valuable tweaks
+like sorting a page are made possible with little effort in the flow of
+normal use to provide a low floor, and a rich formula system creates a
+high ceiling on possible modifications.
 
-We also explore related work on web augmentation tools,
-spreadsheet-based app builders, and web scraping tools. Our novel
-contribution in this paper is to combine ideas from each of these areas
-into a new kind of tool: a programming environment which maps data from
-web pages onto a live table view, enabling end users to modify the data.
+\emph{Build for multiple tiers of users}: We aim to make Wildcard as
+easy as possible for end users, by designing roles for tiers of users
+with differing levels of technical sophistication. Programmers or
+tech-savvy end users are responsible for the data extraction step of
+mapping existing UIs to structured data. Most end users only interact
+with the structured data, providing a predictable and user-friendly
+experience.
 
@@
 other plugins. Ultimately, we hope that Wildcard makes modifying
-websites into a natural part of using the web, as mundane and ubiquitous
-as using spreadsheets.
+websites into a natural part of using the web, as ubiquitous as using
+spreadsheets in everyday computing.
 
-\hypertarget{demos}{%
-\section{Demos}\label{demos}}
+\hypertarget{demo-booking-a-trip-with-wildcard}{%
+\section{Demo: booking a trip with
+Wildcard}\label{demo-booking-a-trip-with-wildcard}}
 
-To get a sense of how it feels to use Wildcard, let's see an example of
-using it to help with booking a trip using the travel sites Airbnb and
-Expedia.
+To get a sense of how it might feel to use Wildcard, let's see an
+example of someone using it to help with booking a trip on the travel
+sites Airbnb and Expedia.
 
-\hypertarget{sorting-and-filtering}{%
-\subsection{Sorting and filtering}\label{sorting-and-filtering}}
+The user starts by opening up the Airbnb search listings page to look
+for a place to stay. The page looks nice and mostly works well, but is
+missing some key features. As mentioned before, this page doesn't allow
+the user to sort by price. It also doesn't let them filter by user
+rating. Using Wildcard, the user can add these small features, while
+leaving the page's design and the rest of its functionality unchanged.
 
-We start by opening up the Airbnb search listings page to look for a
-place to stay. As mentioned before, this page doesn't let us sort by
-price, but we can use Wildcard to fix that. First, we open up the
-Wildcard panel, which shows a table corresponding to the search results
-in the page. As we click around in the table, the corresponding row in
-the page is highlighted so we can see the connection between the views.
+First, the user opens up the Wildcard panel, which shows a table
+corresponding to the search results in the page. As they click around in
+the table, the corresponding row in the page is highlighted so they can
+see the connection between the views.
 
-When we click the Price column header in the table, both the table and
-the original page become sorted by price. We can also filter listings by
-rating, another feature not offered in the Airbnb site.
+Then, the user can use standard spreadsheet column header features to
+sort the page by price and filter by rating:
 
-After finishing the sorting and filtering, we can close the table view
-and continue using the website in its original polished design.
-
-\hypertarget{row-actions}{%
-\subsection{Row actions}\label{row-actions}}
+Notice how after manipulating the data, the user was able to close the
+table view and continue using the website with its original visual
+design. The table view offers a way to change the data backing a page,
+but does not need to replace the original interface entirely.
 
@@
 
-Within the site adapter, each action is implemented as a regular
-Javascript function running in the context of the page, which can click
-on buttons in the UI, launch AJAX requests, navigate to a new page, etc.
+Now the user would like to jot down some notes on the pros and cons of
+each listing. To do this, they can simply type notes into an additional
+column next to each listing, and the notes appear inside the listings in
+the original UI. These annotations are also persisted in the browser for
+future retrieval.
 
-\hypertarget{user-annotations}{%
-\subsection{User annotations}\label{user-annotations}}
-
-It's a common practice in spreadsheets to add a column next to a table
-for recording notes about the item in each row. In Wildcard, this
-pattern can be used to create private annotations in a website. Here, we
-use this feature to jot down pros and cons of various listings:
-
-In order to match the annotations with the existing design, the site
-adapter specifies a reasonable location in each row for injecting new
-data.
-
-Annotations are persisted in browser local storage so the user can come
-back and view them later. The site adapter associates each table row
-with a stable identifier from the original site so that annotations can
-be displayed with the appropriate item.
-
-Currently annotations are private to the browser of the user that
-created them, but it could be useful in the future to allow users to
-share annotations.
-
-\hypertarget{computation-with-formulas}{%
-\subsection{Computation with formulas}\label{computation-with-formulas}}
-
-The demos so far have shown straightforward tweaks that provide useful
-conveniences with little effort. But Wildcard also supports more
-sophisticated tweaks that fetch external data and perform computations,
-using a formula language.
+Wildcard also includes a formula language which enables more
+sophisticated tweaks that fetch external data and perform computations.
 
@@
 stay based on how walkable the surroundings. Using Wildcard formulas, we
-can integrate Airbnb with Walkscore, an API that can rate the
-walkability of any location on a 1-100 scale. When we call the
-\texttt{WALKSCORE} formula with the latitude and longitude of the
-listing, it returns the score as the cell value. Because the cell is
-shown in the page, the score also shows up in the page body.
-
-Wildcard formulas are just Javascript functions that take table data as
-input and return new data. Formulas can have limited side effects like
-accessing external APIs to fetch data, but do not directly manipulate
-the page. Instead, they return data into the table, which is then
-injected into the page with the same mechanism used for injecting user
-annotations.
-
-\hypertarget{custom-ui-elements}{%
-\subsection{Custom UI elements}\label{custom-ui-elements}}
+can integrate Airbnb with Walkscore, an API that rates the walkability
+of any location on a 1-100 scale. When we call the \texttt{WALKSCORE}
+formula with the latitude and longitude of the listing, it returns the
+walk score as the cell value. Because the cell's contents are injected
+into the page, the score also shows up in the page body.
 
@@
 In previous examples the data extracted from the site was marked as
-read-only (it's not allowed to change the name of an Airbnb listing),
-but in this case, the cells are marked as writable, so that changes in
-the table are reflected in the original form. This becomes particularly
-useful when combined with custom cell editing UIs, which allow users to
-graphically edit the value of a cell.
+read-only; users cannot change the name or price of an Airbnb listing.
+In this next case, the cells are marked as writable, so that changes in
+the table are reflected in the form inputs. This becomes useful when
+combined with GUI widgets for editing the value of a table cell.
 
-Here's an example of using those features to help with filling in a
-form. Filling in dates for a flight search typically requires opening up
-a separate calendar app to look up the correct dates, and then manually
-copying them into the form. In Wildcard, we can make this easier by
-editing the date cell using a datepicker widget that has privileged
-access to our calendar information.
+Filling in dates for a flight search typically requires opening up a
+separate calendar app to find the right dates, and then manually copying
+them into the form. In Wildcard, we can make this easier by providing a
+datepicker widget that has privileged access to the user's calendar
+information.
 
-Custom UIs enable people to use a consistent UI to enter common types of
-data across the web. They also allow a user to access their own private
-data as part of a web interface, without needing to expose it to the
-website server.
-
-Overall, an interactive data table is a computation model that presents
-a surprisingly large range of useful possibilities for end user
-modification of websites, while also remaining familiar and easy to use.
-While we've presented concretely illustrated here just a few of these
-possibilities, we plan to continue developing site adapters, formulas,
-and UI elements to explore more use cases, and to eventually publicly
-release the tool to allow real end users to discover their own uses.
+Here we've presented just a few possibilities for how to use Wildcard.
+We think the interactive data table offers a flexible computational
+model that can support a wide range of other useful modifications, all
+while remaining familiar and easy to use.
 
@@
 Airbnb listing. It specifies the field name and some metadata
-properties, and then a function that extracts the name div from a row
-div.
+properties, and then a function that, given the entire listing, returns
+the DOM element representing the name of the listing.
 
@@
 DOM to reflect the table state. The only exception is site-specific
-actions (like favoriting in Airbnb) which runs Javascript code that can
-mutate the DOM or perform other arbitrary behaviors.
-
-So far we have only implemented three site adapters in Wildcard, but we
-plan to implement many more in the near future to continue to test the
-limits of the API. Eventually we will also encourage third party
-programmers to build and maintain adapters for popular sites.
+actions (like favoriting in Airbnb): actions are implemented as regular
+Javascript functions running in the context of the page, which can
+mutate the DOM, simulate clicks on buttons in the UI, launch AJAX
+requests, navigate to new pages, etc.
 
@@
 prior work and our own experimentation. We think these principles extend
-beyond this specific system and can more generally inform the design of
-end user programming tools, particularly systems enabling users to
-modify GUI apps.
+beyond Wildcard and might inform the design of other end user
+programming tools, particularly systems enabling users to modify GUI
+apps.
 
@@
 
-In \emph{Changing Minds} \autocite{disessa2000}, Andrea diSessa
-critiques the design of modern software with a story about a
-hypothetical ``nightmare bike.'' Each gear on the nightmare bike is
-labeled not with a number, but with an icon describing its intended use:
-smooth pavement uphill, smooth pavement downhill, gravel, etc. This
-might seem more ``user-friendly'' than numbered gears, but in fact, it
-makes it harder to operate the bike. An ordered numerical sequence
-allows the user to intuitively grasp the structure of the system---``if
-I shift up, it gets harder to pedal''---but isolated modes provide a
-superficial understanding with no grounding in structure. This
-understanding might be sufficient for the most common cases, but breaks
-down in unfamiliar situations. If someone needs to go uphill on gravel,
-do they need to try every mode at random?
+Today, most personal computing consists of using distinct apps, each one
+an isolated silo of data and functionality. While there are some shared
+idioms and limited points of interoperability, apps fundamentally
+operate independently from one another. Computing does not need to
+operate this way---for example, UNIX offers a compelling alternate
+vision: many small tools, all built on the consistent abstraction of
+text files. Users can invest in deeply mastering a text editor, and then
+use it for a vast array of tasks, even as an interactive input mechanism
+in shell programs (e.g.~for editing git commit messages).
+Beaudouin-Lafon and Mackay call this style of interaction
+\emph{polymorphic} \autocite{beaudouin-lafon2000}, since a single
+interface widget can be used in many contexts. diSessa
+\autocite{disessa2000} has also noted the deep connection between
+polymorphism and literacy: textual literacy rests on a single rich
+medium of writing, which once mastered, can be adapted to many different
+genres and uses. If people needed to relearn from scratch when switching
+from writing essays to writing emails, the medium would lose most of its
+potency.
 
-Today, most personal computing consists of using distinct apps, each one
-an isolated silo of concepts, interfaces, data, and functionality. While
-there are some shared idioms and limited points of interoperability, the
-fundamental expectation is that each app operates independently. If a
-new need emerges, a user must hope ``there's an app for that''; creating
-a new tool to meet the need is often out of the question.
-
-Computing does not need to operate this way. As one example, UNIX offers
-a compelling alternate vision of small tools connected through a common
-abstraction of text files. Just like riding a bike with numbered gears,
-UNIX requires users to build up a structured understanding of the
-system, which then allows them to flexibly react to new needs not served
-by existing tools. Users can also deeply master a text editor and use it
-for a vast array of tasks, even as an interactive input mechanism in
-shell programs (e.g.~for editing git commit messages). Beaudouin-Lafon
-and Mackay call this style of interaction \emph{polymorphic}
-\autocite{beaudouin-lafon2000}, since a single interface can be used in
-many contexts. diSessa \autocite{disessa2000} notes the deep connection
-between polymorphism and literacy: textual literacy rests on a single
-rich medium of writing, which once mastered, can be adapted to many
-different genres and uses. If people needed to relearn from scratch when
-switching from writing essays to writing emails, the medium would lose
-most of its potency; this is the situation in software today.
-
-With Wildcard, we hope to make isolated Web applications work a little
-bit more like UNIX, by imposing a consistent and simple structure onto
-many isolated applications. Just as UNIX uses text files to represent
-all data, Wildcard uses relational tables. This abstraction strikes a
-balance between being simple and generic: a data table is much simpler
-than the DOM tree that describes all the details of the UI, but is also
-generic enough to describe the essence of many different interfaces.
+With Wildcard, we hope to make isolated Web applications work more like
+UNIX, by imposing a consistent and simple data structure onto many
+applications. Just as UNIX uses text files to represent all data,
+Wildcard uses relational tables. This abstraction strikes a balance
+between being simple and generic: a data table is much simpler than the
+DOM tree that describes all the details of the UI, but is also generic
+enough to describe the essence of many different interfaces.
 
@@
 
-Exposing more structured data to users provides improved quality and
-consistency of user interfaces. When UI widgets can compete on their
-merits rather than having a monopoly on a given data silo or
+Exposing structure gives users more choice When UI widgets can compete
+on their merits rather than having a monopoly on a given data silo or
 application, it creates stronger competition at the interface level. For
@@
 about these benefits \autocite{berners-lee2018} in the context of the
-SOLID project, which envisions user-controlled data as a mechanism for
-decoupling data from interfaces, e.g.~giving users a choice of which
-client to use to consume a given social media feed. Wildcard has
-overlapping goals with SOLID, but does not require decentralized user
-control of data---the data can remain on a centralized server, as long
-as the interface can be tweaked by end users.
+SOLID project, which envisions giving users control of their own data as
+a mechanism for decoupling data from interfaces (e.g., giving users a
+choice of which client to use to consume a given social media feed).
+Wildcard has overlapping goals with SOLID, but does not require
+decentralized user control of data---the data can remain on a
+centralized server, as long as the interface can be tweaked by end
+users.
+
+\emph{todo: move the sloppy contrast, and SOLID, down into related work}
 
 Layering a structured view on top of an existing application is not the
-only option available. Systems like Chickenfoot \autocite{bolin2005} and
-CoScripter \autocite{leshed2008} forego the intermediate layer of
-structure, instead allowing users to create scripts in an informal
-language and then perform fuzzy pattern matching to find elements in the
-DOM. For example, to find a button after a textbox in Chickenfoot, the
-user could type \texttt{click(find(“button\ just\ after\ textbox”))}.
-These designs allow for expressing a wide range of operations, but they
-don't explicitly indicate what operations are possible---the user can
-only see the original page and imagine the possibilities. In contrast,
-Wildcard provides affordances that clearly suggest the availability of
-certain actions (e.g.~sorting, editing a cell, adding a column with a
-derived value), especially to users who are familiar with spreadsheets.
-In addition to giving users more certainty about whether a modification
-is possible, these affordances might improve discoverability and give
-users new ideas for things to try. We also think that sloppy programming
-does little to solve the problem of thinking of applications as isolated
-worlds, whereas Wildcard contributes to giving users a structured
-understanding of the different apps that they use.
+only option available. ``Sloppy programming'' systems like Chickenfoot
+\autocite{bolin2005} and CoScripter \autocite{leshed2008} forego the
+intermediate layer of structure, instead allowing users to create
+scripts in an informal language and then perform fuzzy pattern matching
+to find elements in the DOM. For example, to find a button after a
+textbox in Chickenfoot, the user could type
+\texttt{click(find(“button\ just\ after\ textbox”))}. These designs
+allow for expressing a wide range of operations, but they don't
+explicitly indicate what operations are possible---the user can only see
+the original page and imagine the possibilities. In contrast, Wildcard
+provides affordances that clearly suggest the availability of certain
+actions (e.g.~sorting, editing a cell, adding a column with a derived
+value), especially to users who are familiar with spreadsheets. In
+addition to giving users more certainty about whether a modification is
+possible, these affordances might improve discoverability and give users
+new ideas for things to try. ``Sloppy programming'' does give end-users
+greater access to modify the apps that they use, but does not seem to
+provide them with a deeper understanding of those apps or how they
+relate to each other, so the apps are more likely to remain siloed.
 
@@
 
-\hypertarget{encourage-casual-tweaking}{%
-\subsection{Encourage casual tweaking}\label{encourage-casual-tweaking}}
+\hypertarget{low-floor-high-ceiling}{%
+\subsection{Low floor, high ceiling}\label{low-floor-high-ceiling}}
 
-In order to serve the widest possible audience, Wildcard is designed for
-tweaking existing software for one's own needs, not for designing a new
-application from scratch. Creating a new application requires motivation
-and other skills besides programming that not everyone has, but having
-frustrations with some aspect of an app which could be solved by a small
-tweak is a nearly universal experience.
+Seymour Papert advocated for programming systems with a ``low floor''
+and a ``high ceiling''---making it easy for novices to get started, but
+also providing possible a large range of possibilities for more
+sophisticated users \autocite{resnick2016}. Unfortunately, many
+programming systems only provide either one or the other
+\autocite{myers2000}.
 
-\begin{longtable}[]{@{}lll@{}}
+\begin{itemize}
+\tightlist
+\item
+  tweaking vs creation from scratch
+\item
+  in-place toolchain
+\item
+  easy things easy
+\item
+  easy things valuable
+\item
+  this seems clearly justified by the demos
+\end{itemize}
+
+rewrite\ldots{}
+
+\begin{longtable}[]{@{}llc@{}}
 \toprule
 \endhead
-& \emph{Casual} & \emph{Not casual}\tabularnewline
+& \emph{In-place} & \emph{Not in-place}\tabularnewline
 \emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
@@
 First, the technical capability required of the user: is programming
-knowledge needed? Second, the level of effort required: how far out of
-their way the user must go to make a change? Can they casually make a
-small change, or do they need to make a larger project out of it? These
-dimensions are not orthogonal, but they are distinct. For example,
-setting up a workflow trigger in an end user programming system like
+knowledge needed? Second, whether the change is ``in-place''
+\autocite{inkandswitch2019}: can it be made within the same environment
+that the software is typically used in? These dimensions are not
+orthogonal, but they are distinct. For example, setting up a workflow
+trigger in an end user programming system like
 \href{https://ifttt.com/}{IFTTT} does not require much technical skill,
@@
 
-In addition to requiring no programming skills, Wildcard also aims to
-support frequent, small modifications. The Wildcard table appears in the
-course of normal web browsing, to ensure that the tools for modification
-are close at hand while using the original software. Ink and Switch
-refers to this property as having an ``in-place toolchain''
-\autocite{inkandswitch2019}. Wildcard also provides live feedback,
-immediately editing the page in response to user interaction, which
-gives users confidence in the changes they are making.
+In addition to requiring no programming skills, Wildcard supports
+frequent, small modifications. The Wildcard table appears in the course
+of normal web browsing, to ensure that the tools for modification are
+close at hand while using the original software.Wildcard also provides
+live feedback, immediately editing the page in response to user
+interaction, which gives users confidence in the changes they are
+making.
 
-Wildcard also aims to make some valuable changes possible with
-particularly low effort: for example, sorting a table in a single click,
-or adding an annotation by filling in a cell. We would expect many
-Wildcard users to start by using these bits of functionality, before
-possibly moving on to more complex features like formulas. This property
-is inspired by spreadsheets, which can be used productively even by
-someone who has learned only a tiny part of their functionality,
-e.g.~for storing tables of numbers or computing simple sums
-\autocite{nardi1991}. In contrast, many traditional programming systems
-require someone to spend days learning complex concepts before being
-able to create a useful piece of software---printing ``hello world'' is
-not a genuinely valuable activity.
+Spreadsheets can be used in genuinely useful ways even by someone who
+has learned only a tiny part of their functionality, e.g.~for storing
+tables of numbers or computing simple sums \autocite{nardi1991}. This
+creates inherent motivation to continue using the tool and eventually
+learn its more powerful features. In contrast, many traditional
+programming systems require an enormous time investment before a
+beginner is able to create any valuable software. Like spreadsheets,
+Wildcard aims to support genuinely useful actions with little effort:
+for example, sorting a table with a single click, or adding an
+annotation by typing into a cell. We would expect many Wildcard users to
+start by using these bits of functionality, before possibly moving on to
+more complex features like formulas.
 
-\hypertarget{build-for-an-ecosystem}{%
-\subsection{Build for an ecosystem}\label{build-for-an-ecosystem}}
+High ceiling?
+
+\hypertarget{build-for-multiple-tiers-of-users}{%
+\subsection{Build for multiple tiers of
+users}\label{build-for-multiple-tiers-of-users}}
+
+\begin{itemize}
+\tightlist
+\item
+  make things easy for the last tier of users (low floor)
+\item
+  expose predictable structure, don't make them participate in data
+  extraction
+\end{itemize}
 
@@
 for. At the most extreme, this can even include modifications that are
-hostile to the desires of the original developers, a notion Cory
-Doctorow calls
-\href{https://www.eff.org/deeplinks/2019/10/adversarial-interoperability/}{adversarial
-interoperability}. Although ad blocking extensions make this style of
-extension very prominent, most ideas for extending software merely fall
-outside the range of possibilities originally imagined by the developer,
-rather than being actively hostile.
+hostile to the desires of the original developers. Although ad blocking
+extensions make this style of extension very prominent, most ideas for
+extending software merely fall outside the range of possibilities
+originally imagined by the developer, rather than being actively
+hostile.
 
 Wildcard takes advantage of this openness, and does not depend on
-cooperation from first-party website developers. Any programmer can add
+cooperation from first-party website developers: any programmer can add
 support for any website to Wildcard by building a third party adapter.
-However, we also hope that eventually first party website developers
-will build in Wildcard support to their applications, since this would
-reduce the burden of maintaining adapters and make Wildcard plugins more
-stable. Implementing the Wildcard adapter API could help developers by
-allowing users to fix some of their own issues, and would not
-necessarily require too much effort---we envision making it
-straightforward to add Wildcard support in a client-side application
-which already has access to the structured data in the page. There is
-also precedent for first parties implementing an official client
-extension API in response to user demand: for several years, Google
-maintained an official extension API in Gmail for Greasemonkey scripts
-to use. (Incidentally, since then, third parties have continued to
-maintain stable Gmail extension APIs used by many browser extensions
-\autocite{streak,talwar2019}, illustrating the potential of
-collaboratively maintaining third party adapters.)
+But the platform is still open to first parties developing their own
+adapters, which would often be in their interests. Providing Wildcard
+support would allow users to build extensions to fulfill their own
+feature requests. It also would not necessarily require much effort:
+adding Wildcard support would be more straightforward for a first-party
+than a third-party because they have direct access to the structured
+data in the page. There is also precedent for first parties implementing
+an official client extension API in response to user demand: for several
+years, Google maintained an official extension API in Gmail for
+Greasemonkey scripts to use. (Incidentally, since then, third parties
+have continued to maintain stable Gmail extension APIs used by many
+browser extensions \autocite{streak,talwar2019}, illustrating the
+potential of collaboratively maintaining third party adapters.)
 
@@
 context, although it also requires working within rigid constraints. We
-also see Wildcard growing beyond its current bounds in the future and
-becoming more similar to these other systems. With substantial future
+also see Wildcard growing beyond its current bounds and becoming more
+similar in the future to these other systems. With substantial future
 work, Wildcard could grow from merely being a platform for tweaking
@@
 
-Wildcard's goals are closely shared with other systems which provide
+Wildcard's goals are closely shared with other systems that provide
 interfaces in the browser for end users to augment and modify websites
@@
 
-It is a powerful realization to notice that a spreadsheet can serve as
-an end-user-friendly backing data store and computation layer for an
-interactive web application. Research projects like Quilt
-\autocite{benson2014}, Gneiss \autocite{chang2014}, and Marmite
-\autocite{wong2007}, as well as commercial tools like Airtable Blocks
-\autocite{zotero-79} and Glide \autocite{zotero-81} allow users to view
-data in a spreadsheet table, compute over the data using formulas, and
-connect the table to a GUI. Because many users are already familiar with
-using spreadsheets, this way of creating applications tends to be much
-easier than traditional software methods; for example, in a user study
-of Quilt, many people were able to create applications in under 10
-minutes, even if they expected it would take them many hours.
+Prior work has made the powerful realization that a spreadsheet can
+serve as an end-user-friendly backing data store and computation layer
+for an interactive web application. Research projects like Object
+Spreadsheets \autocite{mccutchen2016}, Quilt \autocite{benson2014},
+Gneiss \autocite{chang2014}, and Marmite \autocite{wong2007}, as well as
+commercial tools like Airtable Blocks \autocite{zotero-79} and Glide
+\autocite{zotero-81} allow users to view data in a spreadsheet table,
+compute over the data using formulas, and connect the table to a GUI.
+Because many users are already familiar with using spreadsheets, this
+way of creating applications tends to be far easier than traditional
+software methods; for example, in a user study of Quilt, many people
+were able to create applications in under 10 minutes, even if they
+expected it would take them many hours.
 
-Wildcard builds on this powerful idea, but applies it to modifying
-existing applications, rather than building new applications from
-scratch. For many people, we suspect that tweaking existing applications
-provides more motivation as a starting point for programming than
-creating a new application from scratch.
+Wildcard builds on this idea, but applies it to modifying existing
+applications, rather than building new applications from scratch. For
+many people, we suspect that tweaking existing applications provides
+more motivation as a starting point for programming than creating a new
+application from scratch.
 
 An important design decision for tools in this space is how to deviate
-from traditional spreadsheets. Quilt and Glide use existing web
-spreadsheet software as a backend, providing maximum familiarity and
-compatibility with existing spreadsheets. Gneiss has its own spreadsheet
-implementation, with additional features useful for building GUIs.
-Marmite provides a live data view that resembles a spreadsheet, but
-programming is actually done using a separate data flow pane rather than
-spreadsheet formulas. (Marmite's approach led to some confusion in a
-user study, because users expected behavior more similar to spreadsheets
-\autocite{wong2007}.) Airtable deviates the furthest from spreadsheets:
-although the user interface has formulas, shows live feedback and is
-easy to use, the underlying structure is a relational database with
+from traditional spreadsheets like Microsoft Excel or Google Sheets.
+Quilt and Glide use existing spreadsheet software as a backend,
+providing maximum familiarity for users, and even compatibility with
+existing spreadsheets. Gneiss has its own spreadsheet implementation
+with additional features useful for building GUIs. Marmite provides a
+live data view that resembles a spreadsheet, but programming is actually
+done using a separate data flow pane rather than spreadsheet formulas.
+(Marmite's approach led to some confusion in a user study, because users
+expected behavior more similar to spreadsheets \autocite{wong2007}.)
+Airtable deviates the furthest: although the user interface resembles a
+spreadsheet, the underlying structure is a relational database with
 typed columns. Wildcard's table is most similar to Airtable; the
@@
 
-\hypertarget{limitations-and-future-work}{%
-\section{Limitations and future
-work}\label{limitations-and-future-work}}
+\hypertarget{future-work}{%
+\section{Future work}\label{future-work}}
 
-Wildcard is still in early development; there are still many limitations
-to resolve and open questions to explore.
+Our prototype of Wildcard is still in early development; there are still
+many limitations to resolve and open questions to explore.
 
@@
   learn from the interface of Airtable which shows related objects in a
-  table cell, or SIEUFERD \autocite{bakke2016} which uses nested rows.
+  table cell, or add nested rows as used in other systems
+  \autocite{mccutchen2016,bakke2016}.
 \item
   Only allowing site adapters to be manually coded means that the number
-  of supported sites is limited. As mentioned previously, we may explore
-  also building wrappers using automated heuristics or end user scraping
-  tools.
+  of supported sites is limited. We plan to explore abstractions that
+  make it as easy as possible for programmers to efficiently create new
+  robust adapters, as well as integrating automated heuristics into the
+  adapter creation process.
 \end{itemize}
@@
 
-We live in a time of pessimism about the effects of the internet,
-especially given the degree of centralized control that large
-corporations wield. We think that one promising solution is to enable
-more people to have deeper control over the Web software they use every
-day. The Web offers a promising opportunity of an open foundation, but
-even the success of browser extensions has only empowered users within
-the limited confines of the ideas that programmers have decided to
-build.
+We live in a time of pessimism about the effects of the internet and the
+degree of centralized control wielded by large corporations. We think
+that one promising approach to this problem is to enable people to have
+deeper control over the Web software they use every day. The Web offers
+the potential of an open foundation, but even the success of browser
+extensions has only empowered users within the limited confines of the
+ideas that programmers have decided to build.
 

### 2020-01-09 paper.tex
updates

@@
 
-\title{Wildcard: End user modification of web applications using a live
-data table}
+\title{Wildcard: Spreadsheet-Driven Customization of Web Applications}
 \subtitle{}% optional
@@
 applications, but most people do not have the programming skills to
-implement their own customizations.
+implement their own extensions.
 
-In this paper, we present a prototype of Wildcard, an extension
-framework embedded in the browser that empowers users to casually tweak
-web applications without needing to program in Javascript or directly
-interact with the DOM.
+We present the idea of \emph{spreadsheet-driven customization}: enabling
+end users to customize existing applications using a live spreadsheet
+view of the data inside the application. By manipulating the
+spreadsheet, users can implement a wide variety of customizations,
+ranging from sorting lists of search results to displaying related data
+from other web services, without doing any traditional programming.
 
-Wildcard shows the data from a web page in a table, and maintains a live
-bidirectional connection between the table and the original page. By
-directly manipulating the table, users can perform a wide variety of
-modifications: sorting/filtering content, adding private annotations,
-using spreadsheet formulas to fetch data from other web services, adding
-custom UI elements to edit form data, and more.
+We built a prototype system called Wildcard that implements
+spreadsheet-driven customization as a web browser extension. Through
+concrete examples, we demonstrate that Wildcard has both a low barrier
+to entry for beginners and enough flexibility to solve many useful
+problems. We also show that Wildcard can work with real existing
+websites, by extracting structured data using web scraping techniques.
 
-We present examples of using Wildcard to solve real world problems, and
-explain the design principles behind the prototype. In the future, we
-envision growing Wildcard into a fully deployed system that helps make
-the web into a more malleable medium for all its users.
+For an online version of this paper with videos demonstrating the interface, see \url{https://www.geoffreylitt.com/wildcard/}
 \end{abstract}
@@
 motivated and skilled they might even be able to implement their own
-fix. But for most people in most situations, the only recourse is to
-complain to the developers and pray that someone listens---or more
-likely, to simply give up. While many have become used to this status
-quo, we see it as a tremendous waste of the openness of the Web
-platform. Back in 1977, in \emph{Personal Dynamic Media}
-\autocite{kay1977}, Alan Kay had originally envisioned personal
-computing as a medium that let a user ``mold and channel its power to
-his own needs,'' but today's software is more like concrete than clay.
+fix. But for most people, the only recourse is to complain to the
+developers and pray that someone listens---or more likely, to simply
+give up. While many have become used to this status quo, we see it as a
+tremendous waste of the openness of the Web platform. Back in 1977, in
+\emph{Personal Dynamic Media} \autocite{kay1977}, Alan Kay originally
+envisioned personal computing as a medium that let a user ``mold and
+channel its power to his own needs,'' but today software feels more like
+concrete than clay.
 
-In this paper, we introduce Wildcard, an extension framework embedded in
-the browser that lets users tweak web applications without programming
-in Javascript or interacting with the DOM. Wildcard shows a structured
-table view of the main data in a webpage, which maintains a live
-bidirectional connection to the original UI: when the user manipulates
-the table, the UI is instantly modified, and vice versa. This simple
-framework allows a wide range of possible modifications, ranging from
-sorting lists to fetching related data from other websites.
+In this paper, we present \emph{spreadsheet-driven customization}, a
+technique for making software customization more accessible to end
+users. The core idea is to show the data inside the application in a
+spreadsheet which maintains a live connection to the original UI. When
+the user manipulates the spreadsheet, the UI is instantly modified, and
+vice versa. We built a research prototype called Wildcard that
+implements spreadsheet-driven customization as a web browser extension,
+showing a spreadsheet panel in the context of existing sites.
 
-The system builds on two insights from prior work. The first insight is
-that a spreadsheet can serve as the backend for a GUI application, which
-enables an end user to build applications in an extremely accessible
-environment without doing traditional programming. The second insight is
-that web scraping---extracting structured data from an unstructured web
-page---unlocks value by making it possible to compute over that data.
-Our contribution in this work is to combine those two ideas in a novel
-way. By extracting structured data from an application and showing that
-data in a live spreadsheet table, Wildcard enables a wide class of end
-user modifications to existing applications. Wildcard makes it seem like
-an application is backed by a spreadsheet, even though the actual
-implementation is not. This allows the user to easily manipulate and
-compute over the application's underlying data just as they would in a
-spreadsheet, and then to immediately see the results of their changes in
-the original application's UI.
+Spreadsheets have proven to be a widely successful computing platform
+for non-programmers. They provide a low barrier to entry for beginners,
+while allowing experts to perform complex computations. The 2D grid is
+simple to grasp, but general enough to represent many types of data.
+Prior work \autocite{mccutchen2016,benson2014,chang2014} has used these
+benefits to enable end users to easily create web applications that use
+spreadsheets as a backing data layer.
 
-Wildcard embodies several principles we propose for designing systems
-that enable end user modification of existing software:
+Spreadsheet-driven customization applies these exact same benefits in a
+different context: customizing existing software, rather than building
+new software from scratch. We inherit the low floor and high ceiling of
+spreadsheets: small tweaks like sorting a list of data in an app can be
+done with a single click, while more complex customizations, like
+joining in related data from a web API, are supported by a rich formula
+system. The tabular data format is versatile, supporting data from a
+wide range of applications.
 
-\emph{Expose consistent structure}: Most applications only expose a user
-interface, without showing the structure of the data underneath.
-Furthermore, every application has a different interface, requiring
-users to learn many ways of using apps. Wildcard shows data from diverse
-applications in a single consistent structure, so that users are able to
-invest in learning one mechanism for manipulating the data in their
-applications, and for programming with that data.
+Spreadsheet-driven customization does not require that the application
+is actually backed by a spreadsheet---it merely presents a model of the
+internal state of the application, which can be exposed by the
+application itself or derived from the user interface by third parties.
+This means that spreadsheet-driven customization works with existing
+applications that people already use. It also allows for exposing a
+common data abstraction across applications, enabling users to have a
+consistent mechanism for manipulating data from all of their
+applications.
 
-\emph{Low floor, high ceiling}: We aim to provide a low barrier to entry
-so that non-programmers can casually use the system, but a high ceiling
-on the complexity it can achieve. Small but genuinely valuable tweaks
-like sorting a page are made possible with little effort in the flow of
-normal use to provide a low floor, and a rich formula system creates a
-high ceiling on possible modifications.
+Prior work \autocite{huynh2006} has shown that the process of extracting
+structured data from existing user interfaces can be confusing and
+unfamiliar for end users. In order to make the customization experience
+as straightforward as possible, we envision a tiered architecture that
+hides this complexity from end users. Wildcard provides a mechanism for
+programmers to write \emph{site adapters}, which use web scraping
+techniques to extract structured data from the DOM and AJAX requests of
+existing applications. End users only interact with the structured
+spreadsheet, providing a predictable experience with clear affordances
+for which customizations are possible.
 
-\emph{Build for multiple tiers of users}: We aim to make Wildcard as
-easy as possible for end users, by designing roles for tiers of users
-with differing levels of technical sophistication. Programmers or
-tech-savvy end users are responsible for the data extraction step of
-mapping existing UIs to structured data. Most end users only interact
-with the structured data, providing a predictable and user-friendly
-experience.
-
-Wildcard is currently an early research prototype, with incomplete
-features and limited coverage of sites. We plan to continue privately
-testing the system with our own use cases, and then to release the tool
-publicly, enabling other programmers to contribute site adapters and
-other plugins. Ultimately, we hope that Wildcard makes modifying
-websites into a natural part of using the web, as ubiquitous as using
-spreadsheets in everyday computing.
+Wildcard is currently an early prototype, with incomplete features and
+limited coverage of sites. We plan to continue building site adapters
+and testing the system with our own use cases, to better understand how
+well the spreadsheet abstraction maps to real websites and customization
+needs. Eventually we also plan to release the tool publicly, to
+determine how end users choose to use spreadsheet-driven customization,
+what usability challenges emerge, and how feasible it is for programmers
+to build and maintain site adapters.
 
@@
 For extracting data from the page and getting it into structured form,
-Wildcard provides ways to concisely and declaratively express scraping
-logic. For example, here is a code snippet for extracting the name of an
-Airbnb listing. It specifies the field name and some metadata
-properties, and then a function that, given the entire listing, returns
-the DOM element representing the name of the listing.
+Wildcard provides ways to concisely express web scraping logic. For
+example, here is a code snippet for extracting the name of an Airbnb
+listing:
 
@@
   \OperatorTok{\{}
-\NormalTok{    fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,}
-\NormalTok{    readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}
-\NormalTok{    type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}
+\NormalTok{    fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,} \CommentTok{// The name of the data field}
+\NormalTok{    readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}    \CommentTok{// Whether the user can edit the field}
+\NormalTok{    type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}      \CommentTok{// The type of the field}
+    \CommentTok{// Given the DOM element for the entire listing,}
+    \CommentTok{// return the DOM element representing this field}
 \NormalTok{    el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
@@
 
+Scraping data out of the UI has the advantage of working for any data
+visible on the page, but it can require a lot of work to convert
+unstructured UI to structured data, and sometimes important data is not
+shown in the UI. To work around these problems, we have also prototyped
+mechanisms for site adapters to observe AJAX requests made by the
+browser and extract data directly from JSON responses. (This mechanism
+was used in the Airbnb Walkscore example: approximate latitude and
+longitude aren't shown in the Airbnb UI, but they are available in AJAX
+responses.)
+
 The site adapter also needs to support the reverse direction: sending
@@
 find the divs representing data rows, and the core platform mutates the
-DOM to reflect the table state. The only exception is site-specific
-actions (like favoriting in Airbnb): actions are implemented as regular
-Javascript functions running in the context of the page, which can
-mutate the DOM, simulate clicks on buttons in the UI, launch AJAX
-requests, navigate to new pages, etc.
+DOM to reflect the table state. The only exception is row actions (like
+favoriting an Airbnb listing), which are implemented as regular
+Javascript functions running in the context of the page and can mutate
+the DOM, simulate clicks on buttons, etc.
 
@@
 
-The design of Wildcard is grounded in several principles, informed by
-prior work and our own experimentation. We think these principles extend
-beyond Wildcard and might inform the design of other end user
-programming tools, particularly systems enabling users to modify GUI
-apps.
+The idea of spreadsheet-driven customization is guided by several design
+principles, inspired by prior work and our own experimentation. We think
+these principles can also broadly inform the design of end user
+programming tools, especially those that enable users to customize
+existing software.
 
-\hypertarget{expose-consistent-structure}{%
-\subsection{Expose consistent
-structure}\label{expose-consistent-structure}}
+\hypertarget{expose-a-universal-data-structure}{%
+\subsection{Expose a universal data
+structure}\label{expose-a-universal-data-structure}}
 
-Today, most personal computing consists of using distinct apps, each one
-an isolated silo of data and functionality. While there are some shared
-idioms and limited points of interoperability, apps fundamentally
-operate independently from one another. Computing does not need to
-operate this way---for example, UNIX offers a compelling alternate
-vision: many small tools, all built on the consistent abstraction of
-text files. Users can invest in deeply mastering a text editor, and then
-use it for a vast array of tasks, even as an interactive input mechanism
-in shell programs (e.g.~for editing git commit messages).
-Beaudouin-Lafon and Mackay call this style of interaction
-\emph{polymorphic} \autocite{beaudouin-lafon2000}, since a single
-interface widget can be used in many contexts. diSessa
-\autocite{disessa2000} has also noted the deep connection between
-polymorphism and literacy: textual literacy rests on a single rich
-medium of writing, which once mastered, can be adapted to many different
-genres and uses. If people needed to relearn from scratch when switching
-from writing essays to writing emails, the medium would lose most of its
-potency.
+Today, most personal computing consists of using applications, which
+bundle together behavior and data to provide some set of functionality.
+While there are limited points of interoperability, applications
+generally are designed to operate independently of one another. Once a
+user has gotten past learning the basic idioms of modern computing like
+windows and scrollbars, most of the effort invested in learning to use
+one application does not carry over to using other applications.
 
-With Wildcard, we hope to make isolated Web applications work more like
-UNIX, by imposing a consistent and simple data structure onto many
-applications. Just as UNIX uses text files to represent all data,
-Wildcard uses relational tables. This abstraction strikes a balance
-between being simple and generic: a data table is much simpler than the
-DOM tree that describes all the details of the UI, but is also generic
-enough to describe the essence of many different interfaces.
+Computing does not need to be organized this way. For example, UNIX
+offers a compelling alternative design: many small single-purpose
+utilities, all of which manipulate a universal format of text streams.
+The universal format creates a high degree of leverage from tools: users
+can get a lot of utility from deeply mastering a text editor and some
+text manipulation utilities, because these tools can be applied to
+nearly any task. As just one example, a user's preferred text editor can
+even serve as an interactive input mechanism in shell programs, e.g.~for
+editing git commit messages.
 
-\begin{figure}
-\centering
-\includegraphics{media/structure.png}
-\caption{Wildcard layers the consistent structure of a data table over
-different isolated applications.}
-\end{figure}
+Spreadsheet-driven customization aims to port this UNIX philosophy to
+the world of isolated Web applications, by creating a consistent data
+structure to represent the data inside many applications. In UNIX, the
+universal format is a text stream; in Wildcard, it is a relational
+table: a simple abstraction generic enough to describe the data used in
+many different applications. Because Wildcard maps the data from all
+applications to the table format, users can invest in mastering the
+Wildcard table editor, the formula language, and cell editor UIs, and
+reuse those same tools to customize many different applications.
 
-Exposing structure gives users more choice When UI widgets can compete
-on their merits rather than having a monopoly on a given data silo or
-application, it creates stronger competition at the interface level. For
-example, there is competition among email clients which consume an open
-protocol, but not among Facebook or Twitter clients. Exposing structure
-also leads to improved UI consistency across apps: UNIX users can reuse
-their text editor everywhere, but Web users must use a different rich
-text editor for each site. Tim Berners-Lee has written in more detail
-about these benefits \autocite{berners-lee2018} in the context of the
-SOLID project, which envisions giving users control of their own data as
-a mechanism for decoupling data from interfaces (e.g., giving users a
-choice of which client to use to consume a given social media feed).
-Wildcard has overlapping goals with SOLID, but does not require
-decentralized user control of data---the data can remain on a
-centralized server, as long as the interface can be tweaked by end
-users.
+This idea relates to Beaudouin-Lafon and Mackay's notion of
+\emph{polymorphic interface instruments} \autocite{beaudouin-lafon2000}:
+UI elements that can be used in different contexts, like a color picker
+that can be used in many different drawing applications. diSessa has
+also noted the connection between literacy and the genericness of a
+medium. Textual literacy rests on the fact that writing can be adapted
+to many different genres and uses \autocite{disessa2000}; if people
+needed to relearn reading and writing from scratch when switching from
+essays to emails, the medium would lose most of its potency. We think
+providing generic tools is especially important for software
+customization, because the most common barrier to customizing software
+is not having enough time \autocite{mackay1991}---it's more likely that
+people will customize software regularly if they can reuse the same tool
+across many applications.
 
-\emph{todo: move the sloppy contrast, and SOLID, down into related work}
+This design principle leads to several challenges. First, any universal
+abstraction has its constraints, and can't necessarily naturally express
+the data in every application. We plan to explore the limits of the
+table abstraction further, by trying to build adapters for more sites
+with varied data formats. We expect that many types of data can fit
+fairly naturally into tables: lists of search results, news articles,
+and messages can all be seen as relations. On sites that use document
+structures (e.g.~Google Docs) or graph structures (e.g.~social friend
+graphs), it may prove more challenging to map internal data to this
+abstraction.
 
-Layering a structured view on top of an existing application is not the
-only option available. ``Sloppy programming'' systems like Chickenfoot
-\autocite{bolin2005} and CoScripter \autocite{leshed2008} forego the
-intermediate layer of structure, instead allowing users to create
-scripts in an informal language and then perform fuzzy pattern matching
-to find elements in the DOM. For example, to find a button after a
-textbox in Chickenfoot, the user could type
-\texttt{click(find(“button\ just\ after\ textbox”))}. These designs
-allow for expressing a wide range of operations, but they don't
-explicitly indicate what operations are possible---the user can only see
-the original page and imagine the possibilities. In contrast, Wildcard
-provides affordances that clearly suggest the availability of certain
-actions (e.g.~sorting, editing a cell, adding a column with a derived
-value), especially to users who are familiar with spreadsheets. In
-addition to giving users more certainty about whether a modification is
-possible, these affordances might improve discoverability and give users
-new ideas for things to try. ``Sloppy programming'' does give end-users
-greater access to modify the apps that they use, but does not seem to
-provide them with a deeper understanding of those apps or how they
-relate to each other, so the apps are more likely to remain siloed.
-
-One key challenge of making a structured view successful is ensuring a
-close mapping in the user's mind between the new representation and the
-original page. Wildcard provides live visual cues as the user navigates
-the data table (similar to the highlighting provided by DOM inspectors
-in browser developer tools). In our own usage, we have found that this
-live highlighting makes it very clear how the two representations map to
-each other.
+Another challenge is ensuring a clear mapping in the user's mind between
+the spreadsheet and the original page. Wildcard provides live visual
+cues as the user navigates the data table (similar to the highlighting
+provided by DOM inspectors in browser developer tools). In our own
+usage, we have found that this live highlighting makes it very clear how
+the two representations map to each other.
 
@@
 
-Seymour Papert advocated for programming systems with a ``low floor''
-and a ``high ceiling''---making it easy for novices to get started, but
-also providing possible a large range of possibilities for more
-sophisticated users \autocite{resnick2016}. Unfortunately, many
-programming systems only provide either one or the other
-\autocite{myers2000}.
+Seymour Papert advocated for programming systems with a ``low floor,''
+making it easy for novices to get started, as well as a ``high
+ceiling,'' providing a large range of possibilities for more
+sophisticated users \autocite{resnick2016}. Our goal is for
+spreadsheet-driven customization to meet both of these criteria.
+Although we need more testing with real users to know for sure, we think
+Wildcard clearly provides a low floor, and likely has a flexible enough
+foundation to eventually provide a high ceiling.
 
-\begin{itemize}
-\tightlist
-\item
-  tweaking vs creation from scratch
-\item
-  in-place toolchain
-\item
-  easy things easy
-\item
-  easy things valuable
-\item
-  this seems clearly justified by the demos
-\end{itemize}
+One of the most interesting properties of spreadsheets is the amount of
+value they can provide to users who are aware of only a tiny sliver of
+their functionality. Not only can a novice perform some basic activities
+in a spreadsheet with almost no training (e.g., storing tables of
+numbers or computing simple sums), these activities are often actually
+valuable for the user! The fact that useful tasks can be performed early
+on supports the user's natural motivation to continue using the tool,
+and to eventually learn its more powerful features if needed
+\autocite{nardi1991}. In contrast, many traditional programming systems
+require an enormous upfront investment of time and practice before
+someone is able to write a program that actually helps them achieve a
+real task.
 
-rewrite\ldots{}
+As part of ensuring a low floor for spreadsheet-driven customization, we
+have focused on including genuinely valuable features for novices. For
+example, a user can sort a table with a single click, or simply type in
+some annotations. We would expect many Wildcard users to start out using
+these simpler features before potentially moving on to more
+sophisticated features like formulas.
+
+Another aspect of providing a low floor is providing an ``in-place
+toolchain'' \autocite{inkandswitch2019}---minimizing the effort of
+moving from using to customizing, by making customization tools
+available in the same environment where the user is already using the
+software. This quality is distinct from the level of technical skill
+needed to use the tool: for example, setting up a workflow trigger in an
+end user programming system like \href{https://ifttt.com/}{IFTTT} does
+not require much technical skill, but does require leaving the user's
+normal software and entering a separate environment; conversely, running
+a Javascript snippet in the browser console requires programming skills,
+but can be done immediately and casually in the flow of using a website.
 
@@
 \emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
-\emph{Requires programming} & browser dev tools & editing open source
+\emph{Requires programming} & browser JS console & forking open source
 software\tabularnewline
@@
 
-We can evaluate a system for tweaking software along two dimensions.
-First, the technical capability required of the user: is programming
-knowledge needed? Second, whether the change is ``in-place''
-\autocite{inkandswitch2019}: can it be made within the same environment
-that the software is typically used in? These dimensions are not
-orthogonal, but they are distinct. For example, setting up a workflow
-trigger in an end user programming system like
-\href{https://ifttt.com/}{IFTTT} does not require much technical skill,
-but it does require leaving the user's normal software and entering a
-separate environment. On the other hand, running a Javascript snippet in
-the browser console requires programming skills, but can be done
-immediately and casually in the flow of using a website.
+Wildcard provides an in-place toolchain---on any site that supports
+Wildcard, the user is one click away from starting to tweak the site
+with minimal friction. There is no need for them to switch into another
+environment or figure out how to extract data from the page. Once the
+user starts editing, Wildcard also provides live feedback to help users
+understand the changes they are making. Even if a user isn't yet totally
+familiar with Wildcard, they can learn to use the system by trying
+things out and seeing what happens.
 
-In addition to requiring no programming skills, Wildcard supports
-frequent, small modifications. The Wildcard table appears in the course
-of normal web browsing, to ensure that the tools for modification are
-close at hand while using the original software.Wildcard also provides
-live feedback, immediately editing the page in response to user
-interaction, which gives users confidence in the changes they are
-making.
-
-Spreadsheets can be used in genuinely useful ways even by someone who
-has learned only a tiny part of their functionality, e.g.~for storing
-tables of numbers or computing simple sums \autocite{nardi1991}. This
-creates inherent motivation to continue using the tool and eventually
-learn its more powerful features. In contrast, many traditional
-programming systems require an enormous time investment before a
-beginner is able to create any valuable software. Like spreadsheets,
-Wildcard aims to support genuinely useful actions with little effort:
-for example, sorting a table with a single click, or adding an
-annotation by typing into a cell. We would expect many Wildcard users to
-start by using these bits of functionality, before possibly moving on to
-more complex features like formulas.
-
-High ceiling?
+Since we have still only built several site adapters and demos, it's
+still too early to tell how high the ceiling is for the customizations
+that can be achieved with Wildcard, but we think that with enough
+operators the formula language could eventually support a wide variety
+of customizations. We plan to explore this aspect further by trying to
+solve more real problems with the system and observing where limitations
+emerge in practice.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  make things easy for the last tier of users (low floor)
-\item
-  expose predictable structure, don't make them participate in data
-  extraction
-\end{itemize}
+Real-world spreadsheet usage in offices is highly collaborative: most
+users just perform simple changes, while a few coworkers help with
+writing more complex formulas or even programming macros
+\autocite{nardi1990}. Inspired by this, we aim to make
+spreadsheet-driven customization a collaborative activity that combines
+the different abilities of many users in a collaborative ecosystem.
 
-Some end user programming systems are designed with a single user in
-mind---a lone non-programmer, needing to achieve a task through some
-combination of individual skills and a powerful tool. But in fact,
-real-world usage of end user programming systems is far more
-collaborative. Spreadsheets are built together by users with different
-levels of technical expertise: some users just write simple formulas,
-while their officemates help write more complex formulas or even program
-macros in traditional languages \autocite{nardi1990}. Our goal with
-Wildcard is to consider the full ecosystem of users and to design a tool
-that best combines all of their abilities.
-
-This design principle is most salient in our approach to building site
-adapters. Many web augmentation projects aim to make the full process of
-tweaking available to end users with no programmer help needed, through
-a combination of automated heuristics and interactive feedback. This is
-a laudable goal because it makes these tools work with any site, and
-these approaches are often quite successful, but they also inevitably
-break down sometimes \autocite{bolin2005} \autocite{little2010}. In
-contrast, human programmers working together have been able to maintain
-stable scraping libraries, even for complex sites like Gmail
-\autocite{streak,talwar2019}. Our primary approach to building site
-adapters is having programmers manually create them, because we are
-willing to trade off lower site coverage in return for more reliable
-results on the sites that are covered, building confidence for users.
-Like the collaboration between an expert and novice spreadsheet user,
-this architecture leverages more skilled programmers while still
-providing end users with a flexible programming environment to meet
-their own needs. On the other hand, we remain very open to incorporating
-other tools in the future that enable end users to also create their own
+The main way we do this is by separating website customization into two
+separate stages: structured data extraction, and using the resulting
+spreadsheet. The first stage is currently only available to programmers
+who can code site adapters in Javascript, whereas the second stage is
+available for any non-programmer end user. This architecture frees end
+users from needing to think about data extraction, and enables a
+community of end users to reuse the efforts of programmers building site
 adapters.
 
-\begin{figure}
-\centering
-\includegraphics{media/adapters.png}
-\caption{Having third party programmers create site adapters balances
-site coverage and adapter quality.}
-\end{figure}
+The group of users building adapters does not necessarily need to be
+limited only to programmers. In the future we might explore enabling end
+users to also create site adapters, drawing on related work on enabling
+end users to extract structured data from websites
+\autocite{chasins2018,huynh2006}. But even in that case, there would
+still be a separation between motivated, tech-savvy end users building
+adapters and more casual end users just using the spreadsheet view.
 
-In any system for tweaking software, another important consideration is
-how to involve the first party developers of the original software. Our
-approach is to avoid depending on their cooperation, but also to aim for
-eventual cooperation.
-
-This approach takes advantage of the unusual openness of the Web
-platform. On many other platforms (e.g.~smartphone operating systems),
-software is locked down unless first-party developers explicitly provide
-hooks for plugins and interoperability, but on the Web, all client-side
-code is available for browser extensions to modify. Application authors
-can use practices that make it easier to modify their apps (e.g.~clean
-semantic markup), or more difficult (e.g.~code obfuscation), but the
-default state is openness. This gives extensions freedom to modify
-applications in creative ways that the original developers did not plan
-for. At the most extreme, this can even include modifications that are
-hostile to the desires of the original developers. Although ad blocking
-extensions make this style of extension very prominent, most ideas for
-extending software merely fall outside the range of possibilities
-originally imagined by the developer, rather than being actively
-hostile.
-
-Wildcard takes advantage of this openness, and does not depend on
-cooperation from first-party website developers: any programmer can add
-support for any website to Wildcard by building a third party adapter.
-But the platform is still open to first parties developing their own
-adapters, which would often be in their interests. Providing Wildcard
-support would allow users to build extensions to fulfill their own
-feature requests. It also would not necessarily require much effort:
-adding Wildcard support would be more straightforward for a first-party
-than a third-party because they have direct access to the structured
-data in the page. There is also precedent for first parties implementing
-an official client extension API in response to user demand: for several
-years, Google maintained an official extension API in Gmail for
-Greasemonkey scripts to use. (Incidentally, since then, third parties
-have continued to maintain stable Gmail extension APIs used by many
-browser extensions \autocite{streak,talwar2019}, illustrating the
-potential of collaboratively maintaining third party adapters.)
+Another stakeholder to consider is the first party developers of the
+original software. Spreadsheet-driven customization does not depend on
+cooperation from first-party website developers: a third party
+programmer can write an adapter for any website, which can access any
+information available in the browser. On the other hand, if first
+parties were to expose structured data from their applications, it would
+avoid the need for adapters and generally make customization a lot
+easier. We think there are compelling reasons for first parties to
+consider doing this. Providing Wildcard support would allow users to
+build extensions to fulfill their own feature requests. It also would
+not necessarily require much effort: adding Wildcard support would be
+more straightforward for a first-party than a third-party because they
+have direct access to the structured data in the page. There is also
+precedent for first parties implementing an official client extension
+API in response to user demand: for several years, Google maintained an
+official extension API in Gmail for Greasemonkey scripts to
+use.\footnote{Incidentally, since then, third parties have continued to
+  maintain stable Gmail extension APIs used by many browser extensions
+  \autocite{streak,talwar2019}, illustrating the potential of
+  collaboratively maintaining third party adapters.}
 
@@
 
-\hypertarget{web-augmentation}{%
-\subsection{Web augmentation}\label{web-augmentation}}
+\hypertarget{web-customization}{%
+\subsection{Web customization}\label{web-customization}}
 
 Wildcard's goals are closely shared with other systems that provide
-interfaces in the browser for end users to augment and modify websites
-while using them.
+interfaces in the browser for end users to augment and customize
+websites while using them.
 
@@
 formulas, these downsides can be mitigated.
+
+\emph{Copy of the sloppy contrast that used to be earlier:}
+
+Layering a structured view on top of an existing application is not the
+only option available. ``Sloppy programming'' systems like Chickenfoot
+\autocite{bolin2005} and CoScripter \autocite{leshed2008} forego the
+intermediate layer of structure, instead allowing users to create
+scripts in an informal language and then perform fuzzy pattern matching
+to find elements in the DOM. For example, to find a button after a
+textbox in Chickenfoot, the user could type
+\texttt{click(find(“button\ just\ after\ textbox”))}. These designs
+allow for expressing a wide range of operations, but they don't
+explicitly indicate what operations are possible---the user can only see
+the original page and imagine the possibilities. In contrast, Wildcard
+provides affordances that clearly suggest the availability of certain
+actions (e.g.~sorting, editing a cell, adding a column with a derived
+value), especially to users who are familiar with spreadsheets. In
+addition to giving users more certainty about whether a modification is
+possible, these affordances might improve discoverability and give users
+new ideas for things to try. ``Sloppy programming'' does give end-users
+greater access to modify the apps that they use, but does not seem to
+provide them with a deeper understanding of those apps or how they
+relate to each other, so the apps are more likely to remain siloed.
 

### 2020-01-10 paper.tex
pdf diagram of airbnb

@@
 system. The tabular data format is versatile, supporting data from a
-wide range of applications.
-
-Spreadsheet-driven customization does not require that the application
-is actually backed by a spreadsheet---it merely presents a model of the
-internal state of the application, which can be exposed by the
-application itself or derived from the user interface by third parties.
-This means that spreadsheet-driven customization works with existing
-applications that people already use. It also allows for exposing a
-common data abstraction across applications, enabling users to have a
-consistent mechanism for manipulating data from all of their
-applications.
+wide range of applications. Spreadsheet-driven customization does not
+require that the application is actually backed by a spreadsheet---it
+merely presents a model of the internal state of the application, which
+can be exposed by the application itself or derived from the user
+interface by third parties. This means that spreadsheet-driven
+customization works with existing applications that people already use.
+It also allows for exposing a common data abstraction across
+applications, enabling users to have a consistent mechanism for
+manipulating data from all of their applications.
 
@@
 example of someone using it to help with booking a trip on the travel
-sites Airbnb and Expedia.
+sites Airbnb and Expedia. These demos are best viewed in the online
+version of this paper, which includes videos:
+\url{https://geoffreylitt.com/wildcard}
+
+\begin{figure}
+\hypertarget{fig:airbnb-demo}{%
+\centering
+\includegraphics{media/airbnb-demo-300dpi.png}
+\caption{Using Wildcard to augment the Airbnb page for searching travel
+accommodations}\label{fig:airbnb-demo}
+}
+\end{figure}
 
@@
 rating. Using Wildcard, the user can add these small features, while
-leaving the page's design and the rest of its functionality unchanged.
+leaving the page's design and the rest of its functionality unchanged
+(shown in fig.~\ref{fig:sorting}).
 
-First, the user opens up the Wildcard panel, which shows a table
-corresponding to the search results in the page. As they click around in
-the table, the corresponding row in the page is highlighted so they can
-see the connection between the views.
-
-Then, the user can use standard spreadsheet column header features to
-sort the page by price and filter by rating:
-
-Notice how after manipulating the data, the user was able to close the
-table view and continue using the website with its original visual
-design. The table view offers a way to change the data backing a page,
-but does not need to replace the original interface entirely.
+After manipulating the data, the user can close the table view and
+continue using the website with its original visual design. The table
+view offers a way to change the data backing a page, but does not need
+to replace the original interface entirely.
 
@@
 taken on a row in the table, like adding an item to a shopping cart.
-Wildcard has the ability to make these actions available in the data
-table if the site adapter implements them. The main advantage this
-provides is the ability to easily perform an action in bulk across
+Wildcard has the ability to make these ``row actions'' available in the
+data table if the site adapter implements them. The main advantage this
+provides is the ability to easily perform a row action in bulk across
 multiple rows.
@@
 For example, it's tedious on Airbnb to click on listings one by one to
-add them to a list of favorites. Using Wildcard, we can just select
-multiple rows and favorite all of them with one click. Similarly, we can
-also open the detailed pages for many listings in new tabs.
+add them to a list of favorites. Using Wildcard, the user can just
+select multiple rows and favorite all of them with one click. Similarly,
+we can also open the detailed pages for many listings in new tabs.
+(Shown in fig.~\ref{fig:actions})
 
-Now the user would like to jot down some notes on the pros and cons of
-each listing. To do this, they can simply type notes into an additional
-column next to each listing, and the notes appear inside the listings in
-the original UI. These annotations are also persisted in the browser for
-future retrieval.
-
-Wildcard also includes a formula language which enables more
-sophisticated tweaks that fetch external data and perform computations.
-
-When traveling without a car, it's nice to evaluate potential places to
-stay based on how walkable the surroundings. Using Wildcard formulas, we
-can integrate Airbnb with Walkscore, an API that rates the walkability
-of any location on a 1-100 scale. When we call the \texttt{WALKSCORE}
-formula with the latitude and longitude of the listing, it returns the
-walk score as the cell value. Because the cell's contents are injected
-into the page, the score also shows up in the page body.
-
-It might seem that Wildcard is only useful on websites that display
-lists of tabular data like search results. But in fact, the table
-metaphor is flexible enough to represent many types of data. For
-example, a form can be represented as a single row, with a column for
-each input.
-
-In previous examples the data extracted from the site was marked as
-read-only; users cannot change the name or price of an Airbnb listing.
-In this next case, the cells are marked as writable, so that changes in
-the table are reflected in the form inputs. This becomes useful when
-combined with GUI widgets for editing the value of a table cell.
-
-Filling in dates for a flight search typically requires opening up a
-separate calendar app to find the right dates, and then manually copying
-them into the form. In Wildcard, we can make this easier by providing a
-datepicker widget that has privileged access to the user's calendar
-information.
+\begin{figure}
+\hypertarget{fig:expedia-demo}{%
+\centering
+\includegraphics{media/expedia-demo-300dpi.png}
+\caption{Using Wildcard to augment the Expedia page for booking a
+flight}\label{fig:expedia-demo}
+}
+\end{figure}
 
@@
 \begin{figure}
+\hypertarget{fig:architecture}{%
 \centering
 \includegraphics{media/architecture-clean.png}
-\caption{The architecture of the Wildcard system}
+\caption{The architecture of the Wildcard
+system}\label{fig:architecture}
+}
 \end{figure}
@@
 
-Scraping data out of the UI has the advantage of working for any data
-visible on the page, but it can require a lot of work to convert
-unstructured UI to structured data, and sometimes important data is not
-shown in the UI. To work around these problems, we have also prototyped
-mechanisms for site adapters to observe AJAX requests made by the
-browser and extract data directly from JSON responses. (This mechanism
-was used in the Airbnb Walkscore example: approximate latitude and
-longitude aren't shown in the Airbnb UI, but they are available in AJAX
-responses.)
+Sometimes important data is not shown in the UI, making it impossible to
+scrape from the DOM. To address this, we have also prototyped mechanisms
+for site adapters to observe AJAX requests made by the browser and
+extract data directly from JSON responses. This mechanism was used to
+implement the Airbnb Walkscore example, since latitude and longitude
+aren't shown in the Airbnb UI, but they are available in AJAX responses.
+We also plan to add the ability for site adapters to scrape data across
+multiple pages in paginated lists of results (as explored in prior work
+\autocite{huynh2006}).
 
@@
 updates from the table to the original page. Most DOM manipulation is
-not performed directly by the site adapter: the adapter specifies how to
-find the divs representing data rows, and the core platform mutates the
-DOM to reflect the table state. The only exception is row actions (like
-favoriting an Airbnb listing), which are implemented as regular
-Javascript functions running in the context of the page and can mutate
-the DOM, simulate clicks on buttons, etc.
+not performed directly by the site adapter; instead, the adapter
+specifies how to find the divs representing data rows, and the core
+platform mutates the DOM to reflect the table state. The only exception
+is row actions (like favoriting an Airbnb listing), which are
+implemented as imperative Javascript functions that can can mutate the
+DOM, simulate clicks on buttons, etc.
 
@@
 
+\hypertarget{malleable-software}{%
+\subsection{Malleable software}\label{malleable-software}}
+
 In the broadest sense, Wildcard is inspired by systems aiming to make
@@
 applications built by programmers. These systems include Smalltalk
-\autocite{kay1977}, Hypercard \autocite{hypercard2019}, Boxer
+\autocite{kay1977}, Hypercard \autocite{hypercard2019} , Boxer
 \autocite{disessa1986}, Webstrates \autocite{klokmose2015}, and
-Dynamicland \autocite{victor}. (The project's name Wildcard comes from
-the internal pre-release name for Hypercard, which doubly inspired our
-work by promoting both software modification by end users and the ideas
-behind the Web.)
+Dynamicland \autocite{victor}.\footnote{The project's name Wildcard
+  comes from the internal pre-release name for Hypercard, which doubly
+  inspired our work by promoting both software modification by end users
+  and the ideas behind the Web.}
 
@@
 maximize the malleability of already existing software. This approach
-has the pragmatic benefit of immediately being useful in a much broader
-context, although it also requires working within rigid constraints. We
-also see Wildcard growing beyond its current bounds and becoming more
-similar in the future to these other systems. With substantial future
-work, Wildcard could grow from merely being a platform for tweaking
-existing software into a platform for building new software from
-scratch, designed from the ground up for end user tweakability.
+has the pragmatic benefit of being immediately useful in more
+situations, although it also requires working within more rigid
+constraints.
 
-More narrowly, Wildcard builds on three areas of existing work: web
-augmentation, spreadsheet-based app development, and web scraping. Our
-contribution in this work is to combine these areas in a new way:
-extracting structured data from a website and directly exposing it to
-the user as a live data table, with the end goal of modifying an
-existing application. We think this combination builds on valuable ideas
-and learnings from each of the areas, while also overcoming some of the
-downsides of each.
+With substantial future work, Wildcard could become more similar to
+these other projects, growing from a platform for tweaking existing
+software into a platform for building new software from scratch. This
+would likely end up resembling existing tools for building
+spreadsheet-driven applications (discussed more below), but with an
+extra focus on customizability by end users of the software.
 
@@
 changes to the heuristics can also cause problems), and it is not easy
-for users to discover the space of possible commands.
+for users to discover the space of possible commands
+\autocite{little2010}.
 
@@
 formulas, these downsides can be mitigated.
-
-\emph{Copy of the sloppy contrast that used to be earlier:}
-
-Layering a structured view on top of an existing application is not the
-only option available. ``Sloppy programming'' systems like Chickenfoot
-\autocite{bolin2005} and CoScripter \autocite{leshed2008} forego the
-intermediate layer of structure, instead allowing users to create
-scripts in an informal language and then perform fuzzy pattern matching
-to find elements in the DOM. For example, to find a button after a
-textbox in Chickenfoot, the user could type
-\texttt{click(find(“button\ just\ after\ textbox”))}. These designs
-allow for expressing a wide range of operations, but they don't
-explicitly indicate what operations are possible---the user can only see
-the original page and imagine the possibilities. In contrast, Wildcard
-provides affordances that clearly suggest the availability of certain
-actions (e.g.~sorting, editing a cell, adding a column with a derived
-value), especially to users who are familiar with spreadsheets. In
-addition to giving users more certainty about whether a modification is
-possible, these affordances might improve discoverability and give users
-new ideas for things to try. ``Sloppy programming'' does give end-users
-greater access to modify the apps that they use, but does not seem to
-provide them with a deeper understanding of those apps or how they
-relate to each other, so the apps are more likely to remain siloed.
 
@@
 
-Our prototype of Wildcard is still in early development; there are still
-many limitations to resolve and open questions to explore.
+There are still many open questions about spreadsheet-driven
+customization, which we hope to answer through targeted development and
+usage of the Wildcard prototype.
 
-The most important question is whether the combination of features
-provided by Wildcard can be combined in enough useful ways to make the
-system worth using in practice. While initial demos are promising, we
-need to develop more site adapters and use cases to more fully assess
-this question. We plan to continue privately testing the system with our
-own needs, and to eventually deploy the tool publicly, once the API is
-stable enough and can support a critical mass of sites and use cases. We
-also plan to run usability studies to evaluate and improve the design of
-the tool.
+The most important question is whether the computational model can
+support a large variety of useful customizations. While initial demos
+are promising, we need to develop more site adapters and use cases to
+explore this further. We plan to continue privately testing the system
+with our own needs, and to eventually deploy the tool publicly, once the
+API is stable enough and can support a critical mass of sites and use
+cases. We also plan to run usability studies to evaluate and improve the
+design of the tool.
 
-Here are a few of the largest limitations in the current system:
-
-\begin{itemize}
-\tightlist
-\item
-  Wildcard only extracts data visible on the page, which means that
-  subsequent pages of results in paginated lists are not included in the
-  table. (Sifter \autocite{huynh2006} uses techniques that might help
-  get around this.)
-\item
-  There is no mechanism for end users to express imperative workflows
-  (e.g.~``click this button, and then\ldots{}''); they can only write
-  formulas that return data and then inject the resulting data into the
-  page. While this makes the system simpler, it also may exclude many
-  valuable use cases. We may add a system for expressing workflows like
-  this, although it's not obvious how it would fit together with the
-  existing table view. Gneiss \autocite{chang2014} contains some
-  mechanisms for writing spreadsheet formulas which can handle behaviors
-  like reloading data in response to a user pressing a button, which
-  might prove helpful.
-\item
-  Wildcard's data model only shows a single table at a time, without any
-  notion of relationships between tables. A richer data model with
-  foreign keys might be necessary to support certain use cases. For
+We suspect that two areas of the current model may prove too limiting.
+First, Wildcard's data model only shows a single table at a time,
+without any notion of relationships between tables. A richer data model
+with foreign keys might be necessary to support certain use cases. For
 designing a tabular interface on top of a richer data model, we could
@@
 \autocite{mccutchen2016,bakke2016}.
-\item
-  Only allowing site adapters to be manually coded means that the number
-  of supported sites is limited. We plan to explore abstractions that
-  make it as easy as possible for programmers to efficiently create new
-  robust adapters, as well as integrating automated heuristics into the
-  adapter creation process.
-\end{itemize}
+
+Second, there is no mechanism for end users to express imperative
+workflows (e.g.~``click this button, and then\ldots{}''); they can only
+write formulas that return data and then inject the resulting data into
+the page. If it proves useful, we might consider adding a system for
+expressing workflows like this, although it would substantially alter
+the computational model and it's not clear how it would fit together
+with the existing table view.
+
+Another open question is how efficiently site adapters can be created
+for real websites. By making adapters for sites with different types of
+data in different domains, we plan to explore how automated heuristics
+can speed up the adapter creation process, and to create abstractions
+that make it easier for programmers to efficiently create new robust
+adapters.
 
@@
 
-We live in a time of pessimism about the effects of the internet and the
-degree of centralized control wielded by large corporations. We think
-that one promising approach to this problem is to enable people to have
-deeper control over the Web software they use every day. The Web offers
-the potential of an open foundation, but even the success of browser
-extensions has only empowered users within the limited confines of the
-ideas that programmers have decided to build.
+In this paper, we have presented \emph{spreadsheet-driven
+customization}, a way to make software customization more accessible to
+end users by showing the data inside the application in a spreadsheet.
+We hope that this technique contributes to making the Web into a more
+dynamic medium that users can mold to their own needs.
 
-In this paper, we have presented Wildcard, a browser-embedded
-programming system that maps websites to a structured data table,
-enabling end users to modify their behavior. We hope that it contributes
-to making the web into a more dynamic medium that users can mold to
-their own needs.
-
-We plan to continue developing the system and to eventually deploy it as
-an open-source tool. To receive future updates on Wildcard and
-notifications about a public release, \href{}{sign up for the email
+We plan to continue developing the Wildcard prototype and to eventually
+deploy it as an open-source tool. To receive future updates on Wildcard
+and notifications about a public release, \href{}{sign up for the email
 newsletter}.
@@
 We are also looking for private beta testers. If you have an idea for
-how you might want to use Wildcard, \href{mailto:glitt@mit.edu}{please
-contact us}. We would love to collaborate on building site adapters and
-formulas to support your use case.
+how you might want to use Wildcard, please
+\href{mailto:glitt@mit.edu}{get in touch}. We would love to hear about
+your needs and help find ways to use Wildcard to solve them.
 

### 2020-01-10 compile-latex.sh

@@ -1,5 +1,8 @@
 set -e
 
+convert media/airbnb-demo.png -density 300 -units pixelsperinch media/airbnb-demo-300dpi.png
+convert media/expedia-demo.png -density 300 -units pixelsperinch media/expedia-demo-300dpi.png
+
 pandoc \
   --filter conditional-render \
   --filter pandoc-crossref \

### 2020-01-10 paper.tex
finished pdf version of demos

@@
 example of someone using it to help with booking a trip on the travel
-sites Airbnb and Expedia. These demos are best viewed in the online
-version of this paper, which includes videos:
-\url{https://geoffreylitt.com/wildcard}
+sites Airbnb and Expedia.{ These demos are best viewed as videos in the
+online version of this paper
+(\url{https://www.geoffreylitt.com/wildcard}).}
 
@@
 \includegraphics{media/airbnb-demo-300dpi.png}
-\caption{Using Wildcard to augment the Airbnb page for searching travel
+\caption{Using Wildcard to augment the Airbnb search page for booking
 accommodations}\label{fig:airbnb-demo}
@@
 rating. Using Wildcard, the user can add these small features, while
-leaving the page's design and the rest of its functionality unchanged
-(shown in fig.~\ref{fig:sorting}).
+leaving the page's design and the rest of its functionality unchanged.{
+fig.~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
+site.}
 
-After manipulating the data, the user can close the table view and
-continue using the website with its original visual design. The table
-view offers a way to change the data backing a page, but does not need
-to replace the original interface entirely.
+First, the user opens up the Wildcard panel, which shows a table
+corresponding to the search results in the page. As they click around in
+the table, the corresponding row in the page is highlighted so they can
+see the connection between the views.
+
+Then, the user can use standard spreadsheet column header features to
+sort the page by price and filter by rating{
+(fig.~\ref{fig:airbnb-demo}B)}.
+
+After manipulating the data, the user was able to close the table view
+and continue using the website with its original visual design. The
+table view offers a way to change the data backing a page, but does not
+need to replace the original interface entirely.
 
@@
 add them to a list of favorites. Using Wildcard, the user can just
-select multiple rows and favorite all of them with one click. Similarly,
-we can also open the detailed pages for many listings in new tabs.
-(Shown in fig.~\ref{fig:actions})
+select multiple rows and favorite all of them with one click{
+(fig.~\ref{fig:airbnb-demo}D)}. Similarly, we can also open the detailed
+pages for many listings in new tabs.
+
+Now the user would like to jot down some notes on the pros and cons of
+each listing. To do this, they can simply type notes into an additional
+column next to each listing, and the notes appear inside the listings in
+the original UI{ (fig.~\ref{fig:airbnb-demo}A)}. These annotations are
+also persisted in the browser for future retrieval.
+
+Wildcard also includes a formula language which enables more
+sophisticated tweaks that fetch external data and perform computations.
+
+When traveling without a car, it's nice to evaluate potential places to
+stay based on how walkable the surroundings. Using Wildcard formulas, we
+can integrate Airbnb with Walkscore, an API that rates the walkability
+of any location on a 1-100 scale. When we call the \texttt{WALKSCORE}
+formula with the latitude and longitude of the listing, it returns the
+walk score as the cell value. Because the cell's contents are injected
+into the page, the score also shows up in the page body{
+(fig.~\ref{fig:airbnb-demo}C)}.
 
@@
 \end{figure}
+
+It might seem that Wildcard is only useful on websites that display
+lists of tabular data like search results. But in fact, the table
+metaphor is flexible enough to represent many types of data. For
+example, a flight search form on Expedia can be represented as a single
+row, with a column corresponding to each input. {
+fig.~\ref{fig:expedia-demo} shows an overview of augmenting the Expedia
+site.}
+
+In previous examples the data extracted from the site was marked as
+read-only; users cannot change the name or price of an Airbnb listing.
+In this next case, the cells are marked as writable, so that changes in
+the table are reflected in the form inputs. This becomes useful when
+combined with GUI widgets for editing the value of a table cell.
+
+Filling in dates for a flight search typically requires opening up a
+separate calendar app to find the right dates, and then manually copying
+them into the form. In Wildcard, we can make this easier by providing a
+datepicker widget that has privileged access to the user's calendar
+information{ (fig.~\ref{fig:expedia-demo}B)}.
 
@@
 \OperatorTok{\{}
-\NormalTok{    fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,} \CommentTok{// The name of the data field}
+\NormalTok{  fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,} \CommentTok{// The name of the field}
+\NormalTok{  type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}      \CommentTok{// The type of the field}
 \NormalTok{  readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}    \CommentTok{// Whether the user can edit the field}
-\NormalTok{    type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}      \CommentTok{// The type of the field}
-    \CommentTok{// Given the DOM element for the entire listing,}
-    \CommentTok{// return the DOM element representing this field}
-\NormalTok{    el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
+\NormalTok{  el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,} \CommentTok{// Extract DOM element}
 \OperatorTok{\}}
@@
 not performed directly by the site adapter; instead, the adapter
-specifies how to find the divs representing data rows, and the core
+specifies how to find the spans representing data rows, and the core
 platform mutates the DOM to reflect the table state. The only exception

### 2020-01-10 compile-html.sh
feeling good about this draft

@@ -5,6 +5,7 @@ set -e
   --filter pandoc-citeproc \
   --metadata=format:html \
   -s \
+  --number-sections \
   -o paper.html \
   --css basic.css \
   --toc \

### 2020-01-10 paper.tex
nice draft

@@
 
-In 2012, the travel site Airbnb removed the ability to sort listings by
-price. Users could still filter by price range, but could no longer view
-the cheapest listings first. Many users complained on online message
-boards that the change seemed hostile to users. ``It's so
-frustrating!..What is the logic behind not having this function?'' said
-one user on the
-\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
-support forum}. Alas, the feature remains missing to this day.
-
-This is a familiar situation in a world of web applications that are
-frequently updated without user consent. Sometimes there is a browser
-extension or user script that fixes an issue, and if the user is both
-motivated and skilled they might even be able to implement their own
-fix. But for most people, the only recourse is to complain to the
-developers and pray that someone listens---or more likely, to simply
-give up. While many have become used to this status quo, we see it as a
-tremendous waste of the openness of the Web platform. Back in 1977, in
-\emph{Personal Dynamic Media} \autocite{kay1977}, Alan Kay originally
-envisioned personal computing as a medium that let a user ``mold and
-channel its power to his own needs,'' but today software feels more like
-concrete than clay.
+Web applications often don't match the particular needs of their users.
+Sometimes there is a browser extension available to patch an issue, and
+if the user is a programmer they might be able to fix it themselves. But
+for most people, the only recourse is to complain to the developers, or
+more likely, to simply give up. Back in 1977, in \emph{Personal Dynamic
+Media} \autocite{kay1977}, Alan Kay envisioned personal computing as a
+medium that let a user ``mold and channel its power to his own needs,''
+but today, software feels more like concrete than clay.
 
 In this paper, we present \emph{spreadsheet-driven customization}, a
-technique for making software customization more accessible to end
-users. The core idea is to show the data inside the application in a
-spreadsheet which maintains a live connection to the original UI. When
-the user manipulates the spreadsheet, the UI is instantly modified, and
-vice versa. We built a research prototype called Wildcard that
-implements spreadsheet-driven customization as a web browser extension,
-showing a spreadsheet panel in the context of existing sites.
+technique that enables end users to customize software. The idea is to
+augment an application's UI with a spreadsheet that is synchronized with
+the application's data. When the user manipulates the spreadsheet, the
+underlying data is modified and the changes are propagated to the UI,
+and vice versa. We have implemented this technique in a prototype
+browser extension called Wildcard. Section 2 presents demos of using
+Wildcard to augment real websites in useful ways, and Section 3
+describes the implementation of the extension.
 
-Spreadsheets have proven to be a widely successful computing platform
-for non-programmers. They provide a low barrier to entry for beginners,
-while allowing experts to perform complex computations. The 2D grid is
-simple to grasp, but general enough to represent many types of data.
-Prior work \autocite{mccutchen2016,benson2014,chang2014} has used these
-benefits to enable end users to easily create web applications that use
-spreadsheets as a backing data layer.
+Spreadsheet-driven customization provides an easy entry point for end
+users, since small tweaks can be performed with a single click. At the
+same time, it also supports a variety of richer customizations, like
+adding private annotations to a webpage or joining in related data from
+a web API. In Section 4, we elaborate on this principle of ``low floor,
+high ceiling,'' as well as other design principles guiding our work.
 
-Spreadsheet-driven customization applies these exact same benefits in a
-different context: customizing existing software, rather than building
-new software from scratch. We inherit the low floor and high ceiling of
-spreadsheets: small tweaks like sorting a list of data in an app can be
-done with a single click, while more complex customizations, like
-joining in related data from a web API, are supported by a rich formula
-system. The tabular data format is versatile, supporting data from a
-wide range of applications. Spreadsheet-driven customization does not
-require that the application is actually backed by a spreadsheet---it
-merely presents a model of the internal state of the application, which
-can be exposed by the application itself or derived from the user
-interface by third parties. This means that spreadsheet-driven
-customization works with existing applications that people already use.
-It also allows for exposing a common data abstraction across
-applications, enabling users to have a consistent mechanism for
-manipulating data from all of their applications.
+Prior work \autocite{mccutchen2016,benson2014,chang2014} has enabled end
+users to create ``spreadsheet-driven applications'' which use
+spreadsheets as a backing data layer. Spreadsheet-driven
+\emph{customization} applies this idea in a different context:
+customizing existing software, rather than building new software from
+scratch. Our technique does not require that the application actually be
+backed by a spreadsheet; it merely uses the spreadsheet as an interface
+for viewing and modifying the internal state of the application.
 
-Prior work \autocite{huynh2006} has shown that the process of extracting
-structured data from existing user interfaces can be confusing and
-unfamiliar for end users. In order to make the customization experience
-as straightforward as possible, we envision a tiered architecture that
-hides this complexity from end users. Wildcard provides a mechanism for
-programmers to write \emph{site adapters}, which use web scraping
-techniques to extract structured data from the DOM and AJAX requests of
-existing applications. End users only interact with the structured
-spreadsheet, providing a predictable experience with clear affordances
-for which customizations are possible.
+This approach requires extracting structured data from the user
+interfaces of existing applications, but we hide the complexity of data
+extraction from end users. Programmers write \emph{site adapters} which
+use web scraping techniques to extract structured data from existing
+applications and map them to the spreadsheet table. End users only
+interact with the structured spreadsheet, providing a straightforward
+customization experience. In Section 5, we explain this architecture in
+more depth, and describe how Wildcard relates to existing work on
+malleable software, web customization, spreadsheet-driven applications,
+and web scraping.
 
-Wildcard is currently an early prototype, with incomplete features and
-limited coverage of sites. We plan to continue building site adapters
-and testing the system with our own use cases, to better understand how
-well the spreadsheet abstraction maps to real websites and customization
-needs. Eventually we also plan to release the tool publicly, to
-determine how end users choose to use spreadsheet-driven customization,
-what usability challenges emerge, and how feasible it is for programmers
-to build and maintain site adapters.
+The Wildcard extension is currently an early research prototype. We plan
+to continue testing the system with our own use cases to understand
+better how well the spreadsheet abstraction maps to real websites and
+customization needs. Eventually we plan to release the tool publicly, in
+order to study how end users choose to customize applications, discover
+usability challenges, and to test the feasibility of programmers
+building and maintaining site adapters.
 
@@
 
-To get a sense of how it might feel to use Wildcard, let's see an
-example of someone using it to help with booking a trip on the travel
-sites Airbnb and Expedia.{ These demos are best viewed as videos in the
-online version of this paper
+To get a sense of the user experience of using Wildcard, let's see an
+example of someone using it to help with booking a trip.{ These demos
+are best viewed as videos in the online version of this paper
 (\url{https://www.geoffreylitt.com/wildcard}).}
@@
 
-The user starts by opening up the Airbnb search listings page to look
-for a place to stay. The page looks nice and mostly works well, but is
-missing some key features. As mentioned before, this page doesn't allow
-the user to sort by price. It also doesn't let them filter by user
-rating. Using Wildcard, the user can add these small features, while
-leaving the page's design and the rest of its functionality unchanged.{
+In 2012, the travel site Airbnb removed the ability to sort
+accommodation listings by price. Users could still filter by price
+range, but could no longer view the cheapest listings first. Many users
+complained on online message boards that the change seemed hostile to
+users. ``It's so frustrating!..What is the logic behind not having this
+function?'' said one user on the
+\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
+support forum}. Alas, the feature remains missing to this day.
+
+Using Wildcard, the user can fix this omission, while leaving the page's
+design and the rest of its functionality unchanged.{
 fig.~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
-site.}
-
-First, the user opens up the Wildcard panel, which shows a table
+site.} First, the user opens up the Wildcard panel, which shows a table
 corresponding to the search results in the page. As they click around in
@@
 
-Then, the user can use standard spreadsheet column header features to
-sort the page by price and filter by rating{
-(fig.~\ref{fig:airbnb-demo}B)}.
+Then, the user clicks on the price column header to sort the
+spreadsheet, and the Airbnb UI, by in ascending order by price{
+(fig.~\ref{fig:airbnb-demo}B)}. They can similarly use the column
+headers to filter to listings with a user rating above 4.5, which the
+original Airbnb UI also doesn't allow.
 
-After manipulating the data, the user was able to close the table view
-and continue using the website with its original visual design. The
-table view offers a way to change the data backing a page, but does not
-need to replace the original interface entirely.
+After manipulating the data, the user closes the table view and
+continues using the website with its original visual design. The table
+view offers a way to change the data backing a page, but does not need
+to replace the original interface entirely.
 
-Most websites that show lists of data also offer actions that can be
-taken on a row in the table, like adding an item to a shopping cart.
-Wildcard has the ability to make these ``row actions'' available in the
-data table if the site adapter implements them. The main advantage this
-provides is the ability to easily perform a row action in bulk across
-multiple rows.
+Most websites that show lists of data also offer actions on rows in the
+table, like adding an item to a shopping cart. Wildcard has the ability
+to make these ``row actions'' available in the data table through the
+site adapter.
 
-For example, it's tedious on Airbnb to click on listings one by one to
-add them to a list of favorites. Using Wildcard, the user can just
-select multiple rows and favorite all of them with one click{
-(fig.~\ref{fig:airbnb-demo}D)}. Similarly, we can also open the detailed
-pages for many listings in new tabs.
+In the Airbnb UI, saving multiple listings to a Favorites list requires
+tediously clicking through them one by one. Using Wildcard, the user can
+select multiple rows and favorite all of them with a single click{
+(fig.~\ref{fig:airbnb-demo}D)}. Similarly, the user can also open the
+detailed pages for many listings at once.
 
-Now the user would like to jot down some notes on the pros and cons of
-each listing. To do this, they can simply type notes into an additional
-column next to each listing, and the notes appear inside the listings in
-the original UI{ (fig.~\ref{fig:airbnb-demo}A)}. These annotations are
-also persisted in the browser for future retrieval.
+Now the user wants to jot down some notes on the pros and cons of each
+listing. To do this, they type some notes into an additional column in
+each listing row, and the notes appear inside the listings in the
+original UI{ (fig.~\ref{fig:airbnb-demo}A)}. The annotations are saved
+in the browser for future sessions.
 
-Wildcard also includes a formula language which enables more
-sophisticated tweaks that fetch external data and perform computations.
+Wildcard also includes a formula language, which enables more
+sophisticated customizations that fetch external data and perform
+computations.
 
-When traveling without a car, it's nice to evaluate potential places to
-stay based on how walkable the surroundings. Using Wildcard formulas, we
-can integrate Airbnb with Walkscore, an API that rates the walkability
-of any location on a 1-100 scale. When we call the \texttt{WALKSCORE}
-formula with the latitude and longitude of the listing, it returns the
-walk score as the cell value. Because the cell's contents are injected
-into the page, the score also shows up in the page body{
-(fig.~\ref{fig:airbnb-demo}C)}.
+When traveling without a car, it's useful to evaluate potential places
+to stay based on how walkable the surroundings are. Using Wildcard
+formulas, the user can integrate Airbnb with Walkscore, an API that
+rates the walkability of any location on a 1-100 scale. When they type
+in a \texttt{WALKSCORE} formula with the latitude and longitude of the
+listing, it returns the walk score as the cell's value. Because the
+cell's contents are injected into the page, the score also shows up in
+the page body{ (fig.~\ref{fig:airbnb-demo}C)}.
 
@@
 
-In previous examples the data extracted from the site was marked as
-read-only; users cannot change the name or price of an Airbnb listing.
-In this next case, the cells are marked as writable, so that changes in
-the table are reflected in the form inputs. This becomes useful when
-combined with GUI widgets for editing the value of a table cell.
+In previous examples the table cells were read-only; users cannot change
+the name or price of an Airbnb listing. In this next case, the cells are
+writable, so that changes in the table are reflected in the form inputs.
+This becomes especially useful when combined with GUI widgets that can
+edit the value of a table cell.
 
 Filling in dates for a flight search typically requires opening up a
-separate calendar app to find the right dates, and then manually copying
-them into the form. In Wildcard, we can make this easier by providing a
-datepicker widget that has privileged access to the user's calendar
-information{ (fig.~\ref{fig:expedia-demo}B)}.
+separate calendar app to find the right dates and then manually copying
+them into the form. In Wildcard, the user can avoid this cumbersome
+workflow by using a datepicker widget that includes the user's personal
+calendar information{ (fig.~\ref{fig:expedia-demo}B)}. The user can
+directly click on the right date, and it gets inserted into the
+spreadsheet and the original form.
 
-Here we've presented just a few possibilities for how to use Wildcard.
-We think the interactive data table offers a flexible computational
-model that can support a wide range of other useful modifications, all
-while remaining familiar and easy to use.
+Here we've presented just a few use cases for spreadsheet-driven
+customization, but we think the interactive data table is flexible
+enough to support a wide range of other useful modifications, all while
+remaining familiar and easy to use.
 
@@
 
-In order to maximize extensibility, Wildcard is implemented as a small
+In order to promote extensibility, Wildcard is implemented as a small
 core program along with several types of plugins: site adapters,
@@
 
-Site adapters specify the bidirectional connection between the web page
-and its structured data representation.
+Site adapters are a key part of Wildcard, since they specify the
+bidirectional connection between the web page and its structured data
+representation.
 
-For extracting data from the page and getting it into structured form,
-Wildcard provides ways to concisely express web scraping logic. For
-example, here is a code snippet for extracting the name of an Airbnb
-listing:
+Wildcard provides an interface for concisely expressing how the DOM
+should be mapped to a data table. For example, here is a code snippet
+for extracting the name of an Airbnb listing:
 
@@
 aren't shown in the Airbnb UI, but they are available in AJAX responses.
-We also plan to add the ability for site adapters to scrape data across
-multiple pages in paginated lists of results (as explored in prior work
-\autocite{huynh2006}).
+In the future we also plan to add the ability for site adapters to
+scrape data across multiple pages in paginated lists of results (as
+explored in prior work \autocite{huynh2006}).
 
@@
 updates from the table to the original page. Most DOM manipulation is
-not performed directly by the site adapter; instead, the adapter
-specifies how to find the spans representing data rows, and the core
-platform mutates the DOM to reflect the table state. The only exception
-is row actions (like favoriting an Airbnb listing), which are
-implemented as imperative Javascript functions that can can mutate the
-DOM, simulate clicks on buttons, etc.
+not performed directly by the site adapter. Instead, Wildcard
+automatically mutates the DOM to reflect the table state, using the same
+declarative spec shown above. The only exception is row actions (like
+favoriting an Airbnb listing), which are implemented as imperative
+Javascript functions that can can mutate the DOM, simulate clicks on
+buttons, etc.
 
@@
 principles, inspired by prior work and our own experimentation. We think
-these principles can also broadly inform the design of end user
-programming tools, especially those that enable users to customize
-existing software.
+these principles can also broadly inform the design of tools for end
+user software customization.
 
@@
 bundle together behavior and data to provide some set of functionality.
-While there are limited points of interoperability, applications
-generally are designed to operate independently of one another. Once a
-user has gotten past learning the basic idioms of modern computing like
-windows and scrollbars, most of the effort invested in learning to use
-one application does not carry over to using other applications.
+While there are some limited points of interoperability, applications
+generally are designed to operate independently of one another.
 
@@
 
-Spreadsheet-driven customization aims to port this UNIX philosophy to
-the world of isolated Web applications, by creating a consistent data
+Spreadsheet-driven customization aims to bring some of this UNIX ethos
+to the world of isolated Web applications, by creating a consistent data
 structure to represent the data inside many applications. In UNIX, the
 universal format is a text stream; in Wildcard, it is a relational
-table: a simple abstraction generic enough to describe the data used in
-many different applications. Because Wildcard maps the data from all
-applications to the table format, users can invest in mastering the
-Wildcard table editor, the formula language, and cell editor UIs, and
-reuse those same tools to customize many different applications.
+table. The table is a simple abstraction which is also generic enough to
+describe the data used in many different applications. Because Wildcard
+maps the data from all applications to the table format, users can
+invest in mastering the Wildcard table editor, the formula language, and
+cell editor UIs, and reuse those same tools to customize many different
+applications.
 
@@
 
-Seymour Papert advocated for programming systems with a ``low floor,''
-making it easy for novices to get started, as well as a ``high
+Seymour Papert advocated for programming systems to have a ``low
+floor,'' making it easy for novices to get started, and a ``high
 ceiling,'' providing a large range of possibilities for more
@@
 spreadsheet-driven customization to meet both of these criteria.
-Although we need more testing with real users to know for sure, we think
-Wildcard clearly provides a low floor, and likely has a flexible enough
-foundation to eventually provide a high ceiling.
 
-One of the most interesting properties of spreadsheets is the amount of
-value they can provide to users who are aware of only a tiny sliver of
-their functionality. Not only can a novice perform some basic activities
-in a spreadsheet with almost no training (e.g., storing tables of
-numbers or computing simple sums), these activities are often actually
-valuable for the user! The fact that useful tasks can be performed early
-on supports the user's natural motivation to continue using the tool,
-and to eventually learn its more powerful features if needed
+One of the most interesting properties of spreadsheets is that users who
+are only know a tiny sliver of their functionality (e.g., storing tables
+of numbers or computing simple sums) can still use them in genuinely
+valuable ways. The fact that useful tasks can be performed early on
+supports the user's natural motivation to continue using the tool, and
+to eventually learn its more powerful features if needed
 \autocite{nardi1991}. In contrast, many traditional programming systems
 require an enormous upfront investment of time and practice before
-someone is able to write a program that actually helps them achieve a
-real task.
+someone is able to write a program that helps them achieve a useful task
+in their life.
 
-As part of ensuring a low floor for spreadsheet-driven customization, we
-have focused on including genuinely valuable features for novices. For
-example, a user can sort a table with a single click, or simply type in
-some annotations. We would expect many Wildcard users to start out using
-these simpler features before potentially moving on to more
-sophisticated features like formulas.
+As part of ensuring a low floor, we have focused on including genuinely
+valuable features for novices. For example, a user can sort a table with
+a single click, or simply type in some annotations. We would expect many
+Wildcard users to start by using these simpler features before
+potentially moving on to more sophisticated features like formulas.
 
 Another aspect of providing a low floor is providing an ``in-place
-toolchain'' \autocite{inkandswitch2019}---minimizing the effort of
-moving from using to customizing, by making customization tools
-available in the same environment where the user is already using the
-software. This quality is distinct from the level of technical skill
-needed to use the tool: for example, setting up a workflow trigger in an
-end user programming system like \href{https://ifttt.com/}{IFTTT} does
-not require much technical skill, but does require leaving the user's
-normal software and entering a separate environment; conversely, running
-a Javascript snippet in the browser console requires programming skills,
+toolchain'' \autocite{inkandswitch2019}: minimizing the effort of moving
+from using to customizing, by making customization tools available in
+the same environment where the user is already using the software. This
+quality is distinct from the level of technical skill needed to use the
+tool. For example, setting up a workflow trigger in an end user
+programming system like \href{https://ifttt.com/}{IFTTT} does not
+require much technical skill, but does require leaving the user's normal
+software and entering a separate environment; conversely, running a
+Javascript snippet in the browser console requires programming skills,
 but can be done immediately and casually in the flow of using a website.
@@
 
-Wildcard provides an in-place toolchain---on any site that supports
-Wildcard, the user is one click away from starting to tweak the site
-with minimal friction. There is no need for them to switch into another
-environment or figure out how to extract data from the page. Once the
-user starts editing, Wildcard also provides live feedback to help users
-understand the changes they are making. Even if a user isn't yet totally
-familiar with Wildcard, they can learn to use the system by trying
-things out and seeing what happens.
+Wildcard provides an in-place toolchain because the spreadsheet can be
+instantly opened in the browser window while using any supported
+website. Once the user starts editing, Wildcard also provides live
+feedback, so that even if a user isn't yet totally familiar with
+Wildcard, they can learn to use the system through experimentation.
 
-Since we have still only built several site adapters and demos, it's
-still too early to tell how high the ceiling is for the customizations
-that can be achieved with Wildcard, but we think that with enough
-operators the formula language could eventually support a wide variety
-of customizations. We plan to explore this aspect further by trying to
-solve more real problems with the system and observing where limitations
-emerge in practice.
+Since we have still only built several site adapters and demos, it is
+still too early to tell exactly how high the ceiling is for the
+customizations that can be achieved with Wildcard. But we think that
+with enough operators, the formula language could support a wide variety
+of customizations---people have managed to use simple spreadsheet
+formula languages to solve a surprisingly large range of problems. We
+plan to explore this aspect further by trying to solve more real
+problems with the system and observing where limitations emerge in
+practice.
 
@@
 writing more complex formulas or even programming macros
-\autocite{nardi1990}. Inspired by this, we aim to make
-spreadsheet-driven customization a collaborative activity that combines
-the different abilities of many users in a collaborative ecosystem.
+\autocite{nardi1990}. Inspired by this collaborative approach, we aim to
+make spreadsheet-driven customization a collaborative activity that
+combines the different abilities of many users.
 
@@
 separate stages: structured data extraction, and using the resulting
-spreadsheet. The first stage is currently only available to programmers
-who can code site adapters in Javascript, whereas the second stage is
-available for any non-programmer end user. This architecture frees end
-users from needing to think about data extraction, and enables a
-community of end users to reuse the efforts of programmers building site
-adapters.
+spreadsheet. The data extraction stage is currently only available to
+programmers who can code site adapters in Javascript, whereas the
+customization stage is available to any non-programmer end user. This
+architecture frees end users from needing to think about data
+extraction, and enables a community of end users to reuse the efforts of
+programmers building site adapters.
 
 The group of users building adapters does not necessarily need to be
-limited only to programmers. In the future we might explore enabling end
-users to also create site adapters, drawing on related work on enabling
-end users to extract structured data from websites
-\autocite{chasins2018,huynh2006}. But even in that case, there would
-still be a separation between motivated, tech-savvy end users building
-adapters and more casual end users just using the spreadsheet view.
+limited only to programmers. In the future, we plan to explore enabling
+end users to also create site adapters, drawing on related work in this
+area \autocite{chasins2018,huynh2006}. But even in that case, we still
+envision a separation between highly motivated, tech-savvy end users
+building adapters, and more casual end users just using the spreadsheet
+view.
 
@@
 original software. Spreadsheet-driven customization does not depend on
-cooperation from first-party website developers: a third party
-programmer can write an adapter for any website, which can access any
-information available in the browser. On the other hand, if first
-parties were to expose structured data from their applications, it would
-avoid the need for adapters and generally make customization a lot
-easier. We think there are compelling reasons for first parties to
-consider doing this. Providing Wildcard support would allow users to
-build extensions to fulfill their own feature requests. It also would
-not necessarily require much effort: adding Wildcard support would be
-more straightforward for a first-party than a third-party because they
-have direct access to the structured data in the page. There is also
-precedent for first parties implementing an official client extension
-API in response to user demand: for several years, Google maintained an
-official extension API in Gmail for Greasemonkey scripts to
-use.\footnote{Incidentally, since then, third parties have continued to
-  maintain stable Gmail extension APIs used by many browser extensions
+cooperation from first-party website developers, but if they were to
+expose structured data in their web application clients, it would make
+site adapters unnecessary, producing more robust results with much less
+work.
+
+We think there are compelling reasons for first parties to consider
+doing this. Providing Wildcard support would allow users to build
+extensions to fulfill their own feature requests. It also would not
+necessarily require much effort: adding Wildcard support could be made
+fairly straightforward for a first-party that already has direct access
+to the structured data in the page. There is also precedent for first
+parties implementing an official client extension API in response to
+user demand: for several years, Google maintained an official extension
+API in Gmail for Greasemonkey scripts to use.\footnote{Incidentally,
+  since then, third parties have continued to maintain stable Gmail
+  extension APIs used by many browser extensions
   \autocite{streak,talwar2019}, illustrating the potential of
@@
 
-While similar in high-level goals, Wildcard employs a different solution
-strategy from these projects: whereas they generally require building
-new software from scratch for that environment, Wildcard instead aims to
-maximize the malleability of already existing software. This approach
-has the pragmatic benefit of being immediately useful in more
-situations, although it also requires working within more rigid
-constraints.
+While similar in broad goals, Wildcard employs a different solution
+strategy. These projects generally require building software from
+scratch in a new environment, whereas Wildcard aims to maximize the
+malleability of software built with existing tools. This approach has
+the pragmatic benefit of enabling customization for much more software,
+although it also imposes more rigid constraints.
 
-With substantial future work, Wildcard could become more similar to
-these other projects, growing from a platform for tweaking existing
-software into a platform for building new software from scratch. This
-would likely end up resembling existing tools for building
+With substantial future work, we think Wildcard could become more
+similar to these other projects, growing from a platform for tweaking
+existing software into a platform for building new software from
+scratch. This would likely end up resembling existing tools for building
 spreadsheet-driven applications (discussed more below), but with an
-extra focus on customizability by end users of the software.
+additional focus on customizability by end users of the software.
 
@@
 
-Wildcard's approach is most similar to a class of tools that identify
-structured data in a web page, and use that structure to support end
-user modification of the page.
+Wildcard's solution approach is most similar to other tools that
+identify structured data in a web page, and then use that structure to
+support end user customization of the page.
 
 Sifter \autocite{huynh2006} enables users to sort and filter lists of
-data on web pages, providing a result similar to Wildcard's sort and
-filter functionality. The underlying mechanism is also similar: Sifter
+data on web pages, which resembles Wildcard's sort and filter
+functionality. The underlying mechanism is also similar, since Sifter
 extracts structured data from the page to enable its user-facing
-functionality. Wildcard aims to extend this approach to support much
-broader functionality than just sorting and filtering. In support of
-this goal, Wildcard also shows the structured data table directly to the
-user, whereas Sifter only shows sort and filter controls, without
-revealing the underlying data table. The extraction mechanism is also
-different: Sifter uses a combination of automated heuristics and
-interactive user feedback to extract data, whereas Wildcard currently
-relies on programmers creating wrappers for extracting structured data,
-likely leading to higher quality extraction but on fewer sites.
+functionality. The systems also have major differences. Wildcard aims to
+extend this approach to support much broader functionality, and shows
+the structured data table directly to the user. In contrast, Sifter only
+supports sorting and filtering, and does not reveal the underlying data
+table. The extraction mechanism is also different: Sifter uses a
+combination of automated heuristics and interactive user feedback to
+enable end users to extract data, whereas Wildcard hides data extraction
+from end users.
 
@@
 Sifter, Thresher augments the experience of original page based on
-identifying structure: once semantic content has been identified, it
+identifying structure; once semantic content has been identified, it
 creates context menus in the original website which allow users to take
-actions based on that content. Wildcard and Thresher share an overall
-approach but focus on complementary parts of the problem: Thresher aims
-to enable end users to create content wrappers, but the actions
-available on the structured data are created by programmers; conversely,
-Wildcard delegates wrapper creation to programmers but gives end users
-more flexibility to use the structured data in an open-ended way.
+actions based on that content. Wildcard and Thresher focus on
+complementary parts of the problem. Thresher aims to enable end users to
+create content wrappers, but the actions available on the structured
+data are created by programmers; conversely, Wildcard delegates wrapper
+creation to programmers but gives end users more flexibility to use the
+structured data in an open-ended way.
 
@@
 expressing a wide variety of commands with minimal training, but it also
-has downsides. It is difficult to know whether a command will
-consistently work over time (in addition to changes to the website,
-changes to the heuristics can also cause problems), and it is not easy
-for users to discover the space of possible commands
-\autocite{little2010}.
+has downsides \autocite{little2010}:
+
+\begin{itemize}
+\tightlist
+\item
+  Reliability: It can be difficult to know whether a command will
+  consistently work over time. In addition to changes to the website,
+  changes to the heuristics can also cause problems.
+\item
+  Discoverability: it can be difficult for users to discover the space
+  of possible commands.
+\end{itemize}
 
@@
 to expose a high degree of structure through the familiar spreadsheet
-table. Wildcard offers more consistency: for example, clicking a sort
+table. Wildcard offers more consistency; for example, clicking a sort
 header will always work correctly as long as the site adapter is
@@
 actions are possible, or, crucially, what actions are \emph{not}
-possible, which is useful to know. On the other hand, Wildcard cannot
-offer coverage of all websites, and has a narrower range of possible
-actions than sloppy tools. We expect that with enough site adapters and
-formulas, these downsides can be mitigated.
+possible, which is useful to know. On the other hand, Wildcard's
+explicit site adapter approach will result in coverage of many fewer
+websites, and also a narrower range of possible actions than ``sloppy''
+tools.
 
@@
 
-Prior work has made the powerful realization that a spreadsheet can
-serve as an end-user-friendly backing data store and computation layer
-for an interactive web application. Research projects like Object
-Spreadsheets \autocite{mccutchen2016}, Quilt \autocite{benson2014},
-Gneiss \autocite{chang2014}, and Marmite \autocite{wong2007}, as well as
+Many others have made the powerful realization that a spreadsheet can
+serve as an backing data store and computation layer for an interactive
+web application, enabling end users to create such applications more
+easily. Research projects like Object Spreadsheets
+\autocite{mccutchen2016}, Quilt \autocite{benson2014}, Gneiss
+\autocite{chang2014}, and Marmite \autocite{wong2007}, as well as
 commercial tools like Airtable Blocks \autocite{zotero-79} and Glide
 \autocite{zotero-81} allow users to view data in a spreadsheet table,
-compute over the data using formulas, and connect the table to a GUI.
-Because many users are already familiar with using spreadsheets, this
-way of creating applications tends to be far easier than traditional
-software methods; for example, in a user study of Quilt, many people
-were able to create applications in under 10 minutes, even if they
-expected it would take them many hours.
+compute over the data using formulas, and then connect the table to a
+GUI. Because many users are already familiar with using spreadsheets,
+this way of creating applications tends to be much easier than
+traditional software methods; for example, in a user study of Quilt,
+many users were able to create applications in under 10 minutes, even if
+they expected it would take them many hours.
 
@@
 generally extracts static data for processing in another environment,
-whereas Wildcard modifies the original page by maintaining a
+whereas Wildcard customizes the application's UI by maintaining a
 bidirectional connection between the extracted data and the page.
@@
 There are still many open questions about spreadsheet-driven
-customization, which we hope to answer through targeted development and
+customization which we hope to answer through targeted development and
 usage of the Wildcard prototype.
 
-The most important question is whether the computational model can
-support a large variety of useful customizations. While initial demos
-are promising, we need to develop more site adapters and use cases to
-explore this further. We plan to continue privately testing the system
-with our own needs, and to eventually deploy the tool publicly, once the
-API is stable enough and can support a critical mass of sites and use
-cases. We also plan to run usability studies to evaluate and improve the
-design of the tool.
+The most important question is: where are the limits of this
+computational model? What types of useful customizations can it support
+or not support? While initial demos suggest a variety of use cases, we
+plan to develop more site adapters and demos to explore this question
+further. We will start with privately testing the system with our own
+needs, and then eventually deploy the tool publicly, once the API is
+stable enough and can support a critical mass of sites and use cases. We
+also plan to run usability studies to evaluate and improve the design of
+the tool.
 
-We suspect that two areas of the current model may prove too limiting.
-First, Wildcard's data model only shows a single table at a time,
+We suspect that two areas of the current model may prove limiting.
+First, Wildcard's data model currently shows a single table at a time,
 without any notion of relationships between tables. A richer data model
-with foreign keys might be necessary to support certain use cases. For
-designing a tabular interface on top of a richer data model, we could
-learn from the interface of Airtable which shows related objects in a
-table cell, or add nested rows as used in other systems
+with foreign keys might help support certain use cases. For designing a
+spreadsheet interface on top of a richer relational model, we could
+learn from other systems with this design
 \autocite{mccutchen2016,bakke2016}.
 
-Second, there is no mechanism for end users to express imperative
-workflows (e.g.~``click this button, and then\ldots{}''); they can only
-write formulas that return data and then inject the resulting data into
-the page. If it proves useful, we might consider adding a system for
-expressing workflows like this, although it would substantially alter
-the computational model and it's not clear how it would fit together
-with the existing table view.
+A second potential limitation is that there is currently no mechanism
+for end users to express imperative workflows with sequences of actions,
+which related work has shown to be useful
+\autocite{bolin2005,leshed2008}. It's not clear how such workflows could
+fit into Wildcard, since spreadsheets have a fundamentally different
+computation model, and site adapters created by programmers cannot
+easily account for every possible action a user would want to take in a
+page.
 
 Another open question is how efficiently site adapters can be created
-for real websites. By making adapters for sites with different types of
-data in different domains, we plan to explore how automated heuristics
-can speed up the adapter creation process, and to create abstractions
-that make it easier for programmers to efficiently create new robust
-adapters.
+for real websites, which often include complex markup. We plan to
+explore this question by creating adapters for sites with different data
+structures and in different domains, and perhaps running user tests with
+programmers. Possible future improvements we've considered include
+developing automated heuristics to assist with the adapter creation
+process, and developing new abstractions that make it easier for
+programmers to efficiently create new robust adapters.
 
@@
 In this paper, we have presented \emph{spreadsheet-driven
-customization}, a way to make software customization more accessible to
-end users by showing the data inside the application in a spreadsheet.
-We hope that this technique contributes to making the Web into a more
-dynamic medium that users can mold to their own needs.
+customization}, a technique that enables end users to customize software
+by augmenting an application's UI with a spreadsheet. We hope that this
+technique contributes to making the Web into a more dynamic medium that
+users can mold to their own needs.
 
@@
 \acks
+Thank you to Glen Chiacchieri, Steve Krouse, Daniel Windham, and Maggie
+Yellen for providing helpful feedback on this work.
+
 \printbibliography

### 2020-01-10 paper.tex
start designing the web version

@@
 
-For an online version of this paper with videos demonstrating the interface, see \url{https://www.geoffreylitt.com/wildcard/}
+This paper is best viewed in the online version, which includes videos demonstrating the interface: \url{https://www.geoffreylitt.com/wildcard/}
 \end{abstract}
@@
 describes the implementation of the extension.
+
+\begin{figure}
+\hypertarget{fig:overview}{%
+\centering
+\includegraphics{media/overview-300dpi.png}
+\caption{An overview of spreadsheet-driven
+customization}\label{fig:overview}
+}
+\end{figure}
 

### 2020-01-10 compile-html.sh

@@ -1,5 +1,7 @@
 set -e
 
+convert media/overview.png -density 300 -units pixelsperinch media/overview-300dpi.png
+
  pandoc \
   --filter conditional-render \
   --filter pandoc-citeproc \

### 2020-01-10 compile-latex.sh

@@ -2,6 +2,7 @@ set -e
 
 convert media/airbnb-demo.png -density 300 -units pixelsperinch media/airbnb-demo-300dpi.png
 convert media/expedia-demo.png -density 300 -units pixelsperinch media/expedia-demo-300dpi.png
+convert media/overview.png -density 300 -units pixelsperinch media/overview-300dpi.png
 
 pandoc \
   --filter conditional-render \

### 2020-01-10 paper.tex
design tweaks

@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\hypertarget{sec:introduction}{%
+\section{Introduction}\label{sec:introduction}}
 
@@
 and vice versa. We have implemented this technique in a prototype
-browser extension called Wildcard. Section 2 presents demos of using
-Wildcard to augment real websites in useful ways, and Section 3
-describes the implementation of the extension.
+browser extension called Wildcard. Section~\ref{sec:demos} presents
+demos of using Wildcard to augment real websites in useful ways, and
+Section~\ref{sec:implementation} describes the implementation of the
+extension.
 
@@
 adding private annotations to a webpage or joining in related data from
-a web API. In Section 4, we elaborate on this principle of ``low floor,
-high ceiling,'' as well as other design principles guiding our work.
+a web API. In Section~\ref{sec:design-principles}, we elaborate on this
+principle of ``low floor, high ceiling,'' as well as other design
+principles guiding our work.
 
@@
 interact with the structured spreadsheet, providing a straightforward
-customization experience. In Section 5, we explain this architecture in
-more depth, and describe how Wildcard relates to existing work on
-malleable software, web customization, spreadsheet-driven applications,
-and web scraping.
+customization experience. In Section~\ref{sec:related-work}, we explain
+this architecture in more depth, and describe how Wildcard relates to
+existing work on malleable software, web customization,
+spreadsheet-driven applications, and web scraping.
 
@@
 
-\hypertarget{demo-booking-a-trip-with-wildcard}{%
-\section{Demo: booking a trip with
-Wildcard}\label{demo-booking-a-trip-with-wildcard}}
+\hypertarget{sec:demos}{%
+\section{Demo: booking a trip with Wildcard}\label{sec:demos}}
 
@@
 
+\hypertarget{customizing-search-results}{%
+\subsection{Customizing search
+results}\label{customizing-search-results}}
+
 In 2012, the travel site Airbnb removed the ability to sort
@@
 design and the rest of its functionality unchanged.{
-fig.~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
+Figure~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
 site.} First, the user opens up the Wildcard panel, which shows a table
@@
 spreadsheet, and the Airbnb UI, by in ascending order by price{
-(fig.~\ref{fig:airbnb-demo}B)}. They can similarly use the column
+(Figure~\ref{fig:airbnb-demo}B)}. They can similarly use the column
 headers to filter to listings with a user rating above 4.5, which the
@@
 select multiple rows and favorite all of them with a single click{
-(fig.~\ref{fig:airbnb-demo}D)}. Similarly, the user can also open the
+(Figure~\ref{fig:airbnb-demo}D)}. Similarly, the user can also open the
 detailed pages for many listings at once.
@@
 each listing row, and the notes appear inside the listings in the
-original UI{ (fig.~\ref{fig:airbnb-demo}A)}. The annotations are saved
+original UI{ (Figure~\ref{fig:airbnb-demo}A)}. The annotations are saved
 in the browser for future sessions.
@@
 cell's contents are injected into the page, the score also shows up in
-the page body{ (fig.~\ref{fig:airbnb-demo}C)}.
+the page body{ (Figure~\ref{fig:airbnb-demo}C)}.
 
@@
 
+\hypertarget{customizing-a-form-input}{%
+\subsection{Customizing a form input}\label{customizing-a-form-input}}
+
 It might seem that Wildcard is only useful on websites that display
@@
 row, with a column corresponding to each input. {
-fig.~\ref{fig:expedia-demo} shows an overview of augmenting the Expedia
-site.}
+Figure~\ref{fig:expedia-demo} shows an overview of augmenting the
+Expedia site.}
 
@@
 workflow by using a datepicker widget that includes the user's personal
-calendar information{ (fig.~\ref{fig:expedia-demo}B)}. The user can
+calendar information{ (Figure~\ref{fig:expedia-demo}B)}. The user can
 directly click on the right date, and it gets inserted into the
@@
 
-\hypertarget{system-architecture}{%
-\section{System Architecture}\label{system-architecture}}
+\hypertarget{sec:implementation}{%
+\section{System Implementation}\label{sec:implementation}}
 
@@
 \OperatorTok{\{}
+  \CommentTok{// Field metadata:}
 \NormalTok{  fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,} \CommentTok{// The name of the field}
@@
 \NormalTok{  readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}    \CommentTok{// Whether the user can edit the field}
-\NormalTok{  el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,} \CommentTok{// Extract DOM element}
+
+  \CommentTok{// Function to extract DOM element:}
+\NormalTok{  el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
 \OperatorTok{\}}
@@
 
-\hypertarget{design-principles}{%
-\section{Design principles}\label{design-principles}}
+\hypertarget{sec:design-principles}{%
+\section{Design principles}\label{sec:design-principles}}
 
@@
 
-\hypertarget{related-work}{%
-\section{Related work}\label{related-work}}
+\hypertarget{sec:related-work}{%
+\section{Related work}\label{sec:related-work}}
 

### 2020-01-10 compile-html.sh

@@ -4,6 +4,7 @@ convert media/overview.png -density 300 -units pixelsperinch media/overview-300d
 
  pandoc \
   --filter conditional-render \
+  --filter pandoc-crossref \
   --filter pandoc-citeproc \
   --metadata=format:html \
   -s \

### 2020-01-10 paper.tex
done for the day

@@
 <concept>
-<concept_id>10002944.10011122.10003459</concept_id>
-<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-<concept>
-<concept_id>10010405.10010476.10010477</concept_id>
-<concept_desc>Applied computing~Publishing</concept_desc>
-<concept_significance>300</concept_significance>
+<concept_id>10011007.10011006.10011066.10011069</concept_id>
+<concept_desc>Software and its engineering~Integrated and visual development environments</concept_desc>
+<concept_significance>500</concept_significance>
 </concept>
@@
 
-\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}
-\ccsdesc[500]{Applied computing~Publishing}
-
+\ccsdesc[500]{Software and its engineering~Integrated and visual development environments}
 % To HERE
@@
 medium that let a user ``mold and channel its power to his own needs,''
-but today, software feels more like concrete than clay.
+but today, software behaves more like concrete than clay.
 
 In this paper, we present \emph{spreadsheet-driven customization}, a
-technique that enables end users to customize software. The idea is to
-augment an application's UI with a spreadsheet that is synchronized with
-the application's data. When the user manipulates the spreadsheet, the
-underlying data is modified and the changes are propagated to the UI,
-and vice versa. We have implemented this technique in a prototype
-browser extension called Wildcard. Section~\ref{sec:demos} presents
-demos of using Wildcard to augment real websites in useful ways, and
+technique that enables end users to customize software without doing any
+traditional programming. The idea is to augment an application's UI with
+a spreadsheet that is synchronized with the application's data. When the
+user manipulates the spreadsheet, the underlying data is modified and
+the changes are propagated to the UI, and vice versa. We have
+implemented this technique in a prototype browser extension called
+Wildcard. Section~\ref{sec:demos} presents demos of using Wildcard to
+augment real websites in useful ways, and
 Section~\ref{sec:implementation} describes the implementation of the
@@
 
-Spreadsheet-driven customization provides an easy entry point for end
-users, since small tweaks can be performed with a single click. At the
-same time, it also supports a variety of richer customizations, like
-adding private annotations to a webpage or joining in related data from
-a web API. In Section~\ref{sec:design-principles}, we elaborate on this
-principle of ``low floor, high ceiling,'' as well as other design
-principles guiding our work.
-
 Prior work \autocite{mccutchen2016,benson2014,chang2014} has enabled end
@@
 for viewing and modifying the internal state of the application.
+
+Spreadsheet-driven customization provides an easy entry point for end
+users, since small tweaks like sorting data can be performed with a
+single click. At the same time, it also supports a variety of richer
+customizations, like adding private annotations to a webpage or joining
+in related data from a web API. In Section~\ref{sec:design-principles},
+we elaborate on this breadth of use cases, as well as other design
+principles guiding our work.
 
@@
 The Wildcard extension is currently an early research prototype. We plan
-to continue testing the system with our own use cases to understand
-better how well the spreadsheet abstraction maps to real websites and
-customization needs. Eventually we plan to release the tool publicly, in
-order to study how end users choose to customize applications, discover
-usability challenges, and to test the feasibility of programmers
-building and maintaining site adapters.
+to continue testing the system with our own use cases to explore how the
+spreadsheet abstraction maps to real websites and customization needs.
+Eventually we plan to release the tool publicly, in order to study real
+applications, discover usability challenges, and to test the feasibility
+of programmers building and maintaining site adapters.
 
@@
 
-To get a sense of the user experience of using Wildcard, let's see an
-example of someone using it to help with booking a trip.{ These demos
-are best viewed as videos in the online version of this paper
+To demonstrate the user experience of Wildcard, here is an example of
+using it to help with booking a trip.{ These demos are best viewed as
+videos in the online version of this paper
 (\url{https://www.geoffreylitt.com/wildcard}).}
@@
 In 2012, the travel site Airbnb removed the ability to sort
-accommodation listings by price. Users could still filter by price
+accommodation searches by price. Users could still filter by price
 range, but could no longer view the cheapest listings first. Many users
-complained on online message boards that the change seemed hostile to
-users. ``It's so frustrating!..What is the logic behind not having this
-function?'' said one user on the
+complained that the change seemed hostile to users. ``It's so
+frustrating!..What is the logic behind not having this function?'' said
+one user on the
 \href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
@@
 
-Then, the user clicks on the price column header to sort the
-spreadsheet, and the Airbnb UI, by in ascending order by price{
-(Figure~\ref{fig:airbnb-demo}B)}. They can similarly use the column
-headers to filter to listings with a user rating above 4.5, which the
-original Airbnb UI also doesn't allow.
+Then, the user clicks on the price column header to sort the spreadsheet
+and the Airbnb UI by price{ (Figure~\ref{fig:airbnb-demo}B)}. They also
+filter to listings with a user rating above 4.5 (another feature missing
+in the original Airbnb UI).
 
@@
 to make these ``row actions'' available in the data table through the
-site adapter.
+site adapter. In the Airbnb UI, saving multiple listings to a Favorites
+list requires tediously clicking through them one by one. Using Wildcard
+row actions, the user can select multiple rows and favorite all of them
+with a single click{ (Figure~\ref{fig:airbnb-demo}D)}. Similarly, the
+user can open the detailed pages for many listings at once.
 
-In the Airbnb UI, saving multiple listings to a Favorites list requires
-tediously clicking through them one by one. Using Wildcard, the user can
-select multiple rows and favorite all of them with a single click{
-(Figure~\ref{fig:airbnb-demo}D)}. Similarly, the user can also open the
-detailed pages for many listings at once.
+Next, the user wants to jot down some notes about each listing. To do
+this, they type some notes into an additional column in each listing
+row, and the notes appear inside the listings in the original UI{
+(Figure~\ref{fig:airbnb-demo}A)}. The annotations are saved in the
+browser for future sessions.
 
-Now the user wants to jot down some notes on the pros and cons of each
-listing. To do this, they type some notes into an additional column in
-each listing row, and the notes appear inside the listings in the
-original UI{ (Figure~\ref{fig:airbnb-demo}A)}. The annotations are saved
-in the browser for future sessions.
-
-Wildcard also includes a formula language, which enables more
-sophisticated customizations that fetch external data and perform
-computations.
-
-When traveling without a car, it's useful to evaluate potential places
-to stay based on how walkable the surroundings are. Using Wildcard
-formulas, the user can integrate Airbnb with Walkscore, an API that
-rates the walkability of any location on a 1-100 scale. When they type
-in a \texttt{WALKSCORE} formula with the latitude and longitude of the
-listing, it returns the walk score as the cell's value. Because the
-cell's contents are injected into the page, the score also shows up in
-the page body{ (Figure~\ref{fig:airbnb-demo}C)}.
+Wildcard also includes a formula language that enables more
+sophisticated customizations. When traveling without a car, it's useful
+to evaluate potential places to stay based on how walkable the
+surroundings are. Using Wildcard formulas, the user can integrate Airbnb
+with Walkscore, an API that rates the walkability of any location on a
+1-100 scale. When they type in a \texttt{walkscore} formula with the
+latitude and longitude of the listing, it sets the walk score as the
+value of the cell. Because the cell's contents are injected into the
+page, the score also shows up in the UI{
+(Figure~\ref{fig:airbnb-demo}C)}.
 
@@
 
-In previous examples the table cells were read-only; users cannot change
-the name or price of an Airbnb listing. In this next case, the cells are
-writable, so that changes in the table are reflected in the form inputs.
-This becomes especially useful when combined with GUI widgets that can
-edit the value of a table cell.
+In previous examples the table cells were read-only, since users cannot
+change the name or price of an Airbnb listing. In this case, the cells
+are writable, which means that changes in the table are reflected in the
+form inputs. This becomes especially useful when combined with GUI
+widgets that can edit the value of a table cell.
 
@@
 workflow by using a datepicker widget that includes the user's personal
-calendar information{ (Figure~\ref{fig:expedia-demo}B)}. The user can
+calendar events{ (Figure~\ref{fig:expedia-demo}B)}. The user can
 directly click on the right date, and it gets inserted into the
@@
 Here we've presented just a few use cases for spreadsheet-driven
-customization, but we think the interactive data table is flexible
-enough to support a wide range of other useful modifications, all while
-remaining familiar and easy to use.
+customization. We think the spreadsheet model is flexible enough to
+support a wide range of other useful modifications, while remaining
+familiar and easy to use.
 
@@
 core program along with several types of plugins: site adapters,
-formulas, and cell renderers/editors. The core contains functionality
-for displaying the data table and handling user interactions, and the
-table implementation is built using the
+formulas, and cell editors. The core contains functionality for
+displaying the data table and handling user interactions, and the table
+implementation is built using the
 \href{https://handsontable.com/}{Handsontable} Javascript library.
@@
 Wildcard provides an interface for concisely expressing how the DOM
-should be mapped to a data table. For example, here is a code snippet
-for extracting the name of an Airbnb listing:
+should be mapped to the spreadsheet table. For example, here is a code
+snippet for extracting the name of an Airbnb listing:
 
@@
 
-Sometimes important data is not shown in the UI, making it impossible to
-scrape from the DOM. To address this, we have also prototyped mechanisms
-for site adapters to observe AJAX requests made by the browser and
-extract data directly from JSON responses. This mechanism was used to
-implement the Airbnb Walkscore example, since latitude and longitude
-aren't shown in the Airbnb UI, but they are available in AJAX responses.
-In the future we also plan to add the ability for site adapters to
-scrape data across multiple pages in paginated lists of results (as
-explored in prior work \autocite{huynh2006}).
+Sometimes sophisticated scraping techniques are necessary to extract the
+necessary data from the application. For example, when we have
+prototyped mechanisms for site adapters to observe AJAX requests made by
+the browser and extract data directly from JSON responses. This
+mechanism was used to implement the Airbnb Walkscore example, since
+latitude and longitude aren't shown in the Airbnb UI, but they are
+available in AJAX responses. This technique seems promising because AJAX
+responses tend to already contain data in a structured form, and web
+applications increasingly load data using AJAX. Another technique we
+might consider adding is the ability for site adapters to scrape data
+across multiple pages for paginated lists of results (as explored in
+\autocite{huynh2006}).
 
@@
 not performed directly by the site adapter. Instead, Wildcard
-automatically mutates the DOM to reflect the table state, using the same
-declarative spec shown above. The only exception is row actions (like
-favoriting an Airbnb listing), which are implemented as imperative
+automatically mutates the DOM to reflect the spreadsheet state, using
+the same declarative spec shown above. The only exception is row actions
+(like favoriting an Airbnb listing), which are implemented as imperative
 Javascript functions that can can mutate the DOM, simulate clicks on
@@
 The universal format creates a high degree of leverage from tools: users
-can get a lot of utility from deeply mastering a text editor and some
-text manipulation utilities, because these tools can be applied to
-nearly any task. As just one example, a user's preferred text editor can
-even serve as an interactive input mechanism in shell programs, e.g.~for
-editing git commit messages.
+can get a lot of utility from deeply mastering a text editor and text
+manipulation utilities like \texttt{cut} and \texttt{sed}, because these
+tools can be applied to a huge variety of tasks. A user's preferred text
+editor can even serve as an interactive input mechanism in shell
+programs, e.g.~for editing git commit messages.
 
-Spreadsheet-driven customization aims to bring some of this UNIX ethos
-to the world of isolated Web applications, by creating a consistent data
+Spreadsheet-driven customization aims to bring some of the UNIX ethos to
+the world of isolated Web applications, by creating a consistent data
 structure to represent the data inside many applications. In UNIX, the
 universal format is a text stream; in Wildcard, it is a relational
-table. The table is a simple abstraction which is also generic enough to
-describe the data used in many different applications. Because Wildcard
-maps the data from all applications to the table format, users can
-invest in mastering the Wildcard table editor, the formula language, and
-cell editor UIs, and reuse those same tools to customize many different
-applications.
+table. Because Wildcard maps the data from all applications to the table
+format, users can invest in mastering the Wildcard table editor, the
+formula language, and cell editor UIs, and reuse those same tools to
+customize many different applications.
 
 This idea relates to Beaudouin-Lafon and Mackay's notion of
-\emph{polymorphic interface instruments} \autocite{beaudouin-lafon2000}:
-UI elements that can be used in different contexts, like a color picker
-that can be used in many different drawing applications. diSessa has
-also noted the connection between literacy and the genericness of a
-medium. Textual literacy rests on the fact that writing can be adapted
-to many different genres and uses \autocite{disessa2000}; if people
-needed to relearn reading and writing from scratch when switching from
-essays to emails, the medium would lose most of its potency. We think
-providing generic tools is especially important for software
-customization, because the most common barrier to customizing software
-is not having enough time \autocite{mackay1991}---it's more likely that
-people will customize software regularly if they can reuse the same tool
-across many applications.
+\emph{polymorphic interface instruments} \autocite{beaudouin-lafon2000},
+which are UI elements that can be used in different contexts (for
+example, a color picker that can be used in many different drawing
+applications). It also relates to ideas of literacy in a medium. diSessa
+notes that textual literacy rests on the fact that writing can be
+adapted to many different genres and uses \autocite{disessa2000}; if
+people needed to relearn reading and writing from scratch when switching
+from essays to emails, the medium would lose most of its potency. Having
+generic tools is especially important for software customization,
+because the most common barrier to customizing software is not having
+enough time \autocite{mackay1991}. It's more likely that people will
+customize software regularly if they can reuse the same tools across
+many applications.
 
@@
 with varied data formats. We expect that many types of data can fit
-fairly naturally into tables: lists of search results, news articles,
-and messages can all be seen as relations. On sites that use document
-structures (e.g.~Google Docs) or graph structures (e.g.~social friend
-graphs), it may prove more challenging to map internal data to this
-abstraction.
-
-Another challenge is ensuring a clear mapping in the user's mind between
-the spreadsheet and the original page. Wildcard provides live visual
-cues as the user navigates the data table (similar to the highlighting
-provided by DOM inspectors in browser developer tools). In our own
-usage, we have found that this live highlighting makes it very clear how
-the two representations map to each other.
+easily into tables: lists of search results, news articles, and messages
+can all naturally be seen as relations. For sites that use document
+structures (e.g.~Google Docs) or graph structures (e.g.~social
+networks), it may prove more challenging to map internal data to the
+table abstraction.
 
@@
 \autocite{nardi1991}. In contrast, many traditional programming systems
-require an enormous upfront investment of time and practice before
-someone is able to write a program that helps them achieve a useful task
-in their life.
+require an enormous upfront time investment before someone is able to
+write a program that helps them achieve a useful task in their life.
 
@@
 
-Real-world spreadsheet usage in offices is highly collaborative: most
-users just perform simple changes, while a few coworkers help with
-writing more complex formulas or even programming macros
-\autocite{nardi1990}. Inspired by this collaborative approach, we aim to
-make spreadsheet-driven customization a collaborative activity that
-combines the different abilities of many users.
+Real-world spreadsheet usage is highly collaborative. Many users just
+perform simple changes, while their coworkers help with writing more
+complex formulas or even programming macros \autocite{nardi1990}.
+Inspired by this, we aim to make spreadsheet-driven customization a
+collaborative activity that combines the different abilities of
+different users.
 
 The main way we do this is by separating website customization into two
-separate stages: structured data extraction, and using the resulting
-spreadsheet. The data extraction stage is currently only available to
-programmers who can code site adapters in Javascript, whereas the
-customization stage is available to any non-programmer end user. This
+separate stages: structured data extraction, performed by programmers
+who code site adapters in Javascript, and then customization using the
+resulting spreadsheet, which is available to all end users. This
 architecture frees end users from needing to think about data
@@
 end users to also create site adapters, drawing on related work in this
-area \autocite{chasins2018,huynh2006}. But even in that case, we still
-envision a separation between highly motivated, tech-savvy end users
-building adapters, and more casual end users just using the spreadsheet
-view.
+area \autocite{chasins2018,huynh2006}. But even then, we still envision
+a separation between highly motivated, tech-savvy end users building
+adapters, and more casual end users just using the spreadsheet view.
 
-Another stakeholder to consider is the first party developers of the
-original software. Spreadsheet-driven customization does not depend on
+Another group to consider is the first party developers of the original
+software. Spreadsheet-driven customization does not depend on
 cooperation from first-party website developers, but if they were to
-expose structured data in their web application clients, it would make
-site adapters unnecessary, producing more robust results with much less
-work.
-
-We think there are compelling reasons for first parties to consider
-doing this. Providing Wildcard support would allow users to build
-extensions to fulfill their own feature requests. It also would not
-necessarily require much effort: adding Wildcard support could be made
-fairly straightforward for a first-party that already has direct access
-to the structured data in the page. There is also precedent for first
-parties implementing an official client extension API in response to
-user demand: for several years, Google maintained an official extension
-API in Gmail for Greasemonkey scripts to use.\footnote{Incidentally,
+expose structured data in their web application clients, it would
+eliminate the need for third party site adapters. We think there are
+compelling reasons for first parties to consider doing this. Providing
+Wildcard support would allow users to build extensions to fulfill their
+own feature requests. It also would not necessarily require much effort,
+since adding Wildcard support could be fairly straightforward for a
+first-party that already has direct access to the structured data in the
+page. There is also precedent for first parties implementing an official
+client extension API: for several years, Google maintained an official
+extension API in Gmail for Greasemonkey scripts to use.\footnote{Incidentally,
   since then, third parties have continued to maintain stable Gmail
@@
 scratch in a new environment, whereas Wildcard aims to maximize the
-malleability of software built with existing tools. This approach has
-the pragmatic benefit of enabling customization for much more software,
-although it also imposes more rigid constraints.
+malleability of software built with existing tools.
 
@@
 spreadsheet-driven applications (discussed more below), but with an
-additional focus on customizability by end users of the software.
+additional focus on customization by end users of the software.
 
@@
 Wildcard's goals are closely shared with other systems that provide
-interfaces in the browser for end users to augment and customize
-websites while using them.
+interfaces in the browser for end users to customize websites.
 
@@
 Sifter \autocite{huynh2006} enables users to sort and filter lists of
-data on web pages, which resembles Wildcard's sort and filter
-functionality. The underlying mechanism is also similar, since Sifter
-extracts structured data from the page to enable its user-facing
-functionality. The systems also have major differences. Wildcard aims to
-extend this approach to support much broader functionality, and shows
-the structured data table directly to the user. In contrast, Sifter only
-supports sorting and filtering, and does not reveal the underlying data
-table. The extraction mechanism is also different: Sifter uses a
+data on web pages, which resembles Wildcard's sorting functionality. The
+underlying mechanism is similar, since Sifter extracts structured data
+from the page to enable its user-facing functionality. However, the
+systems also have significant differences:
+
+\begin{itemize}
+\tightlist
+\item
+  Wildcard aims support a broad set of customizations, but Sifter only
+  supports sorting and filtering.
+\item
+  Wildcard shows the structured data table directly to the user. Sifter
+  only shows sorting controls, and does not reveal the underlying data
+  table.
+\item
+  Wildcard hides data extraction from end users. Sifter uses a
   combination of automated heuristics and interactive user feedback to
-enable end users to extract data, whereas Wildcard hides data extraction
-from end users.
+  involve end users in extracting data.
+\end{itemize}
 
@@
 unstructured website content to Semantic Web content. Like Wildcard and
-Sifter, Thresher augments the experience of original page based on
-identifying structure; once semantic content has been identified, it
-creates context menus in the original website which allow users to take
-actions based on that content. Wildcard and Thresher focus on
-complementary parts of the problem. Thresher aims to enable end users to
-create content wrappers, but the actions available on the structured
-data are created by programmers; conversely, Wildcard delegates wrapper
-creation to programmers but gives end users more flexibility to use the
-structured data in an open-ended way.
+Sifter, Thresher augments the page; once semantic content has been
+identified, it creates context menus in the original website which allow
+users to take actions based on that content. Wildcard and Thresher focus
+on complementary parts of the customization process. Thresher aims to
+enable end users to create content wrappers, but the actions available
+on the structured data are determined in advance. Conversely, Wildcard
+delegates wrapper creation to programmers, but gives end users more
+flexibility to use the structured data in an open-ended way.
 
@@
 \autocite{bolin2005} and Coscripter \autocite{leshed2008} enable users
-to create scripts that perform actions like filling in text boxes and
-clicking buttons, without directly interacting with the DOM. Users
-express the desired page elements in natural, informal terms
-(e.g.~writing ``the username box'' to represent the textbook closest to
-the label ``username''), and then using heuristics to determine which
-elements most likely match the user's intent. This approach allows for
-expressing a wide variety of commands with minimal training, but it also
-has downsides \autocite{little2010}:
+to create web automation scripts without directly interacting with the
+DOM. The scripts can perform actions like filling in text boxes and
+clicking on buttons. Users express the desired page elements in natural,
+informal terms (e.g.~writing ``the username box'' to represent the
+textbox closest to the label ``username''), and then the system uses a
+set of heuristics to determine which elements most likely match the
+user's intent. This approach allows for expressing a wide variety of
+commands with minimal training, but it also has downsides
+\autocite{little2010}:
 
@@
 \item
-  Reliability: It can be difficult to know whether a command will
+  \emph{Reliability}: It can be difficult to know whether a command will
   consistently work over time. In addition to changes to the website,
@@
 \item
-  Discoverability: it can be difficult for users to discover the space
-  of possible commands.
+  \emph{Discoverability}: it can be difficult for users to discover the
+  space of possible commands.
 \end{itemize}
@@
 possible, which is useful to know. On the other hand, Wildcard's
-explicit site adapter approach will result in coverage of many fewer
-websites, and also a narrower range of possible actions than ``sloppy''
-tools.
+explicit site adapter approach means that fewer websites can be
+customized, and also that users can't perform customizations if the
+relevant data is not in the spreadsheet.
 
@@
 
-Many others have made the powerful realization that a spreadsheet can
+Many others have had the powerful realization that a spreadsheet can
 serve as an backing data store and computation layer for an interactive
@@
 \autocite{chang2014}, and Marmite \autocite{wong2007}, as well as
-commercial tools like Airtable Blocks \autocite{zotero-79} and Glide
+commercial tools like Airtable \autocite{zotero-79} and Glide
 \autocite{zotero-81} allow users to view data in a spreadsheet table,
@@
 \autocite{huynh2005} more ambitiously map the data to a rich semantic
-schema. In Wildcard, we chose to avoid schemas, in order to minimize the
-work associated with creating a site adapter.
+schema. In Wildcard, we chose to avoid semantic schemas, in order to
+minimize the work associated with creating a site adapter.
 
-In the future, we might be able to integrate web scraping tools to help
-create more reliable site adapters for Wildcard with less work, and to
-open up adapter creation to end users. Sifter was built on top of the
-Piggy Bank scraping library, suggesting precisely this type of
-architecture where web scraping tools are used to support interactive
-page modification.
+In the future, we might try to integrate existing web scraping tools, to
+help create more reliable site adapters for Wildcard with less work, and
+to open up adapter creation to end users.
 
@@
 
-The most important question is: where are the limits of this
+The most important question is: what are the limits of this
 computational model? What types of useful customizations can it support
@@
 
-A second potential limitation is that there is currently no mechanism
-for end users to express imperative workflows with sequences of actions,
-which related work has shown to be useful
-\autocite{bolin2005,leshed2008}. It's not clear how such workflows could
-fit into Wildcard, since spreadsheets have a fundamentally different
-computation model, and site adapters created by programmers cannot
-easily account for every possible action a user would want to take in a
-page.
+A second limitation is that there is currently no way for end users to
+express imperative workflows with sequences of actions. Related work has
+shown these types of workflows to be useful
+\autocite{bolin2005,leshed2008} in practice. It's not clear how such
+workflows could fit into Wildcard, since spreadsheets have a
+fundamentally different computation model, and site adapters created by
+programmers cannot easily account for every possible action a user would
+want to take in a page.
 
-Another open question is how efficiently site adapters can be created
-for real websites, which often include complex markup. We plan to
-explore this question by creating adapters for sites with different data
-structures and in different domains, and perhaps running user tests with
-programmers. Possible future improvements we've considered include
-developing automated heuristics to assist with the adapter creation
-process, and developing new abstractions that make it easier for
-programmers to efficiently create new robust adapters.
+Another open question is how easily site adapters can be created and
+maintained for real websites, which often include complex markup and
+change frequently. We plan to explore this question by creating adapters
+for sites with different data structures and in different domains, and
+perhaps running user tests with programmers. Possible future
+improvements we've considered include developing automated heuristics to
+assist with the adapter creation process, and developing new
+abstractions that make it easier for programmers to efficiently create
+new robust adapters.
 
@@
 \acks
-Thank you to Glen Chiacchieri, Steve Krouse, Daniel Windham, and Maggie
-Yellen for providing helpful feedback on this work.
+Thanks to Glen Chiacchieri, David Karger, Steve Krouse, Rob Miller, Arvind Satyanarayan, Daniel Windham, and Maggie Yellen for providing valuable feedback on this work.
 

### 2020-01-11 paper.tex
tweaks

@@
 \author{Geoffrey Litt}
-\authorinfo{is a PhD student at MIT, researching programming tools for end users and developers.}
+\authorinfo{is a PhD student at MIT in the Software Design Group. His research focuses on programming tools, especially for live programming and end user programming.}
 \author{Daniel Jackson}
@@
 
+Spreadsheet-driven customization provides an easy entry point for end
+users, since small tweaks like sorting data can be performed with a
+single click. At the same time, it also supports a variety of richer
+customizations, like adding private annotations to a webpage or joining
+in related data from a web API. In Section~\ref{sec:design-principles},
+we elaborate on this breadth of use cases, as well as other design
+principles guiding our work.
+
 Prior work \autocite{mccutchen2016,benson2014,chang2014} has enabled end
@@
 for viewing and modifying the internal state of the application.
-
-Spreadsheet-driven customization provides an easy entry point for end
-users, since small tweaks like sorting data can be performed with a
-single click. At the same time, it also supports a variety of richer
-customizations, like adding private annotations to a webpage or joining
-in related data from a web API. In Section~\ref{sec:design-principles},
-we elaborate on this breadth of use cases, as well as other design
-principles guiding our work.
 
@@
 
-To demonstrate the user experience of Wildcard, here is an example of
+To demonstrate the user experience of Wildcard, here's an example of
 using it to help with booking a trip.{ These demos are best viewed as
@@
 \item
-  Wildcard aims support a broad set of customizations, but Sifter only
+  Wildcard supports a broad set of customizations, but Sifter only
   supports sorting and filtering.

### 2020-01-13 compile-html.sh
todomvc demo

@@ -12,7 +12,7 @@ convert media/overview.png -density 300 -units pixelsperinch media/overview-300d
   -o paper.html \
   --css basic.css \
   --toc \
-  --toc-depth=2 \
+  --toc-depth=1 \
   --variable=toc-title:"Contents" \
   --template=templates/pandoc-template-html.html \
   paper.md

### 2020-01-13 paper.tex
dnj edits, new video

@@
 \author{Geoffrey Litt}
-\authorinfo{is a PhD student at MIT in the Software Design Group. His research focuses on programming tools, especially for live programming and end user programming.}
+\authorinfo{is a PhD student at MIT in the Software Design Group.
+His research focuses on programming tools, especially for live programming and end user programming.
+Contact him at \email{glitt@mit.edu}.}
 \author{Daniel Jackson}
-\authorinfo{is a professor in the Department of Electrical Engineering and Computer Science at MIT, associate director of CSAIL, and a MacVicar Fellow. He leads the Software Design Group.}
+\authorinfo{is Professor in the Department of Electrical Engineering and Computer Science at MIT,  a MacVicar teaching fellow, and Associate Director of the Computer Science and Artificial Intelligence Laboratory.
+He leads the Software Design Group.
+Contact him at \email{dnj@csail.mit.edu}.}
 \affiliation{Massachusetts Institute of Technology}
@@
 view of the data inside the application. By manipulating the
-spreadsheet, users can implement a wide variety of customizations,
-ranging from sorting lists of search results to displaying related data
-from other web services, without doing any traditional programming.
+spreadsheet, users can implement a variety of customizations without
+doing any traditional programming.
 
-We built a prototype system called Wildcard that implements
-spreadsheet-driven customization as a web browser extension. Through
-concrete examples, we demonstrate that Wildcard has both a low barrier
-to entry for beginners and enough flexibility to solve many useful
-problems. We also show that Wildcard can work with real existing
-websites, by extracting structured data using web scraping techniques.
+We have implemented this technique in a prototype browser extension
+called Wildcard. Through concrete examples, we demonstrate that Wildcard
+can be used to implement useful customizations for real websites.
 
@@
 user manipulates the spreadsheet, the underlying data is modified and
-the changes are propagated to the UI, and vice versa. We have
-implemented this technique in a prototype browser extension called
-Wildcard. Section~\ref{sec:demos} presents demos of using Wildcard to
-augment real websites in useful ways, and
-Section~\ref{sec:implementation} describes the implementation of the
-extension.
+the changes are propagated to the UI, and vice versa.
+
+We have implemented this technique in a prototype browser extension
+called Wildcard. We've built demos of using Wildcard (shown in
+Section~\ref{sec:demos}) which suggest that this paradigm can support
+useful customizations, ranging from sorting lists of data to adding
+whole new features to applications.
 
@@
 \end{figure}
+
+Our approach requires extracting structured data from the user
+interfaces of existing applications, but we hide the complexity of data
+extraction from end users. Programmers write \emph{site adapters} which
+use web scraping techniques to extract structured data from existing
+applications and map them to the spreadsheet table. Our prototype
+suggests that it is possible to implement site adapters for real
+websites; Section~\ref{sec:implementation} describes some of the
+techniques and challenges involved.
 
@@
 backed by a spreadsheet; it merely uses the spreadsheet as an interface
-for viewing and modifying the internal state of the application.
-
-This approach requires extracting structured data from the user
-interfaces of existing applications, but we hide the complexity of data
-extraction from end users. Programmers write \emph{site adapters} which
-use web scraping techniques to extract structured data from existing
-applications and map them to the spreadsheet table. End users only
-interact with the structured spreadsheet, providing a straightforward
-customization experience. In Section~\ref{sec:related-work}, we explain
-this architecture in more depth, and describe how Wildcard relates to
-existing work on malleable software, web customization,
-spreadsheet-driven applications, and web scraping.
+for viewing and modifying the internal state of the application. In
+Section~\ref{sec:related-work}, we describe how Wildcard relates to
+existing work on spreadsheet-driven apps, as well as other areas like
+web customization and web scraping.
 
@@
 spreadsheet abstraction maps to real websites and customization needs.
-Eventually we plan to release the tool publicly, in order to study real
-applications, discover usability challenges, and to test the feasibility
-of programmers building and maintaining site adapters.
+Eventually we plan to release the tool publicly in order to study real
+use cases, discover usability challenges, and to test the feasibility of
+programmers building and maintaining site adapters.
 
 \hypertarget{sec:demos}{%
-\section{Demo: booking a trip with Wildcard}\label{sec:demos}}
+\section{Demos}\label{sec:demos}}
 
-To demonstrate the user experience of Wildcard, here's an example of
-using it to help with booking a trip.{ These demos are best viewed as
-videos in the online version of this paper
-(\url{https://www.geoffreylitt.com/wildcard}).}
+Here are some examples of using Wildcard to customize websites in useful
+ways.{ These demos are best viewed as videos in the online version of
+this paper (\url{https://www.geoffreylitt.com/wildcard}).}
 
@@
 complained that the change seemed hostile to users. ``It's so
-frustrating!..What is the logic behind not having this function?'' said
+frustrating! What is the logic behind not having this function?'' said
 one user on the
@@
 Then, the user clicks on the price column header to sort the spreadsheet
-and the Airbnb UI by price{ (Figure~\ref{fig:airbnb-demo}B)}. They also
-filter to listings with a user rating above 4.5 (another feature missing
-in the original Airbnb UI).
+and the Airbnb UI by price{ (Figure~\ref{fig:airbnb-demo}, Note A)}.
+They also filter to listings with a user rating above 4.5 (another
+feature missing in the original Airbnb UI).
 
@@
 row actions, the user can select multiple rows and favorite all of them
-with a single click{ (Figure~\ref{fig:airbnb-demo}D)}. Similarly, the
-user can open the detailed pages for many listings at once.
+with a single click{ (Figure~\ref{fig:airbnb-demo}, Note B)}. Similarly,
+the user can open the detailed pages for many listings at once.
 
@@
 row, and the notes appear inside the listings in the original UI{
-(Figure~\ref{fig:airbnb-demo}A)}. The annotations are saved in the
-browser for future sessions.
+(Figure~\ref{fig:airbnb-demo}, Note C)}. The annotations are saved in
+the browser and associated with the ID of the listing, so they will
+appear in future sessions that display the same listing.
 
@@
 to evaluate potential places to stay based on how walkable the
-surroundings are. Using Wildcard formulas, the user can integrate Airbnb
-with Walkscore, an API that rates the walkability of any location on a
-1-100 scale. When they type in a \texttt{walkscore} formula with the
-latitude and longitude of the listing, it sets the walk score as the
-value of the cell. Because the cell's contents are injected into the
-page, the score also shows up in the UI{
-(Figure~\ref{fig:airbnb-demo}C)}.
+surroundings are. Using a formula, the user can integrate Airbnb with
+Walkscore, an API that rates the walkability of any location on a 1-100
+scale. When the user uses the \texttt{walkscore} function with the
+listing's latitude and longitude as inputs, it returns the walk score
+for that location and shows it as the cell value. Because the cell's
+contents are injected into the page, the score also appears in the UI{
+(Figure~\ref{fig:airbnb-demo}, Note D)}.
+
+\hypertarget{snoozing-todos}{%
+\subsection{Snoozing todos}\label{snoozing-todos}}
+
+\begin{figure}
+\hypertarget{fig:todomvc-demo}{%
+\centering
+\includegraphics{media/todomvc-demo-300dpi.png}
+\caption{Using Wildcard to add a ``snooze'' feature to the TodoMVC todo
+list app}\label{fig:todomvc-demo}
+}
+\end{figure}
+
+In addition to fetching data from other sources, Wildcard formulas can
+also perform computations. In this example, the user wants to augment
+the TodoMVC todo list app with a ``snooze'' feature, which will
+temporarily hide a todo from the list until a certain date.{
+Figure~\ref{fig:todomvc-demo} shows an overview of this customization.}
+
+The user starts by opening the table view, which shows the text and
+completed status of each todo, and live updates as the list changes.
+They start the customization by adding a new \texttt{snoozeDate} column
+to store the snooze date for each todo{ (Figure~\ref{fig:todomvc-demo},
+Note A)}.
+
+The next step is to hide snoozed todos. The user creates a new
+\texttt{snoozed?} column, which uses a formula to compute whether a todo
+has a snooze date in the future{ (Figure~\ref{fig:todomvc-demo}, Note
+B)}. Then, they simply filter the table to hide the snoozed todos{
+(Figure~\ref{fig:todomvc-demo}, Note C)}.
+
+As time passes, the built-in \texttt{NOW()} function will return the
+current date, and snoozed todos will automatically appear at the right
+time.
+
+Because this implementation of snoozing was built on top of the
+spreadsheet abstraction, it is completely decoupled from this particular
+todo list app. We envision that users could package and share these
+types of customizations as generic browser extensions, which could then
+be instantly applied to any site supported by Wildcard.
+
+\hypertarget{customizing-a-form-input}{%
+\subsection{Customizing a form input}\label{customizing-a-form-input}}
 
@@
 
-\hypertarget{customizing-a-form-input}{%
-\subsection{Customizing a form input}\label{customizing-a-form-input}}
+It might seem that Wildcard is only useful on websites that display
+lists of tabular data, but the table metaphor is flexible enough to
+represent many types of data. For example, a flight search form on
+Expedia can be represented as a single row, with a column corresponding
+to each input{ (Figure~\ref{fig:expedia-demo}, Note A)}.
 
-It might seem that Wildcard is only useful on websites that display
-lists of tabular data like search results. But in fact, the table
-metaphor is flexible enough to represent many types of data. For
-example, a flight search form on Expedia can be represented as a single
-row, with a column corresponding to each input. {
-Figure~\ref{fig:expedia-demo} shows an overview of augmenting the
-Expedia site.}
+In some of the previous examples, the table cells were read-only
+(because users can't, for example, change the name or price of an Airbnb
+listing). In this case, the cells are writable, which means that changes
+in the table are reflected in the form inputs. This becomes especially
+useful when combined with GUI widgets that can edit the value of a table
+cell.
 
-In previous examples the table cells were read-only, since users cannot
-change the name or price of an Airbnb listing. In this case, the cells
-are writable, which means that changes in the table are reflected in the
-form inputs. This becomes especially useful when combined with GUI
-widgets that can edit the value of a table cell.
-
-Filling in dates for a flight search typically requires opening up a
-separate calendar app to find the right dates and then manually copying
-them into the form. In Wildcard, the user can avoid this cumbersome
-workflow by using a datepicker widget that includes the user's personal
-calendar events{ (Figure~\ref{fig:expedia-demo}B)}. The user can
-directly click on the right date, and it gets inserted into the
+Filling in dates for a flight search often requires a cumbersome
+workflow: open up a separate calendar app, find the dates for the trip,
+and then carefully copy them into the form. In Wildcard, the user can
+avoid this by using a datepicker widget that shows the user's personal
+calendar events{ (Figure~\ref{fig:expedia-demo}, Note B)}. The user can
+directly click on the correct date, and it gets inserted into both the
 spreadsheet and the original form.
 
-Here we've presented just a few use cases for spreadsheet-driven
-customization. We think the spreadsheet model is flexible enough to
-support a wide range of other useful modifications, while remaining
-familiar and easy to use.
+In this section we've presented just a few use cases for
+spreadsheet-driven customization, to hint at the possibilities of the
+paradigm. We think the spreadsheet model is flexible enough to support a
+wide range of other useful modifications, while remaining familiar and
+easy to use.
 
@@
   \CommentTok{// Field metadata:}
-\NormalTok{  fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,} \CommentTok{// The name of the field}
-\NormalTok{  type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}      \CommentTok{// The type of the field}
-\NormalTok{  readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}    \CommentTok{// Whether the user can edit the field}
+\NormalTok{  fieldName}\OperatorTok{:} \StringTok{"name"}\OperatorTok{,}
+\NormalTok{  type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,}
+\NormalTok{  readOnly}\OperatorTok{:} \KeywordTok{true}\OperatorTok{,}
 
@@
 Sometimes sophisticated scraping techniques are necessary to extract the
-necessary data from the application. For example, when we have
-prototyped mechanisms for site adapters to observe AJAX requests made by
-the browser and extract data directly from JSON responses. This
-mechanism was used to implement the Airbnb Walkscore example, since
-latitude and longitude aren't shown in the Airbnb UI, but they are
-available in AJAX responses. This technique seems promising because AJAX
-responses tend to already contain data in a structured form, and web
-applications increasingly load data using AJAX. Another technique we
-might consider adding is the ability for site adapters to scrape data
-across multiple pages for paginated lists of results (as explored in
-\autocite{huynh2006}).
+necessary data from the application. For example, we have prototyped a
+mechanism for site adapters to observe AJAX requests made by the browser
+and extract data directly from JSON responses. This mechanism was used
+to implement the Airbnb Walkscore example, since latitude and longitude
+aren't shown in the Airbnb UI, but they are available in AJAX responses.
+This technique seems promising because AJAX responses tend to already
+contain data in a structured form, and web applications increasingly
+load data using AJAX. Another technique we might consider adding is the
+ability for site adapters to scrape data across multiple pages for
+paginated lists of results (as explored in \autocite{huynh2006}).
 
@@
 \hypertarget{sec:design-principles}{%
-\section{Design principles}\label{sec:design-principles}}
+\section{Design Principles}\label{sec:design-principles}}
 
@@
 principles, inspired by prior work and our own experimentation. We think
-these principles can also broadly inform the design of tools for end
+these principles might also broadly inform the design of tools for end
 user software customization.
@@
 offers a compelling alternative design: many small single-purpose
-utilities, all of which manipulate a universal format of text streams.
+programs, all of which manipulate a universal format of text streams.
 The universal format creates a high degree of leverage from tools: users
-can get a lot of utility from deeply mastering a text editor and text
-manipulation utilities like \texttt{cut} and \texttt{sed}, because these
-tools can be applied to a huge variety of tasks. A user's preferred text
-editor can even serve as an interactive input mechanism in shell
-programs, e.g.~for editing git commit messages.
+benefit from deeply mastering a text editor and other text manipulation
+tools, since they can be applied to a huge variety of tasks. A user's
+preferred text editor can even serve as an interactive input mechanism
+in shell programs, e.g.~for editing git commit messages.
 
@@
 example, a color picker that can be used in many different drawing
-applications). It also relates to ideas of literacy in a medium. diSessa
+applications). It also relates to ideas of literacy in a medium. DiSessa
 notes that textual literacy rests on the fact that writing can be
@@
 This design principle leads to several challenges. First, any universal
-abstraction has its constraints, and can't necessarily naturally express
-the data in every application. We plan to explore the limits of the
-table abstraction further, by trying to build adapters for more sites
-with varied data formats. We expect that many types of data can fit
-easily into tables: lists of search results, news articles, and messages
-can all naturally be seen as relations. For sites that use document
+abstraction has its constraints, and may not naturally express the data
+in every application. We plan to explore the limits of the table
+abstraction further, by trying to build adapters for more sites with
+varied data formats. We expect that many types of data can fit easily
+into tables: lists of search results, news articles, and messages can
+all naturally be seen as relations. For sites that use document
 structures (e.g.~Google Docs) or graph structures (e.g.~social
@@
 
-One of the most interesting properties of spreadsheets is that users who
-are only know a tiny sliver of their functionality (e.g., storing tables
-of numbers or computing simple sums) can still use them in genuinely
-valuable ways. The fact that useful tasks can be performed early on
-supports the user's natural motivation to continue using the tool, and
-to eventually learn its more powerful features if needed
-\autocite{nardi1991}. In contrast, many traditional programming systems
-require an enormous upfront time investment before someone is able to
-write a program that helps them achieve a useful task in their life.
+One of the most interesting properties of spreadsheets is that users
+familiar with only a tiny sliver of their functionality (e.g., storing
+tables of numbers and computing simple sums) can still use them in
+genuinely valuable ways. This supports the user's natural motivation to
+continue using the tool, and to eventually learn its more powerful
+features if needed \autocite{nardi1991}. In contrast, many traditional
+programming systems require an enormous upfront time investment before
+someone is able to write a program that helps them achieve a useful task
+in their life.
 
-As part of ensuring a low floor, we have focused on including genuinely
-valuable features for novices. For example, a user can sort a table with
-a single click, or simply type in some annotations. We would expect many
-Wildcard users to start by using these simpler features before
-potentially moving on to more sophisticated features like formulas.
+As part of ensuring a low floor, we have focused on including features
+that have immediate value for novices. For example, a user can sort a
+table with a single click, or simply type in some annotations. We would
+expect many Wildcard users to start by using these simpler features
+before potentially moving on to more sophisticated features like
+formulas.
 
@@
 \emph{End user friendly} & \textbf{Wildcard} & IFTTT\tabularnewline
-\emph{Requires programming} & browser JS console & forking open source
-software\tabularnewline
+\emph{Requires programming} & browser JS console & using Web APIs in
+Python\tabularnewline
 \bottomrule
@@
 
-Since we have still only built several site adapters and demos, it is
+Since we have still only built a few site adapters and demos, it is
 still too early to tell exactly how high the ceiling is for the
 customizations that can be achieved with Wildcard. But we think that
-with enough operators, the formula language could support a wide variety
-of customizations---people have managed to use simple spreadsheet
-formula languages to solve a surprisingly large range of problems. We
-plan to explore this aspect further by trying to solve more real
-problems with the system and observing where limitations emerge in
-practice.
+with enough built-in formulas (including standard spreadsheet formulas
+and formulas for fetching data from other sources), the language could
+support a wide variety of customizations. We plan to explore this aspect
+further by trying to solve more real problems with the system and
+observing where limitations emerge in practice.
 
-\hypertarget{build-for-multiple-tiers-of-users}{%
-\subsection{Build for multiple tiers of
-users}\label{build-for-multiple-tiers-of-users}}
+\hypertarget{build-for-an-ecosystem}{%
+\subsection{Build for an ecosystem}\label{build-for-an-ecosystem}}
 
@@
 
-Another group to consider is the first party developers of the original
-software. Spreadsheet-driven customization does not depend on
-cooperation from first-party website developers, but if they were to
-expose structured data in their web application clients, it would
-eliminate the need for third party site adapters. We think there are
-compelling reasons for first parties to consider doing this. Providing
-Wildcard support would allow users to build extensions to fulfill their
-own feature requests. It also would not necessarily require much effort,
-since adding Wildcard support could be fairly straightforward for a
-first-party that already has direct access to the structured data in the
-page. There is also precedent for first parties implementing an official
-client extension API: for several years, Google maintained an official
-extension API in Gmail for Greasemonkey scripts to use.\footnote{Incidentally,
-  since then, third parties have continued to maintain stable Gmail
-  extension APIs used by many browser extensions
-  \autocite{streak,talwar2019}, illustrating the potential of
+The first party developers of the original software can also play a role
+in enabling customization. Spreadsheet-driven customization does not
+depend on cooperation from first-party website developers, but if they
+were to expose structured data in their web application clients, it
+would eliminate the need for third party site adapters. We think there
+are compelling reasons for first parties to consider doing this.
+Providing Wildcard support would allow users to build extensions to
+fulfill their own feature requests. It also would not necessarily
+require much effort, since adding Wildcard support could be fairly
+straightforward for a first-party that already has direct access to the
+structured data in the page. There is also precedent for first parties
+implementing an official client extension API: for several years, Google
+maintained an official extension API in Gmail for Greasemonkey scripts
+to use.\footnote{Incidentally, since then, third parties have continued
+  to maintain stable Gmail extension APIs used by many browser
+  extensions \autocite{streak,talwar2019}, illustrating the potential of
   collaboratively maintaining third party adapters.}
@@
 \hypertarget{sec:related-work}{%
-\section{Related work}\label{sec:related-work}}
+\section{Related Work}\label{sec:related-work}}
 
@@
 
-An important design decision for tools in this space is how to deviate
-from traditional spreadsheets like Microsoft Excel or Google Sheets.
-Quilt and Glide use existing spreadsheet software as a backend,
+An important design decision for tools in this space is how far to
+deviate from traditional spreadsheets like Microsoft Excel or Google
+Sheets. Quilt and Glide use existing spreadsheet software as a backend,
 providing maximum familiarity for users, and even compatibility with
@@
 \hypertarget{future-work}{%
-\section{Future work}\label{future-work}}
+\section{Future Work}\label{future-work}}
 
@@
 deploy it as an open-source tool. To receive future updates on Wildcard
-and notifications about a public release, \href{}{sign up for the email
+and notifications about a public release,
+\href{https://geoffreylitt.substack.com/welcome}{sign up for the email
 newsletter}.

### 2020-01-13 compile-latex.sh

@@ -2,6 +2,7 @@ set -e
 
 convert media/airbnb-demo.png -density 300 -units pixelsperinch media/airbnb-demo-300dpi.png
 convert media/expedia-demo.png -density 300 -units pixelsperinch media/expedia-demo-300dpi.png
+convert media/todomvc-demo.png -density 300 -units pixelsperinch media/todomvc-demo-300dpi.png
 convert media/overview.png -density 300 -units pixelsperinch media/overview-300dpi.png
 
 pandoc \

### 2020-01-13 paper.tex
final tweaks for paper checklist

@@
 \usepackage[backend=biber]{biblatex}
-\addbibresource{references.bib}
+\addbibresource{references2.bib}
 
@@
 \authorinfo{is a PhD student at MIT in the Software Design Group.
-His research focuses on programming tools, especially for live programming and end user programming.
+His research focuses on programming languages and environments, particularly tools for end user programming and live programming.
 Contact him at \email{glitt@mit.edu}.}
@@
 Many Web applications do not meet the particular needs of their users.
-Browser extensions and user scripts offer a way to customize web
-applications, but most people do not have the programming skills to
-implement their own extensions.
+Browser extensions offer a way to customize web applications, but most
+people do not have the programming skills to implement their own
+extensions.
 
@@
 called Wildcard. Through concrete examples, we demonstrate that Wildcard
-can be used to implement useful customizations for real websites.
+can support useful customizations---ranging from sorting lists of search
+results to showing related data from web APIs---on top of existing
+websites. We also present the design principles underlying the
+prototype.
+
+Enabling easier customization can dramatically improve people's
+experiences with software. We think that spreadsheet-driven
+customization offers a promising new approach to the problem.
 
@@
 
-The user starts by opening the table view, which shows the text and
-completed status of each todo, and live updates as the list changes.
-They start the customization by adding a new \texttt{snoozeDate} column
-to store the snooze date for each todo{ (Figure~\ref{fig:todomvc-demo},
-Note A)}.
+The user opens the table view, which shows the text and completed status
+of each todo, and updates live in response to changes. They start the
+customization by adding a new \texttt{snoozeDate} column to store the
+snooze date for each todo{ (Figure~\ref{fig:todomvc-demo}, Note A)}.
 
@@
 formulas, and cell editors. The core contains functionality for
-displaying the data table and handling user interactions, and the table
-implementation is built using the
-\href{https://handsontable.com/}{Handsontable} Javascript library.
+displaying the data table and handling user interactions, and is built
+on the \href{https://handsontable.com/}{Handsontable} Javascript
+library.
 
@@
 \caption{The architecture of the Wildcard
-system}\label{fig:architecture}
+prototype}\label{fig:architecture}
 }
@@
 bidirectional connection between the web page and its structured data
-representation.
-
-Wildcard provides an interface for concisely expressing how the DOM
-should be mapped to the spreadsheet table. For example, here is a code
-snippet for extracting the name of an Airbnb listing:
+representation. Wildcard represents data in a relational table with
+named and typed columns, and provides an interface for concisely
+expressing how the DOM should be mapped to this structure. For example,
+here is a code snippet for extracting the name of an Airbnb listing.
 
@@
 
-Sometimes sophisticated scraping techniques are necessary to extract the
-necessary data from the application. For example, we have prototyped a
-mechanism for site adapters to observe AJAX requests made by the browser
-and extract data directly from JSON responses. This mechanism was used
-to implement the Airbnb Walkscore example, since latitude and longitude
-aren't shown in the Airbnb UI, but they are available in AJAX responses.
-This technique seems promising because AJAX responses tend to already
-contain data in a structured form, and web applications increasingly
-load data using AJAX. Another technique we might consider adding is the
-ability for site adapters to scrape data across multiple pages for
-paginated lists of results (as explored in \autocite{huynh2006}).
+Sometimes, more advanced scraping techniques are necessary to extract
+the necessary data from the application. For example, we have prototyped
+a mechanism for site adapters to observe AJAX requests made by the
+browser and extract data directly from JSON responses. This mechanism
+was used to implement the Airbnb Walkscore example, since latitude and
+longitude aren't shown in the Airbnb UI, but they are available in AJAX
+responses. This technique seems promising because AJAX responses already
+contain data in a structured form, and web applications are increasingly
+loading data using AJAX. Another technique we might consider is enabling
+site adapters to scrape data across multiple pages for paginated lists
+of results (as explored in \autocite{huynh2006}).
 
-The site adapter also needs to support the reverse direction: sending
-updates from the table to the original page. Most DOM manipulation is
-not performed directly by the site adapter. Instead, Wildcard
-automatically mutates the DOM to reflect the spreadsheet state, using
-the same declarative spec shown above. The only exception is row actions
-(like favoriting an Airbnb listing), which are implemented as imperative
-Javascript functions that can can mutate the DOM, simulate clicks on
-buttons, etc.
+Because Wildcard requires a live bidirectional connection between the UI
+and the structured data, there are some additional challenges beyond
+those typically associated with web scraping.
+
+First, the site adapter also needs to support sending updates from the
+table to the original page. We have made this fairly straightforward,
+because most DOM manipulation is not performed directly by the site
+adapter---instead, Wildcard automatically mutates the DOM to reflect the
+spreadsheet state, using the same declarative spec shown above. The only
+exception to this is row actions (like favoriting an Airbnb listing),
+which are simply implemented as Javascript functions with the freedom to
+perform arbitrary behaviors like simulating clicks on buttons and
+mutating the DOM.
+
+Another challenge is triggering updates to the spreadsheet data in
+response to UI changes that happen after initial page load. Site
+adapters are responsible for implementing site-specific logic for
+recognizing these changes only by observing the DOM. So far, we have
+been able to use event listeners and the MutationObserver API to
+successfully observe changes, but it may prove challenging to correctly
+respond to updates on some sites only by observing the DOM.
 
@@
 
-\hypertarget{build-for-an-ecosystem}{%
-\subsection{Build for an ecosystem}\label{build-for-an-ecosystem}}
+\hypertarget{design-for-an-ecosystem-of-users}{%
+\subsection{Design for an ecosystem of
+users}\label{design-for-an-ecosystem-of-users}}
 
@@
 maintained an official extension API in Gmail for Greasemonkey scripts
-to use.\footnote{Incidentally, since then, third parties have continued
-  to maintain stable Gmail extension APIs used by many browser
-  extensions \autocite{streak,talwar2019}, illustrating the potential of
-  collaboratively maintaining third party adapters.}
+to use.
+
+Google no longer maintains its own API, but third parties have continued
+to maintain successful replacements---Gmail.js \autocite{talwar2019} is
+an open source wrapper API with 77 contributors and hundreds of commits.
+This shows another benefit of designing for an ecosystem: it's easier
+for many programmers to collectively maintain a single adapter for a
+site and keep it updated in response to changes, instead of creating
+custom scraping logic for individual extensions.
 
@@
 \autocite{disessa1986}, Webstrates \autocite{klokmose2015}, and
-Dynamicland \autocite{victor}.\footnote{The project's name Wildcard
-  comes from the internal pre-release name for Hypercard, which doubly
-  inspired our work by promoting both software modification by end users
-  and the ideas behind the Web.}
+Dynamicland \autocite{victor}. (The project's name Wildcard comes from
+the internal pre-release name for Hypercard, which doubly inspired our
+work by promoting both software modification by end users and the ideas
+behind the Web.)
 

### 2020-01-13 compile-html.sh

@@ -17,6 +17,6 @@ convert media/overview.png -density 300 -units pixelsperinch media/overview-300d
   --template=templates/pandoc-template-html.html \
   paper.md
 
-cp ./paper.html ~/dev/homepage/source/wildcard/index.html
-cp ./basic.css ~/dev/homepage/source/wildcard
-cp -r ./media ~/dev/homepage/source/wildcard
+cp ./paper.html ~/dev/homepage/source/wildcard/convivial/index.html
+cp ./basic.css ~/dev/homepage/source/wildcard/convivial
+cp -r ./media ~/dev/homepage/source/wildcard/convivial

### 2020-01-13 paper.tex
finished

@@
 
-This paper is best viewed in the online version, which includes videos demonstrating the interface: \url{https://www.geoffreylitt.com/wildcard/}
+This paper is best viewed in the online version, which includes videos demonstrating the interface: \url{https://www.geoffreylitt.com/wildcard/salon2020}
 \end{abstract}
@@
 ways.{ These demos are best viewed as videos in the online version of
-this paper (\url{https://www.geoffreylitt.com/wildcard}).}
+this paper (\url{https://www.geoffreylitt.com/wildcard/salon2020}).}
 
@@
 Another aspect of providing a low floor is providing an ``in-place
-toolchain'' \autocite{inkandswitch2019}: minimizing the effort of moving
-from using to customizing, by making customization tools available in
-the same environment where the user is already using the software. This
-quality is distinct from the level of technical skill needed to use the
-tool. For example, setting up a workflow trigger in an end user
-programming system like \href{https://ifttt.com/}{IFTTT} does not
-require much technical skill, but does require leaving the user's normal
-software and entering a separate environment; conversely, running a
-Javascript snippet in the browser console requires programming skills,
-but can be done immediately and casually in the flow of using a website.
+toolchain'' \autocite{2019}: minimizing the effort of moving from using
+to customizing, by making customization tools available in the same
+environment where the user is already using the software. This quality
+is distinct from the level of technical skill needed to use the tool.
+For example, setting up a workflow trigger in an end user programming
+system like \href{https://ifttt.com/}{IFTTT} does not require much
+technical skill, but does require leaving the user's normal software and
+entering a separate environment; conversely, running a Javascript
+snippet in the browser console requires programming skills, but can be
+done immediately and casually in the flow of using a website.
 

### 2020-01-13 compile-html.sh

@@ -17,6 +17,6 @@ convert media/overview.png -density 300 -units pixelsperinch media/overview-300d
   --template=templates/pandoc-template-html.html \
   paper.md
 
-cp ./paper.html ~/dev/homepage/source/wildcard/convivial/index.html
-cp ./basic.css ~/dev/homepage/source/wildcard/convivial
-cp -r ./media ~/dev/homepage/source/wildcard/convivial
+cp ./paper.html ~/dev/homepage/source/wildcard/salon2020/index.html
+cp ./basic.css ~/dev/homepage/source/wildcard/salon2020
+cp -r ./media ~/dev/homepage/source/wildcard/salon2020

### 2020-01-14 paper.tex
submission ready

@@
 \begin{abstract}
-Many Web applications do not meet the particular needs of their users.
+Many Web applications do not meet the precise needs of their users.
 Browser extensions offer a way to customize web applications, but most
@@
 
-We present the idea of \emph{spreadsheet-driven customization}: enabling
-end users to customize existing applications using a live spreadsheet
-view of the data inside the application. By manipulating the
-spreadsheet, users can implement a variety of customizations without
-doing any traditional programming.
+In this paper, we present \emph{spreadsheet-driven customization}, a
+technique that enables end users to customize software without doing any
+traditional programming. The idea is to augment an application's UI with
+a spreadsheet that is synchronized with the application's data. When the
+user manipulates the spreadsheet, the underlying data is modified and
+the changes are propagated to the UI, and vice versa.
 
@@
 results to showing related data from web APIs---on top of existing
-websites. We also present the design principles underlying the
+websites. We also present the design principles underlying our
 prototype.
 
-Enabling easier customization can dramatically improve people's
-experiences with software. We think that spreadsheet-driven
-customization offers a promising new approach to the problem.
+Customization can lead to dramatically better experiences with software.
+We think that spreadsheet-driven customization offers a promising new
+approach to unlocking this benefit for all users, not just programmers.
 
-This paper is best viewed in the online version, which includes videos demonstrating the interface: \url{https://www.geoffreylitt.com/wildcard/salon2020}
+This paper is best viewed in the online version (\url{https://www.geoffreylitt.com/wildcard/salon2020}), which includes videos demonstrating the interface.
 \end{abstract}
@@
 
-Web applications often don't match the particular needs of their users.
+Web applications often don't meet the precise needs of their users.
 Sometimes there is a browser extension available to patch an issue, and
@@
 We have implemented this technique in a prototype browser extension
-called Wildcard. We've built demos of using Wildcard (shown in
-Section~\ref{sec:demos}) which suggest that this paradigm can support
-useful customizations, ranging from sorting lists of data to adding
-whole new features to applications.
+called Wildcard, and used it to build demos which suggest that this
+paradigm can support useful customizations, ranging from sorting lists
+of data to adding whole new features to applications (shown in
+Section~\ref{sec:demos}).
 
@@
 use web scraping techniques to extract structured data from existing
-applications and map them to the spreadsheet table. Our prototype
-suggests that it is possible to implement site adapters for real
-websites; Section~\ref{sec:implementation} describes some of the
-techniques and challenges involved.
+applications and map it to the spreadsheet table. Our prototype suggests
+that it is possible to implement site adapters for real websites;
+Section~\ref{sec:implementation} describes some of the techniques and
+challenges involved.
 
-Spreadsheet-driven customization provides an easy entry point for end
-users, since small tweaks like sorting data can be performed with a
-single click. At the same time, it also supports a variety of richer
-customizations, like adding private annotations to a webpage or joining
-in related data from a web API. In Section~\ref{sec:design-principles},
-we elaborate on this breadth of use cases, as well as other design
-principles guiding our work.
+Spreadsheet-driven customization is based on three design principles,
+described in Section~\ref{sec:design-principles}:
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Expose a universal structure}: By mapping the data in all
+  applications to the same universal table abstraction, Wildcard enables
+  users to learn a generic customization toolkit that they can apply to
+  any site.
+\item
+  \textbf{Low floor, high ceiling}: Wildcard provides an easy entry
+  point for end users, since small tweaks like sorting data can be
+  performed with a single click. At the same time, it also supports a
+  variety of richer customizations, like adding private annotations to a
+  webpage or joining in related data from a web API.
+\item
+  \textbf{Design for an ecosystem of users}: Spreadsheet-driven
+  customization combines the efforts of programmers and end users,
+  rather than putting the full burden on end users. We also envision a
+  role for first party developers to help make their applications more
+  customizable.
+\end{itemize}
 
@@
 for viewing and modifying the internal state of the application. In
-Section~\ref{sec:related-work}, we describe how Wildcard relates to
-existing work on spreadsheet-driven apps, as well as other areas like
-web customization and web scraping.
+Section~\ref{sec:related-work}, we describe further how Wildcard relates
+to existing work on spreadsheet-driven apps, as well as work in other
+areas like web customization and web scraping.
 
@@
 spreadsheet abstraction maps to real websites and customization needs.
-Eventually we plan to release the tool publicly in order to study real
-use cases, discover usability challenges, and to test the feasibility of
-programmers building and maintaining site adapters.
+Eventually we plan to release the tool publicly in order to learn from
+real use cases, discover usability challenges, and to test the
+feasibility of building and maintaining site adapters.
+Section~\ref{sec:conclusion} describes open questions we plan to
+continue exploring, as well as our future plans.
 
@@
 
-\hypertarget{customizing-search-results}{%
-\subsection{Customizing search
-results}\label{customizing-search-results}}
+\hypertarget{augmenting-search-results}{%
+\subsection{Augmenting search results}\label{augmenting-search-results}}
 
@@
 corresponding to the search results in the page. As they click around in
-the table, the corresponding row in the page is highlighted so they can
-see the connection between the views.
+the table, the corresponding row in the page is highlighted to indicate
+the mapping between the views.
 
@@
 After manipulating the data, the user closes the table view and
-continues using the website with its original visual design. The table
-view offers a way to change the data backing a page, but does not need
-to replace the original interface entirely.
+continues using the website. Because the application's UI usually has a
+nicer visual design than a spreadsheet, Wildcard does not aim to replace
+it---at any time, the user can use either the UI, the spreadsheet, or
+both together.
 
-Most websites that show lists of data also offer actions on rows in the
+Many websites that show lists of data also offer actions on rows in the
 table, like adding an item to a shopping cart. Wildcard has the ability
@@
 Next, the user wants to jot down some notes about each listing. To do
-this, they type some notes into an additional column in each listing
-row, and the notes appear inside the listings in the original UI{
+this, they type some notes into an additional column in each row, and
+the notes appear inside the listings in the original UI{
 (Figure~\ref{fig:airbnb-demo}, Note C)}. The annotations are saved in
-the browser and associated with the ID of the listing, so they will
-appear in future sessions that display the same listing.
+the browser and associated with the backend ID of the listing, so they
+will appear in future browser sessions that display the same listing.
 
@@
 Walkscore, an API that rates the walkability of any location on a 1-100
-scale. When the user uses the \texttt{walkscore} function with the
-listing's latitude and longitude as inputs, it returns the walk score
+scale. When the user calls the \texttt{walkscore} formula with the
+listing's latitude and longitude as arguments, it returns the walk score
 for that location and shows it as the cell value. Because the cell's
@@
 The user opens the table view, which shows the text and completed status
-of each todo, and updates live in response to changes. They start the
-customization by adding a new \texttt{snoozeDate} column to store the
-snooze date for each todo{ (Figure~\ref{fig:todomvc-demo}, Note A)}.
+of each todo. They start the customization by adding a new column to
+store the snooze date for each todo{ (Figure~\ref{fig:todomvc-demo},
+Note A)}.
 
-The next step is to hide snoozed todos. The user creates a new
+The next step is to hide snoozed todos. The user creates a
 \texttt{snoozed?} column, which uses a formula to compute whether a todo
-has a snooze date in the future{ (Figure~\ref{fig:todomvc-demo}, Note
-B)}. Then, they simply filter the table to hide the snoozed todos{
-(Figure~\ref{fig:todomvc-demo}, Note C)}.
+is snoozed---i.e., whether it has a snooze date in the future{
+(Figure~\ref{fig:todomvc-demo}, Note B)}. Then, they simply filter the
+table to hide the snoozed todos{ (Figure~\ref{fig:todomvc-demo}, Note
+C)}.
 
-As time passes, the built-in \texttt{NOW()} function will return the
-current date, and snoozed todos will automatically appear at the right
-time.
+Because the built-in \texttt{NOW()} function always returns the current
+datetime, snoozed todos will automatically appear once their snooze date
+arrives.
 
-Because this implementation of snoozing was built on top of the
-spreadsheet abstraction, it is completely decoupled from this particular
-todo list app. We envision that users could package and share these
-types of customizations as generic browser extensions, which could then
-be instantly applied to any site supported by Wildcard.
+Because this implementation of snoozing was built on the spreadsheet
+abstraction, it is completely decoupled from this particular todo list
+app. We envision that users could share these types of customizations as
+generic browser extensions, which could be applied to any site supported
+by Wildcard with no additional effort.
 
-\hypertarget{customizing-a-form-input}{%
-\subsection{Customizing a form input}\label{customizing-a-form-input}}
+\hypertarget{adding-a-custom-datepicker}{%
+\subsection{Adding a custom
+datepicker}\label{adding-a-custom-datepicker}}
 
@@
 In this section we've presented just a few use cases for
-spreadsheet-driven customization, to hint at the possibilities of the
+spreadsheet-driven customization, to suggest some possibilities of the
 paradigm. We think the spreadsheet model is flexible enough to support a
@@
 \centering
-\includegraphics{media/architecture-clean.png}
+\includegraphics{media/architecture-300dpi.png}
 \caption{The architecture of the Wildcard
@@
   \CommentTok{// Function to extract DOM element:}
-\NormalTok{  el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.}\SpecialCharTok{$\{}\NormalTok{titleClass}\SpecialCharTok{\}}\VerbatimStringTok{\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
+\NormalTok{  el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.\_1ebt2xej\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
 \OperatorTok{\}}
@@
 
-First, the site adapter also needs to support sending updates from the
-table to the original page. We have made this fairly straightforward,
-because most DOM manipulation is not performed directly by the site
+First, the site adapter needs to support sending updates from the table
+to the original page. We have made this fairly straightforward, because
+most DOM manipulation is not performed directly by the site
 adapter---instead, Wildcard automatically mutates the DOM to reflect the
@@
 exception to this is row actions (like favoriting an Airbnb listing),
-which are simply implemented as Javascript functions with the freedom to
-perform arbitrary behaviors like simulating clicks on buttons and
-mutating the DOM.
+which are implemented as Javascript functions that can perform arbitrary
+behaviors like simulating clicks on buttons or mutating the DOM.
 
@@
 adapters are responsible for implementing site-specific logic for
-recognizing these changes only by observing the DOM. So far, we have
-been able to use event listeners and the MutationObserver API to
-successfully observe changes, but it may prove challenging to correctly
-respond to updates on some sites only by observing the DOM.
+recognizing these changes by observing the DOM. So far, we have been
+able to use event listeners and the MutationObserver API to successfully
+observe changes, but it may prove challenging to correctly respond to
+updates on some sites purely by observing the DOM.
 
@@
 programs, all of which manipulate a universal format of text streams.
-The universal format creates a high degree of leverage from tools: users
-benefit from deeply mastering a text editor and other text manipulation
-tools, since they can be applied to a huge variety of tasks. A user's
-preferred text editor can even serve as an interactive input mechanism
-in shell programs, e.g.~for editing git commit messages.
+Users get high leverage from mastering a text editor and other text
+manipulation tools, since they can be applied to a huge variety of
+tasks. A user's preferred text editor can even serve as an interactive
+input mechanism in shell programs, e.g.~for editing git commit messages.
 
@@
 table. Because Wildcard maps the data from all applications to the table
-format, users can invest in mastering the Wildcard table editor, the
-formula language, and cell editor UIs, and reuse those same tools to
-customize many different applications.
+format, users can master the tools provided by Wildcard and apply them
+to customize many different applications.
 
@@
 applications). It also relates to ideas of literacy in a medium. DiSessa
-notes that textual literacy rests on the fact that writing can be
+notes that textual literacy depends on the fact that writing can be
 adapted to many different genres and uses \autocite{disessa2000}; if
 people needed to relearn reading and writing from scratch when switching
-from essays to emails, the medium would lose most of its potency. Having
-generic tools is especially important for software customization,
-because the most common barrier to customizing software is not having
-enough time \autocite{mackay1991}. It's more likely that people will
-customize software regularly if they can reuse the same tools across
-many applications.
+from essays to emails, the medium would lose its potency. Having generic
+tools is especially important for software customization, because a
+common barrier to customizing software is not having enough time
+\autocite{mackay1991}. It's more likely that people will customize
+software regularly if they can reuse the same tools across many
+applications.
 
@@
 abstraction further, by trying to build adapters for more sites with
-varied data formats. We expect that many types of data can fit easily
+varied data formats. We expect that many types of data will fit easily
 into tables: lists of search results, news articles, and messages can
-all naturally be seen as relations. For sites that use document
-structures (e.g.~Google Docs) or graph structures (e.g.~social
-networks), it may prove more challenging to map internal data to the
-table abstraction.
+all naturally be seen as relations. Documents (e.g.~Google Docs) or
+graphs (e.g.~social network friend graphs) may prove more challenging to
+map to the table abstraction.
 
@@
 programming systems require an enormous upfront time investment before
-someone is able to write a program that helps them achieve a useful task
-in their life.
+someone is able to write a program that helps them achieve a useful
+task.
 
@@
 expect many Wildcard users to start by using these simpler features
-before potentially moving on to more sophisticated features like
-formulas.
+before moving on to more sophisticated features like formulas.
 
@@
 is distinct from the level of technical skill needed to use the tool.
-For example, setting up a workflow trigger in an end user programming
-system like \href{https://ifttt.com/}{IFTTT} does not require much
-technical skill, but does require leaving the user's normal software and
+For example, setting up a workflow in an end user programming system
+like \href{https://ifttt.com/}{IFTTT} does not require much programming
+skill, but it does require leaving the user's normal software and
 entering a separate environment; conversely, running a Javascript
 snippet in the browser console requires programming skills, but can be
-done immediately and casually in the flow of using a website.
+done immediately in the flow of using a website.
 
@@
 Since we have still only built a few site adapters and demos, it is
-still too early to tell exactly how high the ceiling is for the
-customizations that can be achieved with Wildcard. But we think that
-with enough built-in formulas (including standard spreadsheet formulas
+still too early to tell how high the ceiling is for the customizations
+that can be achieved with Wildcard. But we think that with enough
+built-in formulas (both standard spreadsheet formulas for computation,
 and formulas for fetching data from other sources), the language could
 support a wide variety of customizations. We plan to explore this aspect
-further by trying to solve more real problems with the system and
-observing where limitations emerge in practice.
+further by trying to solve real problems and observing where limitations
+emerge in practice. Section~\ref{sec:demos} also discusses some
+limitations we expect to face.
 
@@
 straightforward for a first-party that already has direct access to the
-structured data in the page. There is also precedent for first parties
-implementing an official client extension API: for several years, Google
-maintained an official extension API in Gmail for Greasemonkey scripts
-to use.
+structured data in the page. (It would certainly be far simpler than
+implementing a third-party site adapter.) There is also precedent for
+first parties implementing an official client extension API: for several
+years, Google maintained an official extension API in Gmail for
+Greasemonkey scripts to use.
 
@@
 to maintain successful replacements---Gmail.js \autocite{talwar2019} is
-an open source wrapper API with 77 contributors and hundreds of commits.
-This shows another benefit of designing for an ecosystem: it's easier
-for many programmers to collectively maintain a single adapter for a
-site and keep it updated in response to changes, instead of creating
-custom scraping logic for individual extensions.
+an open source wrapper API with over 70 contributors and hundreds of
+commits. This library demonstrates another benefit of designing for an
+ecosystem: it's easier for many programmers to collectively maintain a
+single adapter for a site and keep it updated in response to changes to
+the site, rather than each programmer creating custom scraping logic for
+their own individual extensions.
 
@@
 \item
-  Wildcard supports a broad set of customizations, but Sifter only
-  supports sorting and filtering.
+  Wildcard hides data extraction from end users by having programmers
+  develop site adapters, whereas Sifter uses a combination of automated
+  heuristics and interactive user feedback to involve end users in
+  extracting data. User studies of Sifter indicated that users were
+  surprised by the need to participate in the data augmentation process,
+  which influenced our decision to hide this process from the user. We
+  suspect Wildcard's workflow results in a simpler end user experience,
+  but at the cost of only working with a smaller subset of supported
+  websites.
+\item
+  Wildcard supports a broad set of customizations. Sifter only supports
+  sorting and filtering.
 \item
@@
   table.
-\item
-  Wildcard hides data extraction from end users. Sifter uses a
-  combination of automated heuristics and interactive user feedback to
-  involve end users in extracting data.
 \end{itemize}
@@
 \item
-  \emph{Discoverability}: it can be difficult for users to discover the
-  space of possible commands.
+  \emph{Discoverability}: it can be hard for users to find discover
+  possible commands.
 \end{itemize}
@@
 Wildcard offers a sharp contrast to sloppy programming, instead choosing
-to expose a high degree of structure through the familiar spreadsheet
-table. Wildcard offers more consistency; for example, clicking a sort
-header will always work correctly as long as the site adapter is
-maintained. Wildcard also offers clearer affordances for what types of
-actions are possible, or, crucially, what actions are \emph{not}
-possible, which is useful to know. On the other hand, Wildcard's
-explicit site adapter approach means that fewer websites can be
-customized, and also that users can't perform customizations if the
-relevant data is not in the spreadsheet.
+to expose a high degree of structure through the spreadsheet table.
+Wildcard offers more consistency; for example, clicking a sort header
+will always work correctly as long as the site adapter is maintained.
+Wildcard also offers clearer affordances for what types of actions are
+possible. On the other hand, Wildcard's explicit site adapter approach
+means that fewer websites can be customized, and also that users can't
+perform customizations if the relevant data is not exposed in the
+spreadsheet.
+
+Another difference between Wildcard and these tools is that there is no
+way for end users to express imperative workflows with sequences of
+actions in Wildcard. Chickenfoot and Coscripter have shown these types
+of workflows to be useful in practice, but it's not clear how such
+workflows could fit into Wildcard. Spreadsheets have a fundamentally
+different computation model, and site adapters created by programmers
+cannot easily account for every possible action a user might want to
+perform in a page.
 
@@
 help create more reliable site adapters for Wildcard with less work, and
-to open up adapter creation to end users.
+to open up adapter creation to end users. Although Wildcard has
+additional requirements beyond traditional web scraping as discussed in
+Section~\ref{sec:implementation}, there is still enough overlap that
+existing tools might prove helpful.
 
-\hypertarget{future-work}{%
-\section{Future Work}\label{future-work}}
+\hypertarget{sec:conclusion}{%
+\section{Conclusion and Future Work}\label{sec:conclusion}}
+
+In this paper, we have presented \emph{spreadsheet-driven
+customization}, a technique that enables end users to customize software
+by augmenting an application's UI with a spreadsheet. We hope that this
+technique contributes to making the Web into a more dynamic medium that
+users can mold to their own needs.
 
@@
 customization which we hope to answer through targeted development and
-usage of the Wildcard prototype.
+usage of Wildcard.{\footnote{Up to date information about Wildcard is
+  available on the project website
+  (\url{https://geoffreylitt.com/wildcard}), which includes a signup for
+  receiving updates and notifications about a public release. We also
+  welcome hearing from readers who might be interested in being private
+  beta testers or who have feedback on the project.}}
 
@@
 
-We suspect that two areas of the current model may prove limiting.
-First, Wildcard's data model currently shows a single table at a time,
-without any notion of relationships between tables. A richer data model
-with foreign keys might help support certain use cases. For designing a
-spreadsheet interface on top of a richer relational model, we could
-learn from other systems with this design
+We suspect that some areas of the current model may prove overly
+simplistic. For example, Wildcard's data model currently shows a single
+table at a time, without any notion of relationships between tables. A
+richer data model with foreign keys might help support certain use
+cases. For designing a spreadsheet interface on top of a richer
+relational model, we could learn from other systems with this design
 \autocite{mccutchen2016,bakke2016}.
-
-A second limitation is that there is currently no way for end users to
-express imperative workflows with sequences of actions. Related work has
-shown these types of workflows to be useful
-\autocite{bolin2005,leshed2008} in practice. It's not clear how such
-workflows could fit into Wildcard, since spreadsheets have a
-fundamentally different computation model, and site adapters created by
-programmers cannot easily account for every possible action a user would
-want to take in a page.
 
@@
 
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-In this paper, we have presented \emph{spreadsheet-driven
-customization}, a technique that enables end users to customize software
-by augmenting an application's UI with a spreadsheet. We hope that this
-technique contributes to making the Web into a more dynamic medium that
-users can mold to their own needs.
-
-We plan to continue developing the Wildcard prototype and to eventually
-deploy it as an open-source tool. To receive future updates on Wildcard
-and notifications about a public release,
-\href{https://geoffreylitt.substack.com/welcome}{sign up for the email
-newsletter}.
-
-We are also looking for private beta testers. If you have an idea for
-how you might want to use Wildcard, please
-\href{mailto:glitt@mit.edu}{get in touch}. We would love to hear about
-your needs and help find ways to use Wildcard to solve them.
-
 \acks

### 2020-01-14 compile-html.sh

@@ -1,6 +1,7 @@
 set -e
 
 convert media/overview.png -density 300 -units pixelsperinch media/overview-300dpi.png
+convert media/architecture.png -density 300 -units pixelsperinch media/architecture-300dpi.png
 
  pandoc \
   --filter conditional-render \

### 2020-01-14 compile-latex.sh

@@ -4,6 +4,7 @@ convert media/airbnb-demo.png -density 300 -units pixelsperinch media/airbnb-dem
 convert media/expedia-demo.png -density 300 -units pixelsperinch media/expedia-demo-300dpi.png
 convert media/todomvc-demo.png -density 300 -units pixelsperinch media/todomvc-demo-300dpi.png
 convert media/overview.png -density 300 -units pixelsperinch media/overview-300dpi.png
+convert media/architecture.png -density 300 -units pixelsperinch media/architecture-300dpi.png
 
 pandoc \
   --filter conditional-render \
@@ -21,4 +22,5 @@ pdflatex paper.tex
 biber paper
 pdflatex -interaction=batchmode paper.tex
 pdflatex -interaction=batchmode paper.tex
+cp paper.pdf ~/dev/homepage/source/wildcard/salon2020
 open paper.pdf

### 2020-01-14 paper.tex
done

@@
 if the user is a programmer they might be able to fix it themselves. But
-for most people, the only recourse is to complain to the developers, or
+for most people, the only recourse is to complain to the developers, or,
 more likely, to simply give up. Back in 1977, in \emph{Personal Dynamic
@@
 We have implemented this technique in a prototype browser extension
-called Wildcard, and used it to build demos which suggest that this
+called Wildcard and used it to build demos which suggest that this
 paradigm can support useful customizations, ranging from sorting lists
@@
 In addition to fetching data from other sources, Wildcard formulas can
-also perform computations. In this example, the user wants to augment
-the TodoMVC todo list app with a ``snooze'' feature, which will
+also perform computations. In this example, the user would like to
+augment the TodoMVC todo list app with a ``snooze'' feature, which will
 temporarily hide a todo from the list until a certain date.{
@@
 paradigm. We think the spreadsheet model is flexible enough to support a
-wide range of other useful modifications, while remaining familiar and
+wide range of other useful modifications while remaining familiar and
 easy to use.
@@
   \CommentTok{// Function to extract DOM element:}
-\NormalTok{  el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\VerbatimStringTok{\textasciigrave{}.\_1ebt2xej\textasciigrave{}}\NormalTok{)}\OperatorTok{,}
+\NormalTok{  el}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \VariableTok{row}\OperatorTok{.}\FunctionTok{querySelector}\NormalTok{(}\StringTok{".\_1ebt2xej"}\NormalTok{)}\OperatorTok{,}
 \OperatorTok{\}}
@@
 Sometimes, more advanced scraping techniques are necessary to extract
-the necessary data from the application. For example, we have prototyped
-a mechanism for site adapters to observe AJAX requests made by the
-browser and extract data directly from JSON responses. This mechanism
-was used to implement the Airbnb Walkscore example, since latitude and
-longitude aren't shown in the Airbnb UI, but they are available in AJAX
-responses. This technique seems promising because AJAX responses already
-contain data in a structured form, and web applications are increasingly
-loading data using AJAX. Another technique we might consider is enabling
-site adapters to scrape data across multiple pages for paginated lists
-of results (as explored in \autocite{huynh2006}).
+data from the application. For example, we have prototyped a mechanism
+for site adapters to observe AJAX requests made by the browser and
+extract data directly from JSON responses. This mechanism was used to
+implement the Airbnb Walkscore example, since latitude and longitude
+aren't shown in the Airbnb UI, but they are available in AJAX responses.
+This technique seems promising because AJAX responses already contain
+data in a structured form, and web applications are increasingly loading
+data using AJAX. Another technique we might consider is enabling site
+adapters to scrape data across multiple pages for paginated lists of
+results (as explored in \autocite{huynh2006}).
 
@@
 to the original page. We have made this fairly straightforward, because
-most DOM manipulation is not performed directly by the site
-adapter---instead, Wildcard automatically mutates the DOM to reflect the
-spreadsheet state, using the same declarative spec shown above. The only
-exception to this is row actions (like favoriting an Airbnb listing),
-which are implemented as Javascript functions that can perform arbitrary
-behaviors like simulating clicks on buttons or mutating the DOM.
+most DOM manipulation is not performed directly by the site adapter.
+Instead, Wildcard automatically mutates the DOM to reflect the
+spreadsheet state, using the same specification that was used to extract
+the data. The only exception to this is row actions (like favoriting an
+Airbnb listing), which are implemented as Javascript functions that can
+perform arbitrary behaviors like simulating clicks on buttons or
+mutating the DOM.
 
@@
 response to UI changes that happen after initial page load. Site
-adapters are responsible for implementing site-specific logic for
-recognizing these changes by observing the DOM. So far, we have been
-able to use event listeners and the MutationObserver API to successfully
-observe changes, but it may prove challenging to correctly respond to
-updates on some sites purely by observing the DOM.
+adapters are responsible for recognizing these changes by observing the
+DOM. So far, we have been able to use event listeners and the
+MutationObserver API to successfully observe changes, but it may prove
+challenging to observe changes on some sites only through the DOM.
 
@@
 
-The idea of spreadsheet-driven customization is guided by several design
+The idea of spreadsheet-driven customization is guided by three design
 principles, inspired by prior work and our own experimentation. We think
@@
 Computing does not need to be organized this way. For example, UNIX
-offers a compelling alternative design: many small single-purpose
-programs, all of which manipulate a universal format of text streams.
-Users get high leverage from mastering a text editor and other text
-manipulation tools, since they can be applied to a huge variety of
-tasks. A user's preferred text editor can even serve as an interactive
-input mechanism in shell programs, e.g.~for editing git commit messages.
+offers a compelling alternative: many small single-purpose programs, all
+of which manipulate a universal format of text streams. Users get high
+leverage from mastering a text editor and other text manipulation tools,
+since they can be applied to a huge variety of tasks. A user's preferred
+text editor can even serve as an interactive input mechanism in shell
+programs, e.g.~for editing git commit messages.
 
-Spreadsheet-driven customization aims to bring some of the UNIX ethos to
-the world of isolated Web applications, by creating a consistent data
+Spreadsheet-driven customization aims to bring some of this UNIX ethos
+to the world of isolated Web applications by creating a consistent data
 structure to represent the data inside many applications. In UNIX, the
@@
 which are UI elements that can be used in different contexts (for
-example, a color picker that can be used in many different drawing
-applications). It also relates to ideas of literacy in a medium. DiSessa
-notes that textual literacy depends on the fact that writing can be
-adapted to many different genres and uses \autocite{disessa2000}; if
-people needed to relearn reading and writing from scratch when switching
-from essays to emails, the medium would lose its potency. Having generic
-tools is especially important for software customization, because a
-common barrier to customizing software is not having enough time
-\autocite{mackay1991}. It's more likely that people will customize
-software regularly if they can reuse the same tools across many
+example, a color picker that can be used in many drawing applications).
+It also relates to ideas of literacy in a medium. DiSessa notes that
+textual literacy depends on the fact that writing can be adapted to many
+different genres \autocite{disessa2000}: if people needed to relearn
+reading and writing from scratch when switching from essays to emails,
+the medium would lose its potency. Generic tools are especially
+important for software customization, because a common barrier to
+customizing software is not having enough time
+\autocite{mackay1991}---it's more likely that people will customize
+software frequently if they can reuse the same tools across many
 applications.
@@
 This design principle leads to several challenges. First, any universal
-abstraction has its constraints, and may not naturally express the data
-in every application. We plan to explore the limits of the table
-abstraction further, by trying to build adapters for more sites with
+abstraction has its limits, and may not naturally express the data in
+every application. We plan to explore the limits of the table
+abstraction further by trying to build adapters for more sites with
 varied data formats. We expect that many types of data will fit easily
@@
 tables of numbers and computing simple sums) can still use them in
-genuinely valuable ways. This supports the user's natural motivation to
-continue using the tool, and to eventually learn its more powerful
-features if needed \autocite{nardi1991}. In contrast, many traditional
-programming systems require an enormous upfront time investment before
-someone is able to write a program that helps them achieve a useful
-task.
+valuable ways. This supports the user's natural motivation to continue
+using the tool, and to eventually learn its more powerful features if
+needed \autocite{nardi1991}. In contrast, many traditional programming
+systems require an enormous upfront time investment before someone is
+able to write a program that helps them achieve a useful task.
 
@@
 toolchain'' \autocite{2019}: minimizing the effort of moving from using
-to customizing, by making customization tools available in the same
+to customizing by making customization tools available in the same
 environment where the user is already using the software. This quality
@@
 
-Since we have still only built a few site adapters and demos, it is
+Since we have only built a few site adapters and demos so far, it is
 still too early to tell how high the ceiling is for the customizations
@@
 further by trying to solve real problems and observing where limitations
-emerge in practice. Section~\ref{sec:demos} also discusses some
-limitations we expect to face.
+emerge in practice (discussed more in Section~\ref{sec:conclusion}).
 
@@
 
-Real-world spreadsheet usage is highly collaborative. Many users just
-perform simple changes, while their coworkers help with writing more
-complex formulas or even programming macros \autocite{nardi1990}.
-Inspired by this, we aim to make spreadsheet-driven customization a
-collaborative activity that combines the different abilities of
-different users.
+Real-world spreadsheet usage is highly collaborative. Many users only
+perform simple changes, while their coworkers help with writing complex
+formulas or programming macros \autocite{nardi1990}. Inspired by this
+collaborative dynamic, we aim for spreadsheet-driven customization to
+bring together the abilities of different users.
 
@@
 The first party developers of the original software can also play a role
-in enabling customization. Spreadsheet-driven customization does not
-depend on cooperation from first-party website developers, but if they
-were to expose structured data in their web application clients, it
-would eliminate the need for third party site adapters. We think there
-are compelling reasons for first parties to consider doing this.
-Providing Wildcard support would allow users to build extensions to
-fulfill their own feature requests. It also would not necessarily
-require much effort, since adding Wildcard support could be fairly
-straightforward for a first-party that already has direct access to the
-structured data in the page. (It would certainly be far simpler than
-implementing a third-party site adapter.) There is also precedent for
-first parties implementing an official client extension API: for several
-years, Google maintained an official extension API in Gmail for
-Greasemonkey scripts to use.
+in enabling customization. Although spreadsheet-driven customization
+does not depend on cooperation from first-party website developers,
+their participation in exposing structured data would eliminate the need
+for third party site adapters. The first parties could themselves
+benefit from such a system, since providing Wildcard support---a fairly
+low-effort undertaking for a developer with direct access to the
+structured data---would allow users to build extensions to fulfill some
+of their own feature requests. There is also precedent for first parties
+implementing an official client extension API: for several years, Google
+maintained an official extension API in Gmail for Greasemonkey scripts
+to use.
 
@@
 ecosystem: it's easier for many programmers to collectively maintain a
-single adapter for a site and keep it updated in response to changes to
-the site, rather than each programmer creating custom scraping logic for
-their own individual extensions.
+generic site adapter that can support many extensions, rather than each
+programmer creating custom scraping logic for their own single-purpose
+extension.
 
@@
 \section{Related Work}\label{sec:related-work}}
+
+Spreadsheet-driven customization relates to existing work in four areas:
+malleable software, web customization, spreadsheet-driven applications,
+and web scraping.
 
@@
 applications built by programmers. These systems include Smalltalk
-\autocite{kay1977}, Hypercard \autocite{hypercard2019} , Boxer
+\autocite{kay1977}, Hypercard \autocite{hypercard2019}, Boxer
 \autocite{disessa1986}, Webstrates \autocite{klokmose2015}, and
-Dynamicland \autocite{victor}. (The project's name Wildcard comes from
-the internal pre-release name for Hypercard, which doubly inspired our
-work by promoting both software modification by end users and the ideas
-behind the Web.)
+Dynamicland \autocite{victor}. In fact, the name Wildcard comes from the
+internal pre-release name for Hypercard, a project which doubly inspired
+our work by promoting both software modification by end users and the
+ideas behind the Web.
 
-While similar in broad goals, Wildcard employs a different solution
-strategy. These projects generally require building software from
-scratch in a new environment, whereas Wildcard aims to maximize the
-malleability of software built with existing tools.
+Although Wildcard shares broad goals with these projects, it employs a
+different solution strategy. These other systems generally require
+building software from scratch in a new environment. On the other hand,
+Wildcard aims to maximize the malleability of software built with
+existing tools.
 
@@
 Wildcard's solution approach is most similar to other tools that
-identify structured data in a web page, and then use that structure to
+identify structured data in a web page and then use that structure to
 support end user customization of the page.
@@
 unstructured website content to Semantic Web content. Like Wildcard and
-Sifter, Thresher augments the page; once semantic content has been
+Sifter, Thresher augments the page: once semantic content has been
 identified, it creates context menus in the original website which allow
@@
   consistently work over time. In addition to changes to the website,
-  changes to the heuristics can also cause problems.
+  changes to the system's heuristics can also cause unexpected changes
+  in behavior.
 \item
@@
 Another difference between Wildcard and these tools is that there is no
-way for end users to express imperative workflows with sequences of
-actions in Wildcard. Chickenfoot and Coscripter have shown these types
-of workflows to be useful in practice, but it's not clear how such
-workflows could fit into Wildcard. Spreadsheets have a fundamentally
-different computation model, and site adapters created by programmers
-cannot easily account for every possible action a user might want to
-perform in a page.
+way for users to express imperative workflows with sequences of actions
+in Wildcard. Chickenfoot and Coscripter have shown these types of
+workflows to be useful in practice so it might be worth considering how
+to support them, but it's not clear how such workflows could fit into
+Wildcard. Spreadsheets have a fundamentally different computation model,
+and site adapters created by programmers cannot easily account for every
+possible action a user might want to perform in a page.
 
-\hypertarget{spreadsheet-based-app-builders}{%
-\subsection{Spreadsheet-based app
-builders}\label{spreadsheet-based-app-builders}}
+\hypertarget{spreadsheet-driven-applications}{%
+\subsection{Spreadsheet-driven
+applications}\label{spreadsheet-driven-applications}}
 
-Many others have had the powerful realization that a spreadsheet can
-serve as an backing data store and computation layer for an interactive
-web application, enabling end users to create such applications more
-easily. Research projects like Object Spreadsheets
-\autocite{mccutchen2016}, Quilt \autocite{benson2014}, Gneiss
-\autocite{chang2014}, and Marmite \autocite{wong2007}, as well as
-commercial tools like Airtable \autocite{zotero-79} and Glide
-\autocite{zotero-81} allow users to view data in a spreadsheet table,
-compute over the data using formulas, and then connect the table to a
-GUI. Because many users are already familiar with using spreadsheets,
-this way of creating applications tends to be much easier than
-traditional software methods; for example, in a user study of Quilt,
-many users were able to create applications in under 10 minutes, even if
-they expected it would take them many hours.
+Many researchers and companies have explored the idea of
+\emph{spreadsheet-driven applications}: using spreadsheets as a backing
+data store and computation layer for interactive web applications,
+enabling end users to create such applications more easily. Research
+projects like Object Spreadsheets \autocite{mccutchen2016}, Quilt
+\autocite{benson2014}, Gneiss \autocite{chang2014}, and Marmite
+\autocite{wong2007}, as well as commercial tools like Airtable
+\autocite{zotero-79} and Glide \autocite{zotero-81} allow users to view
+data in a spreadsheet table, compute over the data using formulas, and
+then connect the table to a GUI. Because many users are already familiar
+with using spreadsheets, this way of creating applications tends to be
+much easier than traditional software methods; for example, in a user
+study of Quilt, many users were able to create applications in under 10
+minutes, even if they expected it would take them many hours.
 
@@
 done using a separate data flow pane rather than spreadsheet formulas.
-(Marmite's approach led to some confusion in a user study, because users
-expected behavior more similar to spreadsheets \autocite{wong2007}.)
+Marmite's approach led to some confusion in a user study, because users
+expected behavior more similar to spreadsheets \autocite{wong2007}.
 Airtable deviates the furthest: although the user interface resembles a
@@
 customization}, a technique that enables end users to customize software
-by augmenting an application's UI with a spreadsheet. We hope that this
-technique contributes to making the Web into a more dynamic medium that
-users can mold to their own needs.
-
-There are still many open questions about spreadsheet-driven
-customization which we hope to answer through targeted development and
+by augmenting an application's UI with a spreadsheet. We still have many
+open questions which we hope to answer through targeted development and
 usage of Wildcard.{\footnote{Up to date information about Wildcard is
@@
 plan to develop more site adapters and demos to explore this question
-further. We will start with privately testing the system with our own
-needs, and then eventually deploy the tool publicly, once the API is
-stable enough and can support a critical mass of sites and use cases. We
-also plan to run usability studies to evaluate and improve the design of
-the tool.
+further. We will start by privately testing the system with our own
+needs and then eventually deploy the tool publicly once it has a stable
+API and can support a critical mass of sites and use cases. We also plan
+to run usability studies to evaluate and improve the design of the tool.
 
@@
 
-Another open question is how easily site adapters can be created and
+Another question is how easily site adapters can be created and
 maintained for real websites, which often include complex markup and
@@
 improvements we've considered include developing automated heuristics to
-assist with the adapter creation process, and developing new
-abstractions that make it easier for programmers to efficiently create
-new robust adapters.
+assist with the adapter creation process and developing new abstractions
+that make it easier for programmers to efficiently create robust
+adapters.
+
+We hope that spreadsheet-driven customization can help the Web reach its
+full potential as a dynamic medium that users can mold to their own
+needs.
 

### 2020-01-14 paper.tex
submission ready

@@
 in enabling customization. Although spreadsheet-driven customization
-does not depend on cooperation from first-party website developers,
+does not depend on cooperation from first party website developers,
 their participation in exposing structured data would eliminate the need
 for third party site adapters. The first parties could themselves
-benefit from such a system, since providing Wildcard support---a fairly
+benefit from such a system; providing Wildcard support---a fairly
 low-effort undertaking for a developer with direct access to the
@@
   \emph{Reliability}: It can be difficult to know whether a command will
-  consistently work over time. In addition to changes to the website,
-  changes to the system's heuristics can also cause unexpected changes
-  in behavior.
+  consistently work over time. Changes to the website or to the system's
+  heuristics can cause unexpected changes in behavior.
 \item
-  \emph{Discoverability}: it can be hard for users to find discover
-  possible commands.
+  \emph{Discoverability}: it can be difficult for users to find explore
+  the space of possible commands.
 \end{itemize}
@@
 workflows to be useful in practice so it might be worth considering how
-to support them, but it's not clear how such workflows could fit into
-Wildcard. Spreadsheets have a fundamentally different computation model,
-and site adapters created by programmers cannot easily account for every
-possible action a user might want to perform in a page.
+to incorporate them, but it's not obvious how such workflows could fit
+into Wildcard. Spreadsheets have a fundamentally different computation
+model, and site adapters created by programmers cannot easily account
+for every possible action a user might want to perform in a page.
 
@@
 then connect the table to a GUI. Because many users are already familiar
-with using spreadsheets, this way of creating applications tends to be
-much easier than traditional software methods; for example, in a user
-study of Quilt, many users were able to create applications in under 10
+with spreadsheets, this way of creating applications tends to be much
+easier than traditional software methods; for example, in a user study
+of Quilt, many users were able to create applications in under 10
 minutes, even if they expected it would take them many hours.
@@
 
-We hope that spreadsheet-driven customization can help the Web reach its
-full potential as a dynamic medium that users can mold to their own
-needs.
+Ultimately, we hope that spreadsheet-driven customization helps the Web
+reach its full potential as a dynamic medium that users can mold to
+their own needs.
 
 \acks
-Thanks to Glen Chiacchieri, David Karger, Steve Krouse, Rob Miller, Arvind Satyanarayan, Daniel Windham, and Maggie Yellen for providing valuable feedback on this work.
+Thanks to Tarfah Alrashed, Glen Chiacchieri, David Karger, Steve Krouse, Rob Miller, Santiago Perez De Rosso, Arvind Satyanarayan, Daniel Windham, and Maggie Yellen for providing valuable feedback on this work.
 
